---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

This page is a short, practical introduction to `aphantasiaWMStrats`: how to
load it, what the data looks like, and how the scores are produced. It is not
where the findings are; for those, start with the
[task design](task-design.html) page and continue through the Extended Online
Report. This page is about *using the package*.

```{r setup}
library(aphantasiaWMStrats)
```

## The data {#start-data}

One built-in dataset, `all_data`, holds every version of the task in one
item-level table: one row per presented item, so three rows per trial. An
item is one word shown at one orientation in one colour; a trial is one
encode-distract-recall cycle, and there are 21 per participant.

```{r glimpse}
dplyr::glimpse(all_data[, 1:12])
```

Columns worth knowing about before anything else.

**`version`** is v1, v2 or v3. These are not interchangeable: the parity
distractor carries a scoring penalty in v2 and v3 but not v1, and recall order
is fixed in v1/v2 and randomised in v3. **v1 (N = 88) is the primary analysis
sample**; v2 (N = 9) and v3 (N = 21) are described but do not carry the
inferential analyses. The [task design](task-design.html) page explains why.

**`expe_phase`** separates `tutorial`, `training` and the three
`expe_block_*` phases. Tutorial rows carry placeholder values in the task's own
feedback columns and should always be filtered out.

**`score_word`, `score_angle`, `score_color`** are the recall scores, on [0, 1]
with 1 as a perfect match. Their `_z` counterparts are the same values z-scored
within version.

**`responded_word`, `responded_angle`, `responded_color`** say whether the
participant answered at all. These matter more than they look: the task encodes
non-response with sentinel values rather than missing values, and non-responses
are scored 0. An analysis of *accuracy* should filter on these flags; one that
does not is measuring a blend of accuracy and willingness to answer.

**`live_diff_*` and `feedback_*`** are the scores the task computed in the
browser to show participants feedback. They are kept for provenance and should
never be analysed. One of them is derived from a defective edit-distance
implementation, which the [scoring](scoring.html) page demonstrates.

## A first look {#start-look}

Experimental blocks only, answered items only:

```{r first-look}
all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase), version == "v1") |>
  dplyr::summarise(
    items = dplyr::n(),
    answered = sum(responded_word),
    mean_score_when_answered = mean(score_word[responded_word]),
    .by = expe_phase
  ) |>
  knitr::kable(digits = 3, caption = "Word recall by block, v1")
```

## The scoring functions {#start-scoring}

The score columns are not read from the raw data; they are computed from raw
target and response values by `compute_scores()`, which is exported so the
pipeline can be re-run or inspected.

```{r scoring-functions}
# Word recall: edit-distance scoring, 1 = exact match.
score_word(c("fenetre", "fenetre", "fenetre"), c("fenetre", "fenrete", "table"))

# Angular features: cosine of the error, on the right period for each.
# Colour is a 360-degree hue wheel; orientation is 180-degree symmetric.
score_angular(target = c(0, 0, 0), response = c(0, 45, 90), period = 180)
```

The period argument is not decoration. A rectangle tilted 90 degrees from
target is as wrong as an orientation can be, and only the 180-degree period
scores that 0.

`compute_scores()` takes a stimulus-level frame and appends the score columns,
their standardised counterparts, and the response flags:

```{r compute-scores}
all_data |>
  dplyr::filter(version == "v1") |>
  head(3) |>
  compute_scores() |>
  dplyr::select(
    target_word, response_word, score_word,
    responded_word, responded_angle, responded_color
  ) |>
  knitr::kable(digits = 3)
```

Two lower-level functions are exported for the same reason,
`normalise_word()` and `damerau_levenshtein()`. The second is a local
implementation rather than a dependency, partly so it could be tested against
known values: the task's own edit distance was wrong for three versions
precisely because nothing tested it.

```{r distance}
damerau_levenshtein("chien", "chein")  # one adjacent transposition
normalise_word("  Fen\u00eatre ")      # lowercase, strip spaces and accents
```

## Beyond scoring {#start-beyond}

The package also exports the machinery the analyses run on, so that a
script and a documentation page compute the same quantity with the same
code rather than two copies that drift apart.

**Building the composition.** Relative allocation across the three
features is a three-part composition, which cannot be analysed as three
proportions. `compose_features()` builds the parts from responded items
only, and `ilr_coords()` moves them off the simplex.

```{r composition}
v1 <- dplyr::filter(all_data, version == "v1", grepl("^expe_block", expe_phase))

parts <- compose_features(v1, id)
head(dplyr::select(parts, id, tidyselect::starts_with("part_")), 3) |>
  knitr::kable(digits = 3)

head(ilr_coords(parts[stats::complete.cases(parts), ]), 3) |>
  knitr::kable(digits = 3)
```

The partition is an explicit argument rather than a hidden constant,
because which contrast the first coordinate names is a substantive choice
and not a statistical one.

**Reporting quantities.** `split_half_reliability()`, `spearman_brown()`,
`correlation_test()`, `mars_knots()`, `partition_variance()` and
`centred_correlations()` are the estimators behind the numbers on the
documentation pages.

**Model specifications.** The formulas and priors are exported objects:

```{r model-spec}
joint_formula()
```

This matters for reproducibility. Documentation pages
load fitted models with `file_refit = "never"`, and brms returns a cached
fit **without checking** that the formula it was handed matches the one the
model was fitted with. A page that re-typed its formula could display
something that is not what produced its numbers. Calling the builder makes
that impossible.

`joint_priors()`, `performance_formula()`, `composition_formula()` and
`fit_brms_model()` complete the set.

**Figures.** `theme_pdf()` and the `scale_*_aphantasia()` scales are shared
with the sibling package `aphantasiaEmotions`, so imagery groups are the
same colours across both. Analysis scripts render vector PDFs at printed
size; these pages render the same figures at `base_size = 16`.

## Where to go next {#start-next}

The documentation is an Extended Online
Report: interlinked pages ordered so
that no quantity appears before the page that defines it. To dive in:

- [**Task design**](task-design.html) for what participants did and why the
  task changed twice.
- [**Scoring**](scoring.html) for how these numbers are made.
- [**Task engagement**](engagement.html) for the two ways participants
  declined to do the task, and why declining is itself informative.
- [**Versions**](version-scope.html) for which of the three task versions
  the analyses use.
- [**Task validity**](task-validity.html) for whether these scores are
  reliable enough to carry an individual-differences claim. One of the
  three features is not.
- [**Questionnaire psychometrics**](psychometrics.html) for the same
  question asked of the VVIQ, OSIVQ and NIEQ.
- [**Codebook**](codebook.html) for every column in `all_data`.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
