---
title: "The analysis: what is modelled, and why"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Analysis strategy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

The preceding pages establish [who took
part](participants.html), what was measured, on which sample, and
whether it measures anything. This page says what is done with it, and in
what order. It contains no results: it is the reasoning a reader needs
before the modelling pages make sense, and it is deliberately written so
that it could have been written before any model was fitted.

Everything it describes runs on **task version v1**, for the reasons the
[version scope](version-scope.html) page sets out. The one exception is
the exploratory strand, which pools all three versions because it uses
questionnaires rather than task behaviour.

## 1. The question the models have to answer

Aphantasic participants perform about as well as typical imagers on visual
working memory tasks, and report compensating with verbal or spatial
strategies. A strategy is, by definition, what someone takes themselves to
be doing, so no performance pattern can outrank their own account of it.
What a behavioural measure can add is **convergence**: a dependent variable
that, if it tracks the same distinctions participants report, corroborates
the subjective category rather than replacing it.

The WM-FTT was built to provide one. Three features per stimulus, partial
credit for each, and instructions to maximise the score, so that a
participant who divides effort unevenly leaves a trace in the scores
themselves.

"Where does effort go" then has two possible readings, and the difference
between them shaped everything here.

- **Absolute performance.** Are aphantasic participants worse at some
  features than typical imagers?
- **Relative allocation.** Independently of overall ability, do they divide
  effort differently across the three features?

These were treated as competing framings for most of this project's
history. They are complementary, and both turned out to be **secondary** to
something neither of them captures.

## 2. What changed: abstention is a behaviour

Recall of each feature is optional, and participants used that. The [task
engagement](engagement.html) page shows that declining to report a feature
is not missing data: under a points-per-feature incentive, **choosing not
to report is itself an allocation decision**, and arguably the most direct
one the task records. A participant who abandons orientation entirely to
protect colour has made a clearer strategic statement than one whose
orientation score is slightly lower.

Both framings in section 1 filter that behaviour out before analysing what
remains. That is a decision about which quantity is interesting, not a
neutral preprocessing step.

There is also a statistical problem with filtering it out. Analysing only
responded trials conditions on a selected subsample. If the participants
who abstain more also perform differently when they do answer, the
conditional estimates are biased in a way **no responders-only analysis can
detect**, because the information needed to detect it has been removed.

That was tested rather than assumed. Fitting response and accuracy jointly
on orientation gives a participant-level correlation between the two of
0.51, with 95% of the posterior between 0.26 and 0.72. The rule for what
that would mean was written down before the number existed: conditioning
was not innocent, and the joint model becomes the primary analysis.

## 3. How imagery enters every model

Before any model, one decision applies to all of them: **what shape does
imagery vividness take as a predictor?**

Not a smooth one. VVIQ scores in this sample are a sharp, isolated spike at
the scale minimum plus an irregular remainder above it, so a single slope
fitted through everything would describe neither part. Participants at the
floor also have **no vividness variance among themselves**: a slope cannot
be estimated for them, while an offset can.

Every model on this site therefore takes the same form:

- a **slope** across participants above the floor, and
- a single **offset** for the group at it.

Rather than assume that form, several were compared, with this one
pre-declared as primary regardless of which fit best. The
[performance](performance.html) page reports the comparison; the data do
not separate the candidates, which was predicted in advance.

**This is not a side decision, it is the spine.** The floor-group offset is
what every result on this site reports: the compositional shift, the
per-feature accuracies, and the joint model's six responses all estimate it
in the same way, which is what makes them comparable.

### The caution that comes with it

The offset is the distance between where the floor group sits and **where a
relationship fitted on everyone else predicts they would sit**. Above the
floor, participants cluster at the top of the vividness range with only a
handful below 40, so that relationship is anchored by a dense group at 50
to 80 plus a scatter, and extending it down to 16 rests on a linearity
assumption this sample barely constrains. The sibling study
[`aphantasiaEmotions`](https://m-delem.github.io/aphantasiaEmotions/) makes
the same move with 1478 participants across five studies; there are 86
here. On every figure, that stretch of the line is drawn **dashed** to mark
it as extrapolation rather than fit.

Two further things belong on the record. The model was first fitted *after*
seeing an association that motivated it, which is a legitimate thing to do
and an illegitimate thing to describe as pre-specified; it was pre-declared
for the models on these pages, not for the one that suggested it. And its
intervals comfortably contain both nothing and a moderate effect, so an
offset that does not clear its region of practical equivalence should be
read as low power rather than as evidence of absence.

## 4. The primary analysis: allocation

Six responses. Three gates, one per feature, for whether it was reported at
all. Three accuracies, one per feature, for how well it was reported when
it was. All six share a participant-level correlation matrix.

```{r formula}
joint_formula()
```

The `subset()` terms are what let a single trial carry six responses with
independent non-response, and what let a participant contribute to a gate
whether or not they ever answered that feature. One participant answered
**zero** orientation trials: invisible to every responders-only analysis,
and highly informative here.

The fifteen correlations are the object of interest, in four families:

| Family | Question |
|---|---|
| Response and accuracy, within a feature | Was conditioning on responding legitimate, and is the effect feature-specific or a general engagement trait? |
| Accuracy and accuracy, across features | Does doing well on one feature cost another, between people? |
| Response and response, across features | Do people who skip one feature skip others? |
| The six cross terms | Everything else |

Which of them are foregrounded was fixed in advance, in three tiers, so
that choosing among fifteen after seeing them is not possible.

## 5. The other models, and what they are for

Two models sit around it, both using the form set out in section 3.

- **[Performance](performance.html).** Absolute per-feature accuracy, with
  a response family chosen per feature. The same data read as *how much*
  rather than *where*, which is section 1's first reading.
- **[The joint model](joint-model.html).** Six responses: whether each
  feature was reported at all, and how well it was reported when it was,
  with the correlations between those six quantities as the object of
  interest.

The joint model is not a bigger version of the composition. It exists
because the composition is computed from **responded trials only**, and
whether that conditioning was innocent cannot be checked from inside a
responders-only analysis. The joint model checks it.

So the precedence between them is specific rather than general. **On
allocation, the composition is primary.** On selection and on willingness
to report, the joint model is authoritative and the composition is silent.
Where the joint model bears on allocation, it speaks to whether the
composition was *computable*, not to what it found.

## 6. Confirmatory and exploratory

Everything above uses groups defined by imagery vividness, **fixed before
any model was fitted**. That is the confirmatory strand.

A second strand asks whether the sample contains structure that vividness
does not capture: how the questionnaire scales relate to each other, and
whether participants cluster into groups the vividness split does not
recover. That is exploratory, is labelled as such on every page, and
generates hypotheses rather than testing them.

The interesting outcome there would be clusters that **split** the group
of complete aphantasics, since that is a distinction the confirmatory
strand cannot make at all. Clusters that merely recover the vividness
grouping would be a reportable null.

## 7. What was fixed in advance, and where to check

Pre-declaration only means anything if it is checkable, so the planning
documents that record each decision ship with the package, in
`inst/planning/`. They are working documents rather than prose, and they
include the reversals: decisions made, tested against data, and withdrawn.

Three examples, each of which changed an analysis rather than a sentence.
The compositional input was specified as standardised scores, then found to
be roughly half negative and therefore impossible for any log-ratio
transform. An association between reporting propensity and imagery was
reported, then dissolved when stratified by task version. A two-way
trade-off between colour and orientation was reported, then found to
describe six low-engagement participants rather than the task.

---

**Continuing through the Extended Online Report:** this page follows [questionnaire psychometrics](psychometrics.html). To keep reading in order, continue to [composition](composition.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
