---
title: "The analysis: what is modelled, and why"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Analysis strategy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

The preceding pages establish [who took
part](participants.html), what was measured, on which sample, and
whether it measures anything. This page says what is done with it, and in
what order. It reports no findings — every result on this site belongs to
a later page — and it is deliberately written so that it could have been
written before any model was fitted. Two numbers do appear, both because a
decision on this page turns on them: the selection correlation in section
2, which is why the joint model exists at all, and the outcome of the
functional-form comparison in section 3, which is why the floor-group form
survived being pre-declared.

Everything it describes runs on **task version v1**, for the reasons the
[version scope](version-scope.html) page sets out. The one exception is
the exploratory strand, which pools all three versions because it uses
questionnaires rather than task behaviour.

## 1. The question the models have to answer

Aphantasic participants perform about as well as typical imagers on visual
working memory tasks, and report compensating with verbal or spatial
strategies. A strategy is, by definition, what someone takes themselves to
be doing, so no performance pattern can outrank their own account of it.
What a behavioural measure can add is **convergence**: a dependent variable
that, if it tracks the same distinctions participants report, corroborates
the subjective category rather than replacing it.

The WM-FTT was built to provide one. Three features per stimulus, partial
credit for each, and instructions to maximise the score, so that a
participant who divides effort unevenly leaves a trace in the scores
themselves.

"Where does effort go" then has two possible readings, and the difference
between them shaped everything here.

- **Absolute performance.** Are aphantasic participants worse at some
  features than typical imagers?
- **Relative allocation.** Independently of overall ability, do they divide
  effort differently across the three features?

These were treated as competing framings for most of this project's
history. They are complementary, and both turned out to be **secondary** to
something neither of them captures.

## 2. What changed: abstention is a behaviour

Recall of each feature is optional, and participants used that. The [task
engagement](engagement.html) page shows that declining to report a feature
is not missing data: under a points-per-feature incentive, **choosing not
to report is itself an allocation decision**, and arguably the most direct
one the task records. A participant who abandons orientation entirely to
protect colour has made a clearer strategic statement than one whose
orientation score is slightly lower.

Both framings in section 1 filter that behaviour out before analysing what
remains. That is a decision about which quantity is interesting, not a
neutral preprocessing step.

There is also a statistical problem with filtering it out. Analysing only
responded trials conditions on a selected subsample. If the participants
who abstain more also perform differently when they do answer, the
conditional estimates are biased in a way **no responders-only analysis can
detect**, because the information needed to detect it has been removed.

That was tested rather than assumed. Fitting response and accuracy jointly
on orientation — in a bivariate model of that feature alone, not the
six-response model of section 5 — gives a participant-level correlation
between the two of 0.51, with 95% of the posterior between 0.26 and 0.72.
That figure is a posterior **mean**; the [joint
model](joint-model.html) page reports the same parameter as a median,
0.52, which is why the two differ in the second digit.

The rule for what that would mean was written down before the number
existed: conditioning was not innocent, so a model that does not condition
on it is needed. What it is needed *for* is set out in section 5 — it
settles whether the composition was computable, not what the composition
found.

## 3. How imagery enters every model

Before any model, one decision applies to all of them: **what shape does
imagery vividness take as a predictor?**

Not a smooth one. VVIQ scores in this sample are a sharp, isolated spike at
the scale minimum plus an irregular remainder above it, so a single slope
fitted through everything would describe neither part. Participants at the
floor also have **no vividness variance among themselves**: a slope cannot
be estimated for them, while an offset can.

Every model on this site therefore takes the same form:

- a **slope** across participants above the floor, and
- a single **offset** for the group at it.

Rather than assume that form, several were compared, with this one
pre-declared as primary regardless of which fit best. Two comparisons were
run: four candidates on orientation accuracy
([performance](performance.html) §4) and five on the first compositional
coordinate ([composition](composition.html) §8). Neither separates the
candidates, which was predicted in advance.

**This is not a side decision, it is the spine.** The floor-group offset is
what every result on this site reports: the compositional shift, the
per-feature accuracies, and the joint model's six responses all estimate it
in the same way, which is what makes them comparable.

### The caution that comes with it

The offset is the distance between where the floor group sits and **where a
relationship fitted on everyone else predicts they would sit**. Above the
floor, participants cluster at the top of the vividness range with only a
handful below 40, so that relationship is anchored by a dense group at 50
to 80 plus a scatter, and extending it down to 16 rests on a linearity
assumption this sample barely constrains. The sibling study
[`aphantasiaEmotions`](https://m-delem.github.io/aphantasiaEmotions/) makes
the same move with 1478 participants across five studies; there are at most
86 here — every v1 participant with a vividness score, and fewer on the
pages that also require an engaged composition (section 6). On every figure, that stretch of the line is drawn **dashed** to mark
it as extrapolation rather than fit.

Two further things belong on the record. The model was first fitted *after*
seeing an association that motivated it, which is a legitimate thing to do
and an illegitimate thing to describe as pre-specified; it was pre-declared
for the models on these pages, not for the one that suggested it. And its
intervals comfortably contain both nothing and a moderate effect, so an
offset that does not clear its region of practical equivalence should be
read as low power rather than as evidence of absence.

## 4. The primary analysis: allocation

Divide each participant's three feature scores by their sum and what
remains is a **composition**: three parts summing to one, carrying only
relative information. Overall ability is gone; only where the effort went
survives.

Three parts summing to a constant cannot vary independently, so the
composition is moved off the simplex into two unconstrained isometric
log-ratio coordinates before anything is modelled: `ilr1` contrasts words
against the geometric mean of the two non-verbal features, `ilr2`
contrasts colour against orientation. Which contrast comes first is a
substantive choice rather than a statistical one, and the
[composition](composition.html) page shows that the omnibus test does not
depend on it.

```{r formula}
composition_formula("vviq + complete_aphant + parity_rate")
```

Both coordinates are modelled together with the residual correlation
between them estimated. That is not cosmetic: the invariance above holds
for a genuine multivariate model, and two separate univariate fits would
lose it.

This is the analysis the study was **designed** around. The thesis's
prediction is about allocation; it is the coordinate on which vividness,
inner experience and participants' own accounts of what they were doing
all bear; and it is the result that clears its pre-declared threshold.

## 5. The other models, and what they are for

Two models sit around it, both using the form set out in section 3.

- **[Performance](performance.html).** Absolute per-feature accuracy, with
  a response family chosen per feature. The same data read as *how much*
  rather than *where*, which is section 1's first reading.
- **[The joint model](joint-model.html).** Six responses: whether each
  feature was reported at all, and how well it was reported when it was,
  with the correlations between those six quantities as the object of
  interest.

The joint model is not a bigger version of the composition. It exists
because the composition is computed from **responded trials only**, and
whether that conditioning was innocent cannot be checked from inside a
responders-only analysis. The joint model checks it.

So the precedence between them is specific rather than general. **On
allocation, the composition is primary.** On selection and on willingness
to report, the joint model is authoritative and the composition is silent.
Where the joint model bears on allocation, it speaks to whether the
composition was *computable*, not to what it found.

## 6. The samples, once {#samples}

Six numbers circulate on the modelling pages and they are easy to
conflate, so they are computed here in one place and referred to rather
than restated. Every one of them is a filter on v1, applied for a reason
given on an earlier page.

```{r samples}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

vividness <- dplyr::summarise(
  v1, vviq = dplyr::first(vviq_total_score), .by = id
)

parts <- compose_features(v1, id) |>
  dplyr::filter(stats::complete.cases(dplyr::pick(tidyselect::everything())))

engaged <- dplyr::filter(parts, id %in% engaged_ids(v1))

with_vviq <- engaged |>
  dplyr::left_join(vividness, by = "id") |>
  dplyr::filter(!is.na(vviq))

with_scales <- dplyr::filter(
  with_vviq, id %in% questionnaire_scales(all_data)$id
)

tibble::tibble(
  sample = c(
    "v1 participants",
    "With a vividness score",
    "All three compositional parts present",
    "Also clearing the engagement thresholds",
    "Also with a vividness score",
    "Also with every questionnaire scale"
  ),
  n = c(
    dplyr::n_distinct(v1$id),
    sum(!is.na(vividness$vviq)),
    nrow(parts), nrow(engaged), nrow(with_vviq), nrow(with_scales)
  ),
  used_by = c(
    "Descriptives",
    "The joint model",
    "Composition §2",
    "Composition §§2, 3, 6",
    "Composition §§4, 7, 8; performance",
    "Composition §5"
  )
) |>
  knitr::kable(caption = "Every modelling sample on this site")
```

Two things follow from the table that are worth stating in words, because
both have been misreported at some point in this project's history.

The **joint model uses more participants than anything else**, not fewer.
The engagement thresholds exist to stop imprecise participant *means*
being treated as measurements; the joint model forms no participant means,
so it does not need them, and the participants they exclude are precisely
the ones with the most to say about abstention.

The **modelling sample is smaller than the engaged sample**, by the two
participants with no VVIQ score. Every model on this site reports a
floor-group offset, and a participant with no vividness score cannot be
placed on either side of the floor.

## 7. Confirmatory and exploratory

Everything above uses groups defined by imagery vividness, **fixed before
any model was fitted**. That is the confirmatory strand.

A second strand asks whether the sample contains structure that vividness
does not capture: how the questionnaire scales relate to each other, and
whether participants cluster into groups the vividness split does not
recover. That is exploratory, is labelled as such on every page, and
generates hypotheses rather than testing them.

The interesting outcome there would be clusters that **split** the group
of complete aphantasics, since that is a distinction the confirmatory
strand cannot make at all. Clusters that merely recover the vividness
grouping would be a reportable null.

## 8. What was fixed in advance, and where to check

Pre-declaration only means anything if it is checkable, so the planning
documents that record each decision ship with the package, in
`inst/planning/`. They are working documents rather than prose, and they
include the reversals: decisions made, tested against data, and withdrawn.

Three examples, each of which changed an analysis rather than a sentence.
The compositional input was specified as standardised scores, then found to
be roughly half negative and therefore impossible for any log-ratio
transform. An association between reporting propensity and imagery was
reported, then dissolved when stratified by task version. A two-way
trade-off between colour and orientation was reported, then found to
describe six low-engagement participants rather than the task.

---

**Continuing through the Extended Online Report:** this page follows [questionnaire psychometrics](psychometrics.html). To keep reading in order, continue to [composition](composition.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
