---
title: "Beyond vividness: what the questionnaires say"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Beyond vividness}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
set.seed(20260825)

pkg <- "aphantasiaWMStrats"
refit <- "never"
```

**Everything on this page is exploratory.** The groups are not fixed in
advance, nothing here licenses a confirmatory claim, and the parts that
were run after seeing a result say so.

The confirmatory strand divides participants by imagery vividness. This
page asks whether the questionnaires contain anything that division misses:
first by looking at how the scales relate to each other, then by asking an
algorithm to find groups without being told about vividness at all.

The short answer is **yes on one dimension and no as a group structure**,
and the two halves are worth reading together.

```{r data}
scales <- questionnaire_scales(all_data)

scales <- dplyr::mutate(
  scales,
  imagery_group = factor(
    dplyr::if_else(vviq == 16, "VVIQ floor", "Above floor"),
    levels = c("Above floor", "VVIQ floor")
  )
)
```

## 1. Pooling across versions, which is allowed here and nowhere else

Every modelling page on this site runs on v1 alone. This one uses all
`r nrow(scales)` participants with complete questionnaire data.

The reasoning in the [version scope](version-scope.html) page is entirely
about **task** comparability: group balance for a behavioural contrast,
non-response differing three-fold, per-version standardisation being
unstable. None of it touches a questionnaire score. The instruments are
identical and were administered identically in all three versions, so
version is a property of the task rather than of the scales, and pooling
recovers a third more participants for the analysis that most needs them.

```{r versions}
scales |>
  dplyr::summarise(
    n = dplyr::n(),
    at_floor = sum(imagery_group == "VVIQ floor"),
    percent_floor = round(100 * mean(imagery_group == "VVIQ floor")),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(caption = "Participants by task version")
```

What pooling changes is the sample's **composition**: later recruitment
targeted aphantasic participants, so the pooled sample is about a third
floor group where v1 alone is a quarter. That matters for clustering,
because centroids are fitted to whatever sample they are given, and
section 4 tests it rather than assuming it away.

## 2. The scales are largely independent, and three of them are one thing

Raw correlations among overlapping instruments are close to unreadable:
everything correlates with everything, because everything shares a general
factor. The partial structure is what says whether a scale carries anything
the others do not.

```{r partial, fig.alt="Partial correlations among the questionnaire scales.", fig.height = 6}
standardised <- standardise_scales(scales) |>
  dplyr::select(-"imagery_group")

partial <- correlation::correlation(
  dplyr::select(standardised, -"id", -"version"),
  partial = TRUE, p_adjust = "bonferroni"
)

as.data.frame(partial) |>
  dplyr::mutate(label = ifelse(p < 0.05, sprintf("%.2f", r), "")) |>
  ggplot(aes(x = Parameter1, y = Parameter2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.4) +
  geom_text(aes(label = label), size = 3) +
  scale_fill_gradient2(limits = c(-1, 1), name = "Partial r",
                       guide = guide_colourbar(barwidth = 10, barheight = 0.6)) +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       caption = "Bonferroni corrected; values shown where p < .05.") +
  theme_pdf(base_size = 16, axis.text.x = element_text(angle = 45, hjust = 1),
            legend.position = "top")
```

After partialling, the only strongly connected block is the three imagery
measures: **VVIQ, OSIVQ's object subscale, and NIEQ's mental-imagery
dimension**. Everything else is near zero.

```{r imagery-block}
imagery_components <- c("vviq", "osivq_object", "nieq_imagery")
components <- dplyr::select(standardised, tidyselect::all_of(imagery_components))

knitr::kable(round(stats::cor(components, method = "spearman"), 3),
             caption = "The three imagery measures, raw correlations")
```

Cronbach's alpha of the three is
`r fmt(cronbach_alpha(components), 3)`, against correlations of 0.15 to
0.38 with the other OSIVQ subscales. **Three instruments, three response
formats, one construct.**

That has a consequence for anything that comes next. Left as three
features, imagery would be **triple-weighted** in any distance metric, so a
clustering would recover imagery groups by construction and could not
possibly find anything else. They are collapsed into one standardised
composite, which is what makes the question in this page's title
answerable.

The rest being largely independent also means dimension reduction should
stop there. Reducing further would destroy exactly the distinctions the
clustering is meant to find.

## 3. The finding: aphantasic participants report more unsymbolised thinking

Before any clustering, the simplest question. The NIEQ asks about five
modes of inner experience. How do the two imagery groups differ on each?

```{r nieq-contrast}
nieq_contrast <- purrr::map(
  grep("^nieq_", names(scales), value = TRUE),
  function(dimension) {
    floor_values <- scales[[dimension]][scales$imagery_group == "VVIQ floor"]
    above_values <- scales[[dimension]][scales$imagery_group == "Above floor"]
    pooled_sd <- sqrt(
      ((length(floor_values) - 1) * stats::var(floor_values) +
         (length(above_values) - 1) * stats::var(above_values)) /
        (length(floor_values) + length(above_values) - 2)
    )
    # computed before the tibble: inside it, later expressions see earlier
    # columns, so naming a column `floor` would make this test two means
    test <- suppressWarnings(stats::wilcox.test(floor_values, above_values))

    tibble::tibble(
      dimension = sub("^nieq_", "", dimension),
      floor = mean(floor_values), above_floor = mean(above_values),
      d = (mean(floor_values) - mean(above_values)) / pooled_sd,
      p = test$p.value
    )
  }
) |>
  purrr::list_rbind() |>
  dplyr::arrange(dplyr::desc(d))

nieq_contrast |>
  # p rounded to 3 decimals reads as 0.000 for the two that matter most
  dplyr::mutate(p = format.pval(p, digits = 2, eps = 1e-4)) |>
  knitr::kable(digits = 2, caption = "NIEQ dimensions by imagery group")
```

**Unsymbolised thinking is the only dimension where the floor group scores
higher**, and the effect is large. Participants who report no voluntary
visual imagery recognise themselves more in what the NIEQ, following
Hurlburt's descriptive experience sampling work, calls thinking that
proceeds without words or images.

```{r nieq-models, fig.alt="Floor-group models for four NIEQ dimensions.", fig.height = 8}
nieq_data <- dplyr::mutate(
  scales,
  complete_aphant = factor(
    dplyr::if_else(vviq == 16, "floor", "above_floor"),
    levels = c("above_floor", "floor")
  )
)

# Beta after a Smithson-Verkuilen squeeze, as on the accuracy pages: every
# dimension has participants at a boundary, so a Gaussian would put mass
# outside the scale. The imagery dimension is excluded because it
# correlates 0.87 with the score that defines the grouping.
nieq_dimensions <- c("nieq_unsymbolised", "nieq_inner_voice",
                     "nieq_emotions", "nieq_sensory_focus")

nieq_panel <- function(dimension, model_only = FALSE) {
  fit_data <- nieq_data
  fit_data$score <- squeeze_boundaries(fit_data[[dimension]] / 100)

  model <- fit_brms_model(
    formula = brms::bf(score ~ vviq + complete_aphant, family = brms::Beta()),
    data    = fit_data,
    prior   = response_priors(NULL, terms = c("vviq", "complete_aphant")),
    file    = system.file(
      "models", paste0("nieq-floor-", sub("^nieq_", "", dimension), ".rds"),
      package = pkg),
    file_refit = refit
  )
  if (model_only) return(model)

  grid <- tibble::tibble(
    vviq = seq(16, 80, length.out = 200),
    complete_aphant = factor("above_floor", levels = c("above_floor", "floor"))
  )
  above <- brms::posterior_epred(model, newdata = grid)

  floor_row <- grid[1, ]
  floor_row$complete_aphant <- factor("floor",
                                      levels = c("above_floor", "floor"))
  floor_draws <- as.vector(brms::posterior_epred(model, newdata = floor_row))

  observed <- tibble::tibble(x = nieq_data$vviq,
                             y = nieq_data[[dimension]] / 100)
  effect <- brms::fixef(model)["complete_aphantfloor", ]

  plot_floor_group(
    observed = dplyr::filter(observed, x > 16),
    fitted = tibble::tibble(
      x = grid$vviq,
      estimate = apply(above, 2, stats::median),
      lower = apply(above, 2, stats::quantile, 0.025),
      upper = apply(above, 2, stats::quantile, 0.975)
    ),
    floor_draws = floor_draws,
    floor_observed = dplyr::filter(observed, x == 16)$y,
    effect_label = sprintf("%.2f\n[%.2f, %.2f]", effect[["Estimate"]],
                           effect[["Q2.5"]], effect[["Q97.5"]]),
    # only the bottom row carries the axis label, and the arrow needs more
    # room at half width than it does on a full-width panel
    x_lab = if (dimension %in% nieq_dimensions[3:4]) "VVIQ total score",
    y_lab = gsub("_", " ", sub("^nieq_", "", dimension)),
    base_size = 16, point_size = 1.8,
    left_expansion = 0.24, arrow_nudge = -7
  ) +
    ggplot2::scale_y_continuous(
      labels = \(x) round(100 * x), limits = c(0, 1))
}

patchwork::wrap_plots(purrr::map(nieq_dimensions, nieq_panel), ncol = 2)
```

Fitted in the same floor-group form as every other model on this site: a
slope among everyone above the vividness floor, plus a single offset for
the group at it. That separates two things a group comparison cannot, and
here it reverses the reading of both.

```{r nieq-estimates}
purrr::map(nieq_dimensions, function(dimension) {
  estimates <- brms::fixef(nieq_panel(dimension, model_only = TRUE))
  tibble::tibble(
    dimension = gsub("_", " ", sub("^nieq_", "", dimension)),
    floor_offset = estimates["complete_aphantfloor", "Estimate"],
    offset_lower = estimates["complete_aphantfloor", "Q2.5"],
    offset_upper = estimates["complete_aphantfloor", "Q97.5"],
    vviq_slope = estimates["vviq", "Estimate"],
    slope_lower = estimates["vviq", "Q2.5"],
    slope_upper = estimates["vviq", "Q97.5"]
  )
}) |>
  purrr::list_rbind() |>
  dplyr::arrange(dplyr::desc(floor_offset)) |>
  knitr::kable(digits = 3, caption = "Offset and slope, separated")
```

**Not one floor offset excludes zero.** Whatever distinguishes the groups
on these dimensions is not a discontinuity at the floor.

**Unsymbolised thinking is a gradient, not a floor effect.** Its slope is
the informative estimate, running downward across the whole vividness
range, while its offset is uninformative. The group contrast in the table
above was real, but it describes the low end of a continuum rather than
something specific to participants with no imagery at all. That is the
same fact as its -0.42 partial correlation with the imagery composite: it
behaves like an inverse imagery measure, consistently, all the way up.

**And a dimension the group contrast missed.** NIEQ emotions has the only
slope clearly excluding zero, and it is positive: more vivid imagers
report more emotional inner experience. Dichotomised, that dimension gave
d = -0.42 at p = .12 and would have been set aside.

Both readings changed when the dichotomy was dropped, which is the
argument for fitting these at all.

This is a positive claim about what aphantasic experience *is*, from an
instrument built to ask, and it needs no clustering at all. It is worth
saying plainly because the rest of this page is a null. The models below
sharpen it into something more specific than a group difference.

Two cautions. It is a group contrast that was **not** pre-declared, so it
is a hypothesis for a future study rather than a test passed. And the
[psychometrics](psychometrics.html) page records that NIEQ's sensory-focus
dimension does not reach its reliability threshold in this sample, which is
a reason for care with the instrument as a whole even though it is not the
dimension at issue here.

## 4. The clustering, and what it was asked to find

The criterion was fixed before the clustering ran, so that the outcome
could not be described as whatever it turned out to be:

- **Success**: clusters that **split** the floor group. That would be a
  distinction the confirmatory strand cannot make at all.
- **Null**: clusters that recover the vividness grouping. The structure is
  imagery, and there is nothing else.

Consensus clustering rather than one algorithm, because the central risk in
this kind of analysis is finding groups that do not exist. A partition that
survives three algorithms and three consensus functions is less likely to
be an artifact of any one of them. The number of clusters is **selected,
not assumed**.

```{r clustering}
features <- standardise_scales(dplyr::select(scales, -"imagery_group")) |>
  add_imagery_composite()

feature_matrix <- as.matrix(dplyr::select(features, -"id", -"version"))
rownames(feature_matrix) <- features$id

clustering <- diceR::dice(
  feature_matrix, nk = 2:5, reps = 10, p.item = 0.8,
  algorithms = c("gmm", "pam", "hc"),
  cons.funs = c("kmodes", "majority", "CSPA"),
  seed = 20260825, progress = FALSE, verbose = FALSE
)

tibble::tibble(
  k = 2:5,
  mean_silhouette = vapply(clustering$indices$ii,
                           \(index) mean(as.data.frame(index)$silhouette),
                           numeric(1))
) |>
  knitr::kable(digits = 3, caption = "Cluster separation by k")
```

Two clusters, on the silhouette and on the proportion of ambiguous
clustering alike, and all three consensus functions returned two
independently.

```{r verdict}
scales$cluster <- factor(clustering$clusters[, "CSPA"])

table(cluster = scales$cluster, imagery = scales$imagery_group) |>
  as.data.frame.matrix() |>
  tibble::rownames_to_column("cluster") |>
  dplyr::mutate(cluster = paste("Cluster", cluster)) |>
  knitr::kable(caption = "Clusters against the vividness split")
```

The adjusted Rand index against the vividness grouping is
`r fmt(adjusted_rand_index(scales$cluster, scales$imagery_group))`, and
**every participant at the vividness floor falls in one cluster.** By the
criterion above, that is the null.

```{r profiles, fig.alt="The two clusters on the imagery composite and unsymbolised thinking."}
features |>
  dplyr::mutate(cluster = scales$cluster,
                imagery_group = scales$imagery_group) |>
  ggplot(aes(x = imagery, y = nieq_unsymbolised, colour = cluster,
             shape = imagery_group)) +
  geom_point(size = 2.4, alpha = 0.85) +
  scale_colour_manual(values = unname(palette.colors()[c(6, 4)]),
                      name = "Cluster") +
  scale_shape_manual(values = c(`Above floor` = 16, `VVIQ floor` = 17),
                     name = NULL) +
  labs(x = "Imagery composite (z)", y = "Unsymbolised thinking (z)",
       caption = "Two clouds, separated along one axis.")
```

The solution is **one axis, dichotomised**. Every feature separates in the
same direction except unsymbolised thinking, which separates in the
opposite one, and section 2 explains why: it is an inverse imagery measure
rather than a second dimension.

One respect in which this is not purely a relabelling: the boundary is not
the scale floor. The low-imagery cluster holds the floor participants
**plus** several low-vividness participants above it, so the line is drawn
at a vividness threshold rather than at the floor.

## 5. Is the pooling innocent?

Section 1 flagged that v1 participants' labels are partly determined by
participants from versions that enter no other analysis. Refit on v1 alone
and compare.

```{r stability}
v1_rows <- features$version == "v1"

restricted <- diceR::dice(
  feature_matrix[v1_rows, , drop = FALSE], nk = 2:5, reps = 10, p.item = 0.8,
  algorithms = c("gmm", "pam", "hc"),
  cons.funs = c("kmodes", "majority", "CSPA"),
  seed = 20260825, progress = FALSE, verbose = FALSE
)

stability <- cluster_stability(
  pooled = clustering$clusters[v1_rows, "CSPA"],
  restricted = restricted$clusters[, "CSPA"]
)

tibble::tibble(
  `v1 participants` = sum(v1_rows),
  `agreement after matching labels` = stability$agreement,
  `adjusted Rand` = stability$adjusted_rand
) |>
  knitr::kable(digits = 3, caption = "Pooled fit against a v1-only fit")
```

The v1 partition is essentially the same either way, so the pooling is
innocent and the larger sample is kept. Had it not been, the pooled
solution would have been describing the versions excluded from everything
else, and it would have been dropped.

## 6. Two things tried afterwards, and reported as such

Everything below was run **after** seeing the null, which makes it
exploration of a null rather than a second attempt at the same test.

**Is there structure inside the floor group?** A better question than
forcing more clusters on the whole sample, because restricting to the floor
group removes the axis that dominated, so anything found there is
necessarily beyond vividness. Features re-standardised within the
subsample, imagery scales dropped since they barely vary in it.

```{r floor-clustering}
floor_features <- scales |>
  dplyr::filter(imagery_group == "VVIQ floor") |>
  dplyr::select(-"vviq", -"osivq_object", -"nieq_imagery", -"imagery_group",
                -"cluster") |>
  standardise_scales()

floor_matrix <- as.matrix(dplyr::select(floor_features, -"id", -"version"))

floor_clustering <- diceR::dice(
  floor_matrix, nk = 2:4, reps = 10, p.item = 0.8,
  algorithms = c("gmm", "pam", "hc"),
  cons.funs = c("kmodes", "majority", "CSPA"),
  seed = 20260825, progress = FALSE, verbose = FALSE
)

knitr::kable(floor_clustering$indices$pac, digits = 3,
             caption = "Proportion of ambiguous clustering, floor group only")
```

The consensus functions returned a partition and agreed with each other.
**The stability indices say not to believe it.** The proportion of
ambiguous clustering runs from about 0.22 to 0.76 across algorithms,
meaning a large share of participant pairs cluster together inconsistently
across resamples. With `r nrow(floor_matrix)` participants and
`r ncol(floor_matrix)` features, an algorithm that always returns a
partition has returned one.

Had the indices supported it, the claim would have been a contrast between
participants high on both inner voice and unsymbolised thinking and
participants low on both. That deserves a properly powered study, and it is
recorded here so the question is visibly closed rather than left for
someone to retry with a different seed.

**Forcing three clusters** on the full sample splits off two above-floor
participants and leaves the floor group otherwise intact. It changes
nothing.

## 7. The validation step is not run, and that is the result

The plan was to refit the [joint model](joint-model.html) on the cluster
labels, using task behaviour that had been held out throughout. Its stated
precondition was clusters that are not simply imagery groups.

That precondition is not met. Refitting anyway would reproduce the
floor-group analysis with the boundary moved slightly, and reporting that
as validation of a cluster solution would be the kind of overstatement this
project has made before and does not intend to repeat.

So the exploratory strand ends here, with two results of unequal weight.
The clustering finds nothing that vividness does not already capture. The
questionnaires themselves do: **unsymbolised thinking rises steadily as
imagery vividness falls**, which points at a description of what inner
experience is like at that end of the scale rather than at a deficit. It
is a gradient rather than a property of aphantasia specifically, and the
models in section 3 are what distinguish those two readings.

One question is left deliberately open. The cluster boundary sits above the
vividness floor, so it is a different dichotomy of the same axis. Whether
it predicts task behaviour better than the floor split is answerable and
cheap, and would need declaring before it is run.

---

**Continuing through the Extended Online Report:** this page follows [composition](composition.html). To keep reading in order, continue to [model diagnostics](model-diagnostics.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
