---
title: "Codebook"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false

knitr::opts_chunk$set(
  echo = FALSE,
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(aphantasiaWMStrats)
```

This page documents every column in `all_data`, the pooled, stimulus-level
dataset from the Working Memory Feature Trade-off Task (WM-FTT):
word, orientation and colour recall under a parity-judgement distractor,
combining all three study versions (v1 original, v2 added a parity-error
penalty, v3 added recall-order randomisation).

The task was called CFA-WM during data collection, and column names, OSF
component names and the front-end code still use that prefix, for example
`strats_cfa_q01_colors_1`. Those identifiers are deliberately unchanged so
that exports stay comparable with earlier ones and with the raw JATOS
data. WM-FTT and CFA-WM refer to the same task.

The tables below are computed directly from `all_data` rather than typed out
by hand, so they cannot drift out of sync with the data itself. `all_data`
has one row per item (three per trial), so a given participant's
demographic and questionnaire-score values repeat across their rows; the
counts and ranges below reflect that (`N missing` counts item rows, not
participants).

The five nested list-columns holding raw questionnaire items are described
only briefly here (see the last section below). For full item-level detail 
(every raw VVIQ/OSIVQ/NIEQ/strategy-report column, computed the same way)
see the codebook that ships alongside `all_data_surveys` and
`all_data_full` on [OSF](https://osf.io/3649s/files/osfstorage), the two flat 
exports built for sharing this data outside the package.

## Identifiers and version

```{r core-summary-fn}
summarise_column <- function(data, colname) {
  x <- data[[colname]]
  n_missing <- sum(is.na(x))

  if (is.numeric(x)) {
    range_txt <- paste0(
      round(min(x, na.rm = TRUE), 2), "-", round(max(x, na.rm = TRUE), 2)
    )
    type_txt <- if (is.integer(x)) "integer" else "float"
  } else {
    lvls <- if (colname != "id") sort(unique(x[!is.na(x)])) else "[free text]"
    range_txt <- paste(lvls, collapse = ", ")
    type_txt <- "string"
  }

  tibble::tibble(
    Variable = colname,
    Type = type_txt,
    `Range / levels` = range_txt,
    `N missing` = n_missing
  )
}
```

```{r identifiers}
identifier_variables <- tibble::tribble(
  ~Variable,  ~Description,
  "id",       "Participant identifier, unique within a version.",
  "version",  "WM-FTT task version: v1 (original), v2 (added a parity-error penalty), v3 (added recall-order randomisation, independent front-end infrastructure).",
  "language", "UI language of the experiment in the browser."
)

identifier_summary <- purrr::map_dfr(
  identifier_variables$Variable, \(v) summarise_column(all_data, v)
)

identifier_variables |>
  dplyr::left_join(identifier_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

## Trial and item structure

```{r trial-structure}
trial_structure_variables <- tibble::tribble(
  ~Variable,        ~Description,
  "trial_number",   "Trial index. A trial is one encode-distract-recall cycle: 21 test trials in three blocks of seven, numbered 1-21, preceded by 3 training trials numbered 1-3 and one tutorial trial numbered 0.",
  "response_order", "Order in which the three response modalities were collected for this trial. Fixed in v1/v2 (always word/angle/colour); randomised per trial in v3.",
  "item_number",    "Item index, running 1-63 across the three test blocks (1-21, 22-42, 43-63) and 1-9 in training. Three items per trial; each item is one word shown at one orientation in one colour, so an item row carries all three features rather than one of them.",
  "expe_phase",     "Task phase: tutorial, training, or one of the three test blocks."
)

trial_structure_summary <- purrr::map_dfr(
  trial_structure_variables$Variable, \(v) summarise_column(all_data, v)
)

trial_structure_variables |>
  dplyr::left_join(trial_structure_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

## Stimulus content and response

```{r stimulus-content}
stimulus_variables <- tibble::tribble(
  ~Variable,               ~Description,
  "target_word",           "The to-be-remembered word.",
  "target_angle",          "The to-be-remembered rectangle orientation, in degrees.",
  "target_color_angle",    "The to-be-remembered colour, as an angle on the colour wheel.",
  "target_color",          "The to-be-remembered colour, as its nearest named colour.",
  "response_word",         "Participant's recalled word.",
  "response_angle",        "Participant's recalled orientation, in degrees.",
  "response_color_angle",  "Participant's recalled colour, as an angle on the colour wheel.",
  "response_color",        "Participant's recalled colour, as its nearest named colour.",
  "rt_word",               "Response time for the word recall response (ms).",
  "rt_angle",              "Response time for the orientation recall response (ms).",
  "rt_color",              "Response time for the colour recall response (ms).",
  "parity_1_stim",         "First parity-judgement distractor digit embedded in this trial.",
  "parity_1_resp",         "Participant's parity judgement for parity_1_stim.",
  "parity_1_acc",          "Accuracy of the first parity judgement (1 = correct).",
  "parity_1_rt",           "Response time for the first parity judgement (ms).",
  "parity_2_stim",         "Second parity-judgement distractor digit embedded in this trial.",
  "parity_2_resp",         "Participant's parity judgement for parity_2_stim.",
  "parity_2_acc",          "Accuracy of the second parity judgement (1 = correct).",
  "parity_2_rt",           "Response time for the second parity judgement (ms)."
)

stimulus_summary <- purrr::map_dfr(
  stimulus_variables$Variable, \(v) summarise_column(all_data, v)
)

stimulus_variables |>
  dplyr::left_join(stimulus_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

v2/v3 carry a real parity-error penalty in the task design; the parity
columns exist for v1 too, but the penalty was not applied at recall.

## Recall scores

These are the analysis variables. They are computed in R from the raw
target and response values by `compute_scores()`, not derived from the
task's own feedback columns, for reasons set out on the
[Scoring](scoring.html) page.

Two things to know before using them. Non-responses are scored 0 and
flagged by the `responded_*` columns, so any analysis of recall *accuracy*
should filter on those flags; leaving them out combines accuracy with
willingness to respond, which are different quantities. And word recall is
near ceiling in this task, with a split-half reliability of 0.45 against
0.82 for orientation and 0.83 for colour, so it does not support
individual-differences claims.

```{r recall-scores}
recall_score_variables <- tibble::tribble(
  ~Variable,          ~Description,
  "score_word",       "Word recall similarity, 1 = exact match. Edit-distance scoring [@gonthierEasyWayImprove2022]: Damerau-Levenshtein distance capped at the target's length, subtracted from it and divided by it. Reliability 0.45; see the caveat above.",
  "score_angle",      "Orientation recall similarity, 1 = exact match. Cosine of the angular error on a 180-degree period, since a rectangle's tilt is symmetric under 180-degree rotation. Reliability 0.82.",
  "score_color",      "Colour recall similarity, 1 = exact match. Cosine of the angular error on the 360-degree hue wheel. Reliability 0.83.",
  "score_word_z",     "`score_word`, z-scored within task version using experimental-block rows only. Raw and standardised are both kept: raw for absolute performance, standardised for cross-feature comparison.",
  "score_angle_z",    "`score_angle`, z-scored within task version.",
  "score_color_z",    "`score_color`, z-scored within task version.",
  "responded_word",   "Whether the participant entered a word at all.",
  "responded_angle",  "Whether the participant moved the orientation widget. Unlike the other two flags this is an *inference* rather than an observation: the widget starts at 90 degrees and returns that value if untouched, so a deliberate 90-degree response cannot be distinguished from no response. No experimental-block target lies within 20 degrees of 90, so such a response can never be near-correct.",
  "responded_color",  "Whether the participant moved the colour wheel. Non-response appears in the raw data as `response_color_angle == 999`."
)

recall_score_summary <- purrr::map_dfr(
  recall_score_variables$Variable, \(v) summarise_column(all_data, v)
)

recall_score_variables |>
  dplyr::left_join(recall_score_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

## Live feedback from the task display

**Not analysis variables.** These drove the feedback screen participants
saw between trials, and are kept for provenance and reproducibility only.
Note that `live_diff_*` are *dissimilarities* (0 = exact match), the
opposite direction from the `score_*` columns above.

`live_diff_word`, and therefore `feedback_score_word` and the word totals
derived from it, rest on a defective edit-distance implementation in the
task's front end: a loop whose condition is an assignment leaves the
distance matrix half-initialised, so the function does not compute the
distance its name claims. The [Scoring](scoring.html) page demonstrates
this against the data. Use `score_word` instead.

```{r feedback-columns}
feedback_variables <- tibble::tribble(
  ~Variable,                     ~Description,
  "live_diff_word",              "Front-end word dissimilarity, 0 = exact match. Not a valid edit distance; see above.",
  "live_diff_angle",             "Front-end orientation error, absolute difference divided by 180. Linear rather than cosine, with no handling of the rectangle's 180-degree symmetry.",
  "live_diff_color",            "Front-end colour error, circular difference divided by 180. Non-responses are mapped to 1 by an explicit guard.",
  "feedback_score_word",         "Points shown for the word response (0, 0.5 or 1), thresholded from `live_diff_word`.",
  "feedback_score_angle",        "Points shown for the orientation response (0, 0.5 or 1).",
  "feedback_score_color",        "Points shown for the colour response (0, 0.5 or 1).",
  "feedback_trial_score",        "Points shown for the whole trial: the three per-feature scores summed, minus 0.5 per incorrect parity judgement, floored at 0. Trial-level, so constant across a trial's three item rows.",
  "feedback_total_score_word",   "Cumulative word points shown at the end of the task.",
  "feedback_total_score_angle",  "Cumulative orientation points shown at the end of the task.",
  "feedback_total_score_color",  "Cumulative colour points shown at the end of the task.",
  "feedback_total_score",        "Combined total points shown to the participant. This is the number they were told to maximise, so it describes what they were optimising, but it is not a performance measure."
)

feedback_summary <- purrr::map_dfr(
  feedback_variables$Variable, \(v) summarise_column(all_data, v)
)

feedback_variables |>
  dplyr::left_join(feedback_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

Tutorial rows carry placeholder values in these columns: every
`live_diff_*` is exactly 1 and every `feedback_score_*` exactly 0,
regardless of what the participant did.

## Questionnaire scores

Raw questionnaire items are in the nested list-columns described at the end
of this page, not here: these are the derived, cross-version-comparable
scores.

```{r questionnaire-scores}
score_variables <- tibble::tribble(
  ~Variable,             ~Description,
  "vviq_total_score",    "Total score on the Vividness of Visual Imagery Questionnaire [@marksVisualImageryDifferences1973], summed across all 16 items. Lower scores indicate weaker/absent visual imagery.",
  "vviq_group_2",        "VVIQ group, 2 levels: aphantasia (<= 32), typical (> 32).",
  "vviq_group_4",        "VVIQ group, 4 levels: aphantasia (= 16), hypophantasia (17-32), typical (33-74), hyperphantasia (>= 75).",
  "nieq_mental_imagery", "NIEQ [@heaveyMeasuringFrequencyInnerExperience2019] mental imagery dimension score (mean of frequency and proportion items).",
  "nieq_inner_voice",    "NIEQ inner voice/speaking dimension score.",
  "nieq_emotions",       "NIEQ feelings/emotions dimension score.",
  "nieq_sensory_focus",  "NIEQ sensory awareness dimension score.",
  "nieq_unsymbolised",   "NIEQ unsymbolized thinking dimension score.",
  "object_mean",         "OSIVQ [@blazhenkovaNewObjectSpatialVerbal2009] object-imagery subscale mean.",
  "spatial_mean",        "OSIVQ spatial-imagery subscale mean.",
  "verbal_mean",         "OSIVQ verbal cognitive-style subscale mean.",
)

score_summary <- purrr::map_dfr(
  score_variables$Variable, \(v) summarise_column(all_data, v)
)

score_variables |>
  dplyr::left_join(score_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

**Item wording is not reproduced here**, as the VVIQ, OSIVQ, and NIEQ
item text is copyrighted by their respective authors - see
@marksVisualImageryDifferences1973, @blazhenkovaNewObjectSpatialVerbal2009,
and @heaveyMeasuringFrequencyInnerExperience2019 for the original items.

## Demographics

```{r demographics}
demographic_variables <- tibble::tribble(
  ~Variable,       ~Description,
  "age",           "Participant age in years, self-reported.",
  "gender",        "Gender, as freely typed by the participant.",
  "country",       "Self-reported country of residence.",
  "job",           "Self-reported occupation.",
  "education",     "Self-reported education level.",
  "field",         "Self-reported field of study or work.",
  "where_from",    "Self-reported source of how the participant heard about the study.",
  "prognosis",     "Self-reported aphantasia status/prognosis, where collected.",
  "treatment",     "Self-reported relevant treatment. NA throughout for v3 (not asked in that version's demographics form).",
  "neuro_trouble", "Self-reported neurological condition, freely typed. NA throughout for v3 (not asked in that version's demographics form)."
)

# country/job/education/field/where_from live in the extra_demographics
# nested column rather than top-level - unnest just those to summarise them.
demographics_unnested <- all_data |>
  tidyr::unnest_wider(extra_demographics, names_repair = "minimal")

demographic_summary <- purrr::map_dfr(
  demographic_variables$Variable, \(v) summarise_column(demographics_unnested, v)
)

demographic_variables |>
  dplyr::left_join(demographic_summary, by = "Variable") |>
  dplyr::select(Variable, Description, Type, `Range / levels`, `N missing`) |>
  knitr::kable()
```

`gender` and `neuro_trouble` are free text (French and English mixed for
`gender`; French for `neuro_trouble`), not a controlled vocabulary - the
table above lists the observed values as-is rather than collapsing them
into fixed categories.

## Nested item list-columns

Five list-columns hold the raw questionnaire and strategy-report items,
kept nested rather than forced into a common wide schema, because v1/v2 and
v3 don't ask exactly the same raw items (different front-end versions).
Unnest with `tidyr::unnest_wider()` to get the raw items as regular
columns.

```{r nested-columns}
nested_variables <- tibble::tribble(
  ~`List-column`,        ~Description,
  "vviq_items",          "Raw VVIQ item responses (vviq_q01-vviq_q16), 1-5 each. Same names in v1/v2 and v3.",
  "osivq_items", "Raw OSIVQ item responses. **Reverse-keyed items are stored inconsistently across versions:** v1 and v2 store `osivq_q02v`, `osivq_q09v`, `osivq_q41v` and `osivq_q42s` *unreversed*, while v3 stores them already reversed. The `object_mean`/`spatial_mean`/`verbal_mean` columns are correct in both; only these nested items differ. Recomputing subscale scores or reliabilities from the items without reversing them for v1/v2 gives wrong answers (verbal alpha comes out at 0.25 rather than 0.84).",
  "nieq_items",          "Raw NIEQ item responses: 2 items per dimension (frequency, proportion), both 0-100 sliders. Same names in v1/v2 and v3.",
  "strategy_items",      "Raw strategy-report answers. Mixed shape: v1/v2's colour/orientation/word strategy fields are free text (mostly French); a separate scoring-strategy field uses a small coded vocabulary, asked slightly differently in v1/v2 (3 legacy slots) vs v3 (2 slots).",
  "extra_demographics",  "Additional demographic items not promoted to top-level columns: country, job, education, field, where_from, language_native, language_usual."
)

knitr::kable(nested_variables)
```

For the full per-item breakdown - every raw column name, its exact range,
missingness, and (where applicable) coded vocabulary - see the codebook
that ships with `all_data_surveys` on 
[OSF](https://osf.io/3649s/files/osfstorage): a
participant-level export where these list-columns are unnested wide,
alongside `dataset_description.json`
([Psych-DS](https://psych-ds.github.io/)-compliant) and a PDF version of
the same information.

---

**Continuing through the Extended Online Report:** this page is a technical
reference, meant to be partially independent from the rest of the
vignettes. Get back to the [home page](https://m-delem.github.io/aphantasiaWMStrats/index.html).

---

## References

<div id="refs"></div>

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
