---
title: "Composition: how effort is divided, independently of how much there is"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Composition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)

# Models are loaded from their saved artefacts and never refitted here.
# See the analysis strategy page for how they were built.
pkg <- "aphantasiaWMStrats"
refit <- "never"
```

Two participants can score the same overall and have divided their effort
completely differently. This page is about that difference: **relative
allocation across the three features, with overall performance removed.**

This is the study's **primary analysis**, and the one it was designed
around. A composition is a different object from an accuracy score, with
its own geometry, and it answers the question the task was built to ask:
not *how well*, but *where*.

It is also where the strands meet. Imagery vividness, inner experience and
participants' own accounts of what they were doing all bear on the same
coordinate, in sections 4 to 6.

Everything here is v1 ([version scope](version-scope.html)) and uses
responded trials only, which carries a caveat set out in section 8.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    imagery_group = dplyr::first(vviq_group_2),
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

# Every participant with all three parts, and the subset clearing the
# engagement thresholds of the task validity page.
all_parts <- compose_features(v1, id) |>
  dplyr::filter(stats::complete.cases(dplyr::pick(tidyselect::everything())))

engaged_parts <- dplyr::filter(all_parts, id %in% engaged_ids(v1))

add_coordinates <- function(parts) {
  parts |>
    dplyr::bind_cols(
      ilr_coords(dplyr::select(parts, tidyselect::starts_with("part_")))
    ) |>
    dplyr::left_join(participant_info, by = "id") |>
    dplyr::mutate(
      complete_aphant = factor(
        dplyr::if_else(vviq == 16, "floor", "above_floor"),
        levels = c("above_floor", "floor")
      )
    )
}

# Three frames, named separately rather than one object reassigned. Each
# section states which it uses, and each is the frame the corresponding
# script fitted on. A single reassigned `model_data` is how a page ends up
# describing one sample and displaying another.
#
# Section 6 needs a composition and a strategy report and nothing else, so
# it keeps every engaged participant, with or without a vividness score.
# `inst/scripts/14-strategy-convergence.R` does the same, deliberately.
reported <- add_coordinates(engaged_parts)

# Sections 4, 7 and 8 additionally need vividness. This is the frame
# `inst/scripts/11-compositional-analysis.R` fits `comp-floor` on.
model_data <- dplyr::filter(reported, !is.na(vviq))

# Section 5 additionally needs the questionnaire scales. Standardised
# because the instruments sit on incompatible ranges (OSIVQ runs 1-5,
# NIEQ 0-100) -- and standardised **after** the join, on the participants
# who enter that model, exactly as
# `inst/scripts/15-allocation-and-style.R` does. A z-score computed on a
# different sample is a different predictor.
style_data <- model_data |>
  dplyr::inner_join(
    questionnaire_scales(all_data) |>
      dplyr::select(id, tidyselect::starts_with("osivq_"),
                    tidyselect::starts_with("nieq_")),
    by = "id"
  ) |>
  dplyr::mutate(
    dplyr::across(
      c("osivq_object", "osivq_spatial", "osivq_verbal",
        "nieq_inner_voice", "nieq_unsymbolised"),
      \(x) as.numeric(scale(x))
    )
  )
```

```{r samples}
tibble::tibble(
  frame = c("All three parts present", "Clearing the thresholds",
            "With a vividness score", "With the questionnaire scales"),
  used_in = c("Section 2", "Sections 2, 3 and 6",
              "Sections 4, 7 and 8", "Section 5"),
  participants = c(nrow(all_parts), nrow(engaged_parts),
                   nrow(model_data), nrow(style_data))
) |>
  knitr::kable(caption = "The frames this page uses, counted rather than stated")
```

## 1. What a composition is, and why proportions will not do

Each participant has three responders-only mean scores. Divide each by
their sum and you have a **composition**: three parts that sum to one,
carrying only relative information.

That closure is not a technicality. Three proportions summing to a
constant cannot vary independently, so a rise in one forces a fall in the
others, and ordinary correlations between them are not interpretable. The
[scoring](scoring.html) page shows what that does in practice:
residualising three parts on their own mean induces a correlation of about
-0.5 whatever the data says.

The standard solution is to leave the simplex. `ilr_coords()` applies an
isometric log-ratio transform, turning three constrained parts into two
unconstrained coordinates:

```{r ilr-demo}
head(model_data[, c("part_word", "part_angle", "part_color", "ilr1", "ilr2")],
     3) |>
  knitr::kable(digits = 3)
```

The two coordinates are contrasts, and **which contrasts is a choice**:

- `ilr1` is words against the geometric mean of orientations and colours,
  the verbal versus non-verbal balance.
- `ilr2` is colours against orientations.

## 2. The partition is a substantive choice, not a statistical one

Any two-coordinate log-ratio basis is a rotation of the same geometry. The
total variance, the distances between participants and the omnibus test are
**identical** whichever feature is contrasted first. Only the split between
the two coordinates moves.

```{r partition}
dplyr::bind_rows(
  partition_variance(all_parts, "All with three parts"),
  partition_variance(engaged_parts, "Clearing thresholds")
) |>
  dplyr::mutate(
    first = c(word = "Word", color = "Colour", angle = "Orientation")[first],
    total = var_ilr1 + var_ilr2
  ) |>
  knitr::kable(digits = 4,
               caption = "Variance by partition, and the invariant total")
```

Within each sample the total is constant, which is the invariance made
visible. So a partition should be chosen for the contrast it names, and
the verbal versus non-verbal one is what this study argues about.

There is a second, quieter lesson in that table. On the looser sample the
orientation-first partition appears to carry far more of the variance,
which would look like a reason to prefer it. On the engaged sample the
three are close. **The apparent asymmetry was six participants**, all of
whom answered very few orientation trials, and choosing a partition on
that basis would have meant choosing it on an artifact.

## 3. The composition barely varies

```{r spread}
composition_summary(engaged_parts, "Clearing thresholds") |>
  knitr::kable(digits = 3, caption = "The three parts")
```

Allocation is close to even and remarkably tight: each part sits near a
third with a standard deviation of about 0.025. Whatever this analysis
finds, it finds inside that range.

```{r ternary, fig.alt="Ternary diagram of the three-part composition, coloured by imagery group.", fig.height = 6}
ternary_data <- engaged_parts |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(imagery_group))

plot_composition_ternary(
  ternary_data,
  group = as.character(ternary_data$imagery_group),
  base_size = 16,
  # the default is sized for a print PDF; screen needs bigger marks
  point_size = 2.5
) +
  labs(caption = "Centred and scaled: the raw spread is very small.")
```

## 4. Does imagery predict allocation?

Both coordinates are modelled together, with the residual correlation
between them estimated. That is not cosmetic: the invariance in section 2
holds for a genuine multivariate model, and two separate fits would lose
it.

```{r model}
composition_formula("vviq + complete_aphant + parity_rate")

composition_model <- fit_brms_model(
  formula = composition_formula("vviq + complete_aphant + parity_rate"),
  data    = model_data,
  prior   = composition_priors(),
  file    = system.file("models", "comp-floor.rds", package = pkg),
  file_refit = refit
)

# `file_refit = "never"` returns the cached fit without checking that the
# data it was handed matches the data it was fitted on, so the check is
# made here instead of assumed.
stopifnot(nrow(model_data) == nrow(composition_model$data))
```

This runs on the `r nrow(model_data)` participants who clear the
engagement thresholds **and** have a vividness score.

Imagery vividness enters in the form the [analysis
strategy](analysis-strategy.html) page describes: a slope among everyone
above the scale floor, plus a single offset for participants at it, who
have no vividness variance among themselves.

```{r results}
knitr::kable(brms::fixef(composition_model), digits = 4,
             caption = "Population-level effects")
```

Read against a region of practical equivalence of a tenth of the outcome's
own standard deviation:

```{r rope}
comp_rope <- readRDS(system.file("results", "comp-rope.rds", package = pkg))

knitr::kable(comp_rope, caption = "Posterior against the ROPE")
```

```{r floor-row, include = FALSE}
# The row the prose below reads from, pulled out by name rather than by
# position so a reordering of the table cannot silently repoint it.
floor_ilr1 <- comp_rope[
  comp_rope$coordinate == "ilr1" & comp_rope$term == "complete_aphantfloor", ]
stopifnot(nrow(floor_ilr1) == 1)
```

**Participants at the imagery floor sit above the line on the verbal
coordinate.** The offset is `r fmt(floor_ilr1$Estimate)`, with 95% of the
posterior in `r floor_ilr1[["95% CrI"]]`, a standardised effect of
`r fmt(floor_ilr1$d, 2)` outcome SD, and
`r fmt(100 * floor_ilr1[["Above ROPE"]], 1)`% of the posterior above the
ROPE. In interpretable terms, their ratio of word allocation to the
geometric mean of the two non-verbal features is roughly
`r round(100 * (exp(floor_ilr1$Estimate / sqrt(2 / 3)) - 1))`% higher.

**Vividness above the floor does nothing at all.** Its slope lies entirely
inside the ROPE on both coordinates. Whatever is happening is a property of
the floor group rather than a gradient.

**On the colour versus orientation coordinate there is no result**, and it
is worth being precise about which kind of no result. The floor offset has
52% of its posterior below the ROPE, 19% inside and 30% above. That is not
evidence of absence, it is an uninformative posterior, and it should be
reported as such.

## 5. Is it imagery, or is it cognitive style?

The study was designed around a prediction from the Object-Spatial-Verbal
framework: that allocation should track **cognitive style**, and not
imagery vividness alone. A verbaliser should protect the word feature, a
spatialiser the orientation.

Section 4 tests only the vividness axis, which is the one the prediction
names as insufficient. This section tests the prediction itself.

This section runs on `r nrow(style_data)` participants rather than the
`r nrow(model_data)` of section 4: it needs the questionnaire scales, and
the `r nrow(model_data) - nrow(style_data)` participant dropped here has
NIEQ scores but no OSIVQ ones. Section 4's result is unaffected, being
fitted on the larger frame.

The step before that one is worth naming too, because both are easy to
conflate. `r nrow(engaged_parts)` participants clear the engagement
thresholds; `r nrow(engaged_parts) - nrow(model_data)` of them have no
VVIQ score, which is what takes section 4 to `r nrow(model_data)`.

**The predictor set was chosen from the questionnaires alone**, never from
how any of them relates to allocation, which is what keeps the choice
independent of the result. Scales correlating above 0.6 with VVIQ are
excluded as re-entering vividness under another name; scales with no
hypothesis attached are excluded whatever their correlation. That leaves
OSIVQ spatial and verbal, and the two NIEQ dimensions the
[psychometrics](psychometrics.html) and [beyond
vividness](beyond-vividness.html) pages identify as least redundant with
imagery.

```{r style-model}
style_model <- fit_brms_model(
  formula = composition_formula(
    paste("vviq + complete_aphant + parity_rate +",
          "osivq_spatial + osivq_verbal +",
          "nieq_inner_voice + nieq_unsymbolised")),
  data    = style_data,
  prior   = composition_priors(),
  file    = system.file("models", "alloc-full.rds", package = pkg),
  file_refit = refit
)

stopifnot(nrow(style_data) == nrow(style_model$data))

knitr::kable(
  brms::fixef(style_model)[, c("Estimate", "Q2.5", "Q97.5")],
  digits = 4, caption = "Allocation predicted by vividness, style and inner experience"
)
```

**The prediction is not supported.** OSIVQ verbal on the verbal coordinate
is 0.005 with 56% of its posterior inside the ROPE; OSIVQ spatial on the
colour-orientation coordinate is -0.001 with 55% inside. A verbaliser does
not measurably protect words, and a spatialiser does not measurably protect
orientations.

```{r offset-comparison}
# The same parameter from both fits, each against its own frame's outcome
# SD, so the ROPE means the same thing in both rows.
floor_offsets <- dplyr::bind_rows(
  dplyr::mutate(
    report_rope(
      brms::as_draws_df(composition_model)$b_ilr1_complete_aphantfloor,
      outcome_sd = stats::sd(model_data$ilr1)),
    model = "Section 4: vividness only"),
  dplyr::mutate(
    report_rope(
      brms::as_draws_df(style_model)$b_ilr1_complete_aphantfloor,
      outcome_sd = stats::sd(style_data$ilr1)),
    model = "Section 5: vividness, style and inner experience")
) |>
  dplyr::relocate(model)

knitr::kable(
  floor_offsets,
  caption = "The floor-group offset on the verbal coordinate, with and without the style predictors")
```

**The imagery effect is untouched by adding them.** The floor-group offset
is `r fmt(floor_offsets$Estimate[2])` `r floor_offsets[["95% CrI"]][2]`
here against `r fmt(floor_offsets$Estimate[1])`
`r floor_offsets[["95% CrI"]][1]` in section 4, with
`r fmt(100 * floor_offsets[["Above ROPE"]][2], 1)`% of the posterior beyond
the ROPE. Whatever drives allocation, it is not a style effect wearing a
vividness label — which is the outcome that would have required every
result on this site to be reread.

**One predictor does something, and the prediction did not name it.**

```{r inner-voice}
draws <- brms::as_draws_df(style_model)

dplyr::bind_rows(
  dplyr::mutate(
    report_rope(draws$b_ilr1_nieq_inner_voice,
                outcome_sd = stats::sd(style_data$ilr1)),
    term = "Inner voice"),
  dplyr::mutate(
    report_rope(draws$b_ilr1_nieq_unsymbolised,
                outcome_sd = stats::sd(style_data$ilr1)),
    term = "Unsymbolised thinking")
) |>
  dplyr::relocate(term) |>
  knitr::kable(caption = "The two inner-experience dimensions, on the verbal coordinate")
```

Reported **inner speech** predicts allocation on the verbal coordinate at
the same evidential standard the floor group clears, and in the opposite
direction to the obvious guess: more inner speech goes with **less**
allocation to words.

One reading, offered as a hypothesis rather than a conclusion: someone
whose inner speech is habitually occupied may have less phonological
capacity free for the memoranda, which would make this a competition effect
rather than a style effect. It is **exploratory** — the thesis's own
inner-experience prediction is labelled exploratory in the strict sense,
since the sample was not sized for it.

**Unsymbolised thinking, by contrast, contributes nothing here**, and that
is informative rather than empty. It correlates -0.40 with VVIQ and 0.43
with the floor indicator, so once vividness is in the model it has nothing
left to explain. It was kept in deliberately so the model could show that
rather than have it assumed. Note what this does *not* say: unsymbolised is
related to allocation marginally, and the [beyond
vividness](beyond-vividness.html) page shows it declining steadily across
the vividness range. Redundancy given vividness is a different claim from
no relationship.

Comparing the two predictor sets formally leans toward the simpler one
without settling it, by 3.8 elpd against a standard error of 3.0. At that
separation the data do not distinguish them, which is what `r nrow(style_data)` participants
and four extra predictors should be expected to produce.

## 6. Do participants know they are doing it?

Everything above is behaviour. After the task, participants were asked
which features they had tried to keep for points — a question about what
they were doing, not about what it felt like.

A strategy is by definition what someone takes themselves to be doing, so
no behavioural pattern can overrule their own account of it. What the
composition can offer is **convergence**: an independent measurement that
either tracks the same distinctions or does not.

```{r convergence}
participants <- dplyr::distinct(get_data("v1"), id, .keep_all = TRUE)

prioritisation <-
  dplyr::bind_rows(participants$strategy_items) |>
  dplyr::mutate(id = participants$id) |>
  dplyr::select(id, tidyselect::starts_with("strats_cfa_q04")) |>
  tidyr::pivot_longer(-id, names_to = "slot", values_to = "feature") |>
  dplyr::filter(!is.na(feature)) |>
  dplyr::summarise(kept = sum(feature != "none"), .by = id)

# `reported`, not `model_data`: this question needs a composition and a
# report, not a vividness score, and restricting it would publish a
# different number from the one `inst/scripts/14` computes.
convergence <- dplyr::inner_join(reported, prioritisation, by = "id") |>
  dplyr::filter(kept > 0)

convergence |>
  dplyr::summarise(
    participants = dplyr::n(),
    mean_ilr1 = mean(ilr1),
    .by = kept
  ) |>
  dplyr::arrange(kept) |>
  knitr::kable(digits = 3,
               caption = "Features participants said they kept, against what they did")
```

```{r convergence-test}
correlation_test(convergence$ilr1, convergence$kept) |>
  knitr::kable(digits = 4, caption = "Spearman correlation")
```

**They converge.** Participants who said they kept one feature really did
concentrate on words; those who said they kept all three really did spread
their effort, monotonically across the three levels.

That matters more than its p-value. The composition comes from 63 trials of
behaviour and the report comes from one question afterwards, so they are
**independent measurements of the same construct**, and their agreement
corroborates the coordinate as measuring what it claims. No amount of
internal reliability could establish that.

It also bears on a worry about the whole enterprise. If self-report about
one's own cognition were simply unreliable here, this is where it would
show. It does not — though note what was asked: *which features did you try
to keep* is a question about behaviour, which people can answer.
Whether the same holds for questions about phenomenology is a different
matter, and nothing here bears on it.

Two further details, reported in
`inst/scripts/14-strategy-convergence.R` rather than here. Naming colours
as well as words goes with a lower word share **and** a higher colour
share, so the agreement holds on the parts and not only on the summary
coordinate. And prioritisation is unrelated to imagery group, which makes
the report an independent measurement rather than a restatement of section
4.

## 7. What the coordinate cannot tell you

A ratio moved. It cannot say which part moved it.

```{r levels}
model_data |>
  dplyr::mutate(group = dplyr::if_else(complete_aphant == "floor",
                                       "At the VVIQ floor", "Above floor")) |>
  dplyr::summarise(
    participants = dplyr::n(),
    word = mean(mean_word), orientation = mean(mean_angle),
    colour = mean(mean_color),
    .by = group
  ) |>
  knitr::kable(digits = 3,
               caption = "Raw responders-only accuracy, not shares")
```

In levels, the floor group is marginally better on words and clearly worse
on orientations and colours. **The balance shifts toward words because the
non-verbal side falls, not because the verbal side rises.**

That distinction matters more than it might appear. "Aphantasic
participants rely more on verbal encoding" and "aphantasic participants are
worse at the non-verbal features" predict the same compositional result,
and this analysis cannot separate them. Doing so needs the absolute
accuracy models, which is exactly why the two framings are complementary
rather than competing.

One further caution. `ilr1` rests partly on word, and the [task
validity](task-validity.html) page shows word does not function as an
individual-differences measure. The *ratio* is stable where the *level* is
not, which is not a contradiction, but no claim here may treat word as a
well-measured quantity. Measurement noise in an outcome widens intervals
rather than biasing estimates, so if anything this result is conservative.

## 8. Two checks, and one caveat that is still open

**The functional form was not assumed.** Five ways of entering imagery were
compared, with the floor-group form pre-declared as primary regardless of
which won.

```{r comparison}
comparison <- readRDS(system.file("results", "comp-loo.rds", package = pkg))
if (!"model" %in% names(comparison)) {
  comparison <- tibble::rownames_to_column(as.data.frame(comparison), "model")
}
comparison |>
  dplyr::select(model, elpd_diff, se_diff) |>
  knitr::kable(digits = 2, caption = "Model comparison, first coordinate")
```

The forms with a discontinuity sit above those without one, by roughly one
to one and a half standard errors. That is weak, it was predicted to be
weak in advance, and it is reported rather than converted into a claim.

**The result survives at trial level.** Each trial is itself a
composition, so the same question can be asked with within-person noise
separated from between-person signal rather than averaged into it. Whether
that is possible at all is an empirical question about zeros, since a
log-ratio transform needs every part strictly positive:

```{r trial-frame}
trial_parts <- v1 |>
  dplyr::filter(id %in% model_data$id) |>
  dplyr::mutate(trial_uid = paste(expe_phase, trial_number, sep = "_")) |>
  compose_features(id, trial_uid, trial_number)

usable <- trial_parts |>
  dplyr::filter(
    stats::complete.cases(dplyr::pick(tidyselect::starts_with("part_"))),
    dplyr::if_all(tidyselect::starts_with("part_"), \(x) x > 0)
  )

tibble::tibble(
  compositions = nrow(trial_parts),
  usable = nrow(usable),
  share = nrow(usable) / nrow(trial_parts)
) |>
  knitr::kable(digits = 3,
               caption = "Trial-level compositions usable without zero replacement")
```

Dropped rather than imputed: there is no zero-replacement strategy worth
defending for that share of rows, and an unbalanced multilevel model
handles the loss natively.

```{r trial-level}
trial_data <- usable |>
  dplyr::bind_cols(
    ilr_coords(dplyr::select(usable, tidyselect::starts_with("part_")))
  ) |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    trial_c = as.numeric(scale(trial_number, scale = FALSE))
  )

trial_model <- fit_brms_model(
  formula = composition_formula(
    "vviq + complete_aphant + parity_rate + trial_c + (1 | p | id)"),
  data    = trial_data,
  prior   = composition_priors(),
  # The per-participant deviations are discarded; the `sd` and `sigma`
  # hyperparameters the intraclass correlation below needs are kept.
  save_pars = brms::save_pars(group = FALSE),
  file    = system.file("models", "comp-trial-multilevel.rds", package = pkg),
  file_refit = refit
)

stopifnot(nrow(trial_data) == nrow(trial_model$data))

knitr::kable(brms::fixef(trial_model)[, c("Estimate", "Q2.5", "Q97.5")],
             digits = 4, caption = "Trial-level model, population-level effects")
```

The floor offset on the verbal coordinate is larger there, and the interval
still excludes zero. These are **different estimands**, not one quantity
measured twice: the log-ratio of a mean is not the mean of the log-ratios.
Agreement between them is convergence rather than replication.

That model also gives a reliability estimate from the same fit as the
effects:

```{r icc}
readRDS(system.file("results", "comp-icc.rds", package = pkg)) |>
  knitr::kable(digits = 3, caption = "Between-person share of trial-level variance")
```

An intraclass correlation near 0.09 sounds alarming until it is aggregated:
at that level, a mean over 21 trials has a reliability near 0.70, which is
close to the 0.77 the [task validity](task-validity.html) page finds by
split-half. They are consistent, not contradictory.

**The caveat, now resolved and narrower than expected.** Every composition
here is built from trials a participant chose to answer, which conditions
on a selected subsample. The [joint model](joint-model.html) tested
whether that mattered, and the answer is **feature-specific**: willingness
and accuracy are related on orientation and unrelated on word and colour.

So one of the three parts is built on a selected subsample and the other
two are not. Both coordinates mix all three parts, so both carry the
caveat, but through one part rather than three. That is weaker than the
blanket warning this section used to carry, and it is a constraint on
interpretation rather than a reason to discount the result.

The same model also settles the question §1 raised about closure: between
people, accuracy on orientation and colour correlates **positively**, so
there is no between-person trade-off for closure to have been hiding.

---

**Continuing through the Extended Online Report:** this page follows [the analysis](analysis-strategy.html). To keep reading in order, continue to [performance](performance.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
