---
title: "Task engagement: two ways participants withdrew"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Task engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_minimal(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
pct <- function(x) sprintf("%.1f%%", 100 * x)
```

Participants could withdraw effort from the WM-FTT in two places: the
parity-judgement distractor between encoding and recall, and the recall
responses themselves. This page establishes that these are **separate
behaviours** rather than one underlying disengagement, describes each, and
sets out why neither can be treated as a nuisance to control away.

It is descriptive groundwork. The models that build on it live elsewhere;
what is settled here is what the variables are and how they relate.

```{r data}
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase), version == "v1")

# Per participant: parity accuracy across both judgements per trial, and
# non-response rate per recall feature.
by_participant <- block_trials |>
  dplyr::summarise(
    parity_accuracy = mean(c(parity_1_acc, parity_2_acc), na.rm = TRUE),
    non_response_word = 1 - mean(responded_word),
    non_response_orientation = 1 - mean(responded_angle),
    non_response_colour = 1 - mean(responded_color),
    vviq = dplyr::first(vviq_total_score),
    imagery_group = dplyr::first(vviq_group_2),
    .by = id
  )
```

## 1. The parity task had no consequence in v1

The scoring design was meant to make the distractor cost something: an
incorrect parity judgement deducts from the trial score. That penalty was
added in v2. **In v1 it is absent**, and since v1 is the primary analysis
sample, the distractor there was free.

```{r parity-penalty}
block_trials |>
  dplyr::summarise(
    feature_scores_summed = sum(feedback_score_word) + sum(feedback_score_angle) +
      sum(feedback_score_color),
    trial_score_shown = dplyr::first(feedback_trial_score),
    .by = c(id, expe_phase, trial_number)
  ) |>
  dplyr::summarise(
    trials = dplyr::n(),
    trials_where_they_match = sum(abs(feature_scores_summed - trial_score_shown) < 1e-9)
  ) |>
  knitr::kable(caption = "v1: the trial score is exactly the sum of the three feature scores")
```

Every trial. No deduction was ever applied, so a participant who worked out
that parity did not count lost nothing by ignoring it.

```{r parity-distribution}
ggplot(by_participant, aes(parity_accuracy)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  labs(x = "Parity judgement accuracy", y = "Participants",
       title = "Parity accuracy in v1, with chance marked")
```

```{r parity-summary}
by_participant |>
  dplyr::summarise(
    participants = dplyr::n(),
    median_accuracy = stats::median(parity_accuracy, na.rm = TRUE),
    at_or_below_chance = sum(parity_accuracy <= 0.5, na.rm = TRUE),
    above_ninety = sum(parity_accuracy > 0.9, na.rm = TRUE)
  ) |>
  knitr::kable(digits = 3, caption = "Parity engagement, v1")
```

The distribution is not unimodal noise around a competence level. It
separates people who did the task from people who did not, which is what
makes a binary split tempting and also what makes "accuracy" the wrong word
for it. In v1 this measures **compliance with a task that did not matter**,
not concurrent memory load.

## 2. Parity engagement is not independent of imagery

This is the finding that stops parity accuracy being usable as a nuisance
covariate.

```{r parity-vviq}
parity_vviq <- by_participant |>
  dplyr::filter(!is.na(parity_accuracy), !is.na(vviq))

tibble::tibble(
  n = nrow(parity_vviq),
  rho = stats::cor(parity_vviq$parity_accuracy, parity_vviq$vviq,
                   method = "spearman"),
  p = suppressWarnings(stats::cor.test(parity_vviq$parity_accuracy,
                                       parity_vviq$vviq,
                                       method = "spearman")$p.value)
) |>
  knitr::kable(digits = 4, caption = "Parity accuracy against VVIQ, v1")
```

Lower imagery vividness goes with *higher* parity accuracy. A covariate
that correlates with the grouping variable is not a nuisance: adjusting for
it removes part of the effect under study. Any model that includes parity
accuracy as a control needs to say what it thinks it is controlling for.

Why the correlation runs in that direction is not established here. A
resource-allocation account would predict it (effort not spent on imagery
is available for the distractor), but so would several duller explanations.

## 3. The second signal: declining to report a feature

Between `r pct(min(c(mean(1 - block_trials$responded_word), mean(1 - block_trials$responded_angle), mean(1 - block_trials$responded_color))))`
and `r pct(max(c(mean(1 - block_trials$responded_word), mean(1 - block_trials$responded_angle), mean(1 - block_trials$responded_color))))`
of v1 recall trials receive no response at all. The task encodes these with
sentinels rather than missing values, and the
[Scoring](scoring.html) page covers how they are handled.

```{r nonresponse-distribution}
by_participant |>
  tidyr::pivot_longer(tidyselect::starts_with("non_response_"),
                      names_to = "feature", values_to = "rate",
                      names_prefix = "non_response_") |>
  dplyr::mutate(feature = stringr::str_to_title(feature)) |>
  ggplot(aes(rate)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  facet_wrap(~feature) +
  labs(x = "Non-response rate", y = "Participants",
       title = "Most participants answer nearly everything; a few answer almost nothing")
```

## 4. The two are not the same behaviour

The natural assumption is that both are effort withdrawal, in which case
one variable would do. They are not.

```{r independence}
by_participant |>
  tidyr::pivot_longer(tidyselect::starts_with("non_response_"),
                      names_to = "feature", values_to = "non_response",
                      names_prefix = "non_response_") |>
  dplyr::filter(!is.na(parity_accuracy)) |>
  dplyr::summarise(
    rho = stats::cor(non_response, parity_accuracy, method = "spearman"),
    p = suppressWarnings(stats::cor.test(non_response, parity_accuracy,
                                         method = "spearman")$p.value),
    .by = feature
  ) |>
  dplyr::mutate(feature = stringr::str_to_title(feature)) |>
  knitr::kable(digits = 3,
               caption = "Feature non-response against parity accuracy, v1")
```

The correlations are weak, mostly non-significant, and **positive**:
participants who engaged more with the parity task declined *more* recall
responses, which is the opposite of what a shared-disengagement account
predicts. Two behaviours, not one.

That has a practical consequence. A single "engagement" covariate would
average two things that point in different directions, and would control
for neither properly.

## 5. What this means downstream

Neither variable is a nuisance to be partialled out, and both correlate
with the grouping variable of interest.

- **Parity accuracy** in v1 measures compliance with a consequence-free
  task, and correlates with VVIQ at
  `r fmt(stats::cor(parity_vviq$parity_accuracy, parity_vviq$vviq, method = "spearman"), 3)`.
  Using it as a control silently adjusts away part of the imagery effect.
- **Non-response** is folded into the recall scores as zeros, which means
  any analysis of accuracy that ignores the `responded_*` flags is
  measuring a blend of accuracy and willingness to answer. The
  [Scoring](scoring.html) page quantifies how much difference that makes.

The honest framing is that participants had two independent ways of not
doing the task, both are measurable, and both carry information about the
people doing them. Treating either as measurement error discards that.

## References
