---
title: "Task engagement: two ways of not doing the task"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Task engagement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
pct <- function(x) sprintf("%.1f%%", 100 * x)
```

Participants could decline to do the WM-FTT in two places: the
parity-judgement distractor between encoding and recall, and the recall
responses themselves. This page establishes what each of those variables
measures, shows that they are **separate behaviours**, and argues that
neither is a nuisance to be controlled away.

The second claim is the one the modelling pages depend on. Under a
points-per-feature incentive with optional recall, **declining to report a
feature is an allocation decision**, not missing data.

It is descriptive groundwork. The models that build on it live elsewhere.

```{r data}
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase), version == "v1")

by_participant <- block_trials |>
  dplyr::summarise(
    # Parity is split into its two components. Section 2 shows why the
    # `parity_*_acc` columns cannot be used directly.
    parity_probes   = 2 * dplyr::n(),
    parity_answered = sum(responded_parity_1) + sum(responded_parity_2),
    parity_correct  = sum(parity_1_acc) + sum(parity_2_acc),
    non_response_word = 1 - mean(responded_word),
    non_response_orientation = 1 - mean(responded_angle),
    non_response_colour = 1 - mean(responded_color),
    vviq = dplyr::first(vviq_total_score),
    imagery_group = dplyr::first(vviq_group_2),
    .by = id
  ) |>
  dplyr::mutate(
    parity_rate = parity_answered / parity_probes,
    parity_accuracy_given_response =
      dplyr::if_else(parity_answered > 0, parity_correct / parity_answered,
                     NA_real_),
    # what a naive mean of `parity_*_acc` gives, kept for section 2
    parity_accuracy_naive = parity_correct / parity_probes
  )

engaged <- engaged_ids(block_trials)
```

## 1. The parity task had no consequence in v1

The scoring design was meant to make the distractor cost something: an
incorrect parity judgement deducts from the trial score. That penalty was
added in v2. **In v1 it is absent**, and since v1 is the primary analysis
sample, the distractor there was free.

```{r parity-penalty}
block_trials |>
  dplyr::summarise(
    feature_scores_summed = sum(feedback_score_word) + sum(feedback_score_angle) +
      sum(feedback_score_color),
    trial_score_shown = dplyr::first(feedback_trial_score),
    .by = c(id, expe_phase, trial_number)
  ) |>
  dplyr::summarise(
    trials = dplyr::n(),
    trials_where_they_match = sum(abs(feature_scores_summed - trial_score_shown) < 1e-9)
  ) |>
  knitr::kable(caption = "v1: the trial score is exactly the sum of the three feature scores")
```

Every trial. No deduction was ever applied, so a participant who worked out
that parity did not count lost nothing by ignoring it.

Some of them worked it out.

```{r parity-optout}
optout <- dplyr::filter(by_participant, parity_rate == 0)

tibble::tibble(
  participants = nrow(by_participant),
  answered_no_parity_probe = nrow(optout),
  of_those_clearing_recall_thresholds = sum(optout$id %in% engaged)
) |>
  knitr::kable(caption = "Participants who ignored the distractor entirely, v1")
```

`r nrow(optout)` of `r nrow(by_participant)` participants answered **no
parity probe at all**, and `r sum(optout$id %in% engaged)` of those clear
the per-feature engagement thresholds on the recall task. That is not
withdrawal. Under v1's incentives, abandoning a consequence-free secondary
task while keeping recall performance intact is the rational move.

## 2. What the parity accuracy column actually measures

`parity_1_acc` and `parity_2_acc` score an **unanswered** probe as 0, the
same convention the recall scores use. That makes a zero ambiguous, and in
v1 the ambiguity is close to total.

```{r parity-zeros}
tibble::tibble(
  zeros = sum(block_trials$parity_1_acc == 0),
  unanswered = sum(block_trials$parity_1_acc == 0 & !block_trials$responded_parity_1),
  answered_and_wrong = sum(block_trials$parity_1_acc == 0 & block_trials$responded_parity_1)
) |>
  knitr::kable(caption = "Zeros on parity_1_acc, v1: what they are")
```

Most of them are probes nobody answered. A participant mean of that column
is therefore not an accuracy measure. Split into its two components, the
two barely relate to each other:

```{r parity-components}
components <- dplyr::filter(by_participant, !is.na(parity_accuracy_given_response))

tibble::tibble(
  quantity = c("Naive mean of parity_*_acc", "Proportion of probes answered",
               "Accuracy given a response"),
  median = c(stats::median(by_participant$parity_accuracy_naive),
             stats::median(by_participant$parity_rate),
             stats::median(components$parity_accuracy_given_response)),
  at_exactly_zero = c(sum(by_participant$parity_accuracy_naive == 0),
                      sum(by_participant$parity_rate == 0), 0L)
) |>
  knitr::kable(digits = 3, caption = "Three quantities, v1")

tibble::tibble(
  pair = c("Naive mean vs proportion answered",
           "Naive mean vs accuracy given a response",
           "Proportion answered vs accuracy given a response"),
  r = c(stats::cor(by_participant$parity_accuracy_naive, by_participant$parity_rate),
        stats::cor(components$parity_accuracy_naive,
                   components$parity_accuracy_given_response),
        stats::cor(components$parity_rate,
                   components$parity_accuracy_given_response))
) |>
  knitr::kable(digits = 3, caption = "What the naive column is made of")
```

The naive column is almost perfectly correlated with the proportion of
probes answered, and weakly with accuracy among those answered. It is a
response-rate variable wearing an accuracy name, and that is the whole
explanation for its bimodality. Meanwhile the two real components are close
to orthogonal, so collapsing them does not blur two related quantities: it
discards one.

```{r parity-distribution}
by_participant |>
  dplyr::select(id, parity_rate, parity_accuracy_given_response) |>
  tidyr::pivot_longer(-id, names_to = "quantity", values_to = "value") |>
  dplyr::mutate(quantity = factor(
    dplyr::recode(quantity,
      parity_rate = "Proportion of probes answered",
      parity_accuracy_given_response = "Accuracy given a response"),
    # narrative order, not alphabetical
    levels = c("Proportion of probes answered", "Accuracy given a response")
  )) |>
  ggplot(aes(value)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  # chance is only a reference point for the accuracy panel; drawing it on
  # a response rate would imply something it does not mean
  geom_vline(
    data = tibble::tibble(quantity = factor("Accuracy given a response",
      levels = c("Proportion of probes answered", "Accuracy given a response"))),
    aes(xintercept = 0.5), linetype = "dashed"
  ) +
  facet_wrap(~quantity) +
  labs(x = NULL, y = "Participants",
       title = "The two components behave completely differently",
       caption = "Dashed line marks chance on the parity judgement.")
```

Whether someone did the distractor is bimodal, with a hard spike at zero.
How well they did it, among those who did, is unimodal and well above the
chance line. Only the second is an accuracy variable.

### The claim this section used to make

An earlier version of this page reported a correlation between the naive
column and VVIQ and concluded that it was "the finding that stops parity
accuracy being usable as a nuisance covariate". Both halves of that were
too strong.

```{r parity-vviq}
usable <- dplyr::filter(by_participant, !is.na(vviq))

naive_vviq <- correlation_test(usable$parity_accuracy_naive, usable$vviq)

dplyr::bind_rows(
  dplyr::mutate(naive_vviq, quantity = "Naive mean of parity_*_acc"),
  dplyr::mutate(correlation_test(usable$parity_rate, usable$vviq),
                quantity = "Proportion of probes answered"),
  dplyr::mutate(
    correlation_test(usable$parity_accuracy_given_response, usable$vviq),
    quantity = "Accuracy given a response")
) |>
  dplyr::select(quantity, n, rho, p) |>
  knitr::kable(digits = 4, caption = "Parity against VVIQ, v1")
```

The original figure was rho = `r fmt(naive_vviq$rho)` with p =
`r fmt(naive_vviq$p)`, which is not evidence for a direction, and it came
from a variable that measures willingness rather than accuracy. Neither
corrected component does any better. **Parity engagement is not related to
imagery vividness in this sample.**

## 3. The second signal: declining to report a feature

Between `r pct(min(c(mean(1 - block_trials$responded_word), mean(1 - block_trials$responded_angle), mean(1 - block_trials$responded_color))))`
and `r pct(max(c(mean(1 - block_trials$responded_word), mean(1 - block_trials$responded_angle), mean(1 - block_trials$responded_color))))`
of v1 recall trials receive no response at all. The task encodes these with
sentinels rather than missing values, and the
[Scoring](scoring.html) page covers how they are handled.

```{r nonresponse-distribution}
by_participant |>
  tidyr::pivot_longer(tidyselect::starts_with("non_response_"),
                      names_to = "feature", values_to = "rate",
                      names_prefix = "non_response_") |>
  dplyr::mutate(feature = stringr::str_to_title(feature)) |>
  ggplot(aes(rate)) +
  geom_histogram(binwidth = 0.05, boundary = 0) +
  facet_wrap(~feature) +
  labs(x = "Non-response rate", y = "Participants",
       title = "Most participants answer nearly everything; a few answer almost nothing")
```

## 4. The two are not the same behaviour

The natural assumption is that both are effort withdrawal, in which case
one variable would do. They are not.

```{r independence}
by_participant |>
  tidyr::pivot_longer(tidyselect::starts_with("non_response_"),
                      names_to = "feature", values_to = "non_response",
                      names_prefix = "non_response_") |>
  dplyr::summarise(
    rho = stats::cor(non_response, parity_rate, method = "spearman"),
    p = suppressWarnings(stats::cor.test(non_response, parity_rate,
                                         method = "spearman")$p.value),
    .by = feature
  ) |>
  dplyr::mutate(feature = stringr::str_to_title(feature)) |>
  knitr::kable(digits = 3,
               caption = "Feature non-response against parity engagement, v1")
```

The correlations are near zero and inconsistent in sign. Ignoring the
distractor tells you essentially nothing about whether someone will answer
a recall prompt, which is what a shared-disengagement account would need.
**Two behaviours, not one.**

That has a practical consequence. A single "engagement" covariate would
average two unrelated things and control for neither.

## 5. What this means downstream

**Parity engagement is not a confound.** It does not differ by imagery
group:

```{r parity-not-confound}
floor_split <- by_participant |>
  dplyr::filter(!is.na(vviq), id %in% engaged) |>
  dplyr::mutate(floor = dplyr::if_else(vviq == 16, "VVIQ floor", "Above floor"))

floor_split |>
  dplyr::summarise(
    participants = dplyr::n(),
    mean_parity_rate = mean(parity_rate),
    .by = floor
  ) |>
  knitr::kable(digits = 3,
               caption = "Parity engagement by imagery group, engaged sample")
```

```{r parity-wilcox}
stats::wilcox.test(parity_rate ~ floor, data = floor_split, exact = FALSE)
```

It is still worth carrying in the accuracy models, but as the **dual-task
load a participant actually incurred**, which is a strategy variable in its
own right, rather than as a control for something it does not predict.

**Non-response is the substantive one.** It is folded into the recall
scores as zeros, so any analysis of accuracy that ignores the `responded_*`
flags is measuring a blend of accuracy and willingness to answer. The
[Scoring](scoring.html) page quantifies how much difference that makes.

More than that: under a points-per-feature incentive with optional recall,
choosing not to report a feature **is** an allocation decision, and
arguably a more direct one than how accurately a reported feature was
recalled. Treating it as missing data discards the clearest strategic
signal the task produces. That is why the primary model of this study
models whether a feature was reported and how well it was reported
together, rather than conditioning on the first and analysing only the
second.

---

**Continuing through the Extended Online Report:** this page follows [scoring](scoring.html). To keep reading in order, continue to [versions](version-scope.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
