---
title: "Implementation notes"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Implementation notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

The conventions behind every model on this site, collected so they can be
checked rather than inferred. None of it is a result; all of it is the kind
of thing that determines whether the results mean what they appear to.

## 1. Fitting defaults

Every model is fitted through `fit_brms_model()`, a thin wrapper over
`brms::brm()` that fixes the conventions below so they cannot drift between
scripts.

Four chains, 2000 **post-warmup** draws each after 1000 warmup iterations,
so `brm()` receives `iter = 3000, warmup = 1000` and keeps 8000 draws in
total. The argument below is named `iterations` rather than `iter` for
exactly that reason: it is draws per chain, not a total, and reading it as
brms' `iter` gives 4000. The joint model additionally uses
`adapt_delta = 0.99` and `max_treedepth = 15`, set from the start rather
than in response to warnings: a model with fifteen correlated random
intercepts from 86 participants is expected to need them.

```{r wrapper}
args(fit_brms_model)
```

## 2. Models are objects, not code {#formulas}

Formulas and priors are **exported functions**, not written inline in the
scripts that fit them.

```{r formula}
joint_formula()
```

This is not tidiness. Documentation pages load fitted models with
`file_refit = "never"`, and brms then returns the cached fit **without
checking** that the formula it was handed matches the one the model was
fitted with. A page that re-typed its formula could display something that
is not what produced its numbers, silently and permanently. Calling the
same builder the script called makes that impossible.

The response order matters for the same reason. It determines the generated
Stan code, so a formula built in a different order is a different model as
far as caching is concerned. An early version of the joint model appended
word's accuracy arm last, an artifact of a conditional block; that fit was
discarded and refitted in the documented order.

## 3. Priors are set per coefficient {#priors}

```{r priors}
response_priors("scoreangle")
```

Not per class. On the logit scale a vividness slope acts over a 64-point
range and a binary group offset does not, so a single `class = "b"` prior
would treat them as the same quantity. In a multivariate model `resp` is
required as well: `class = "b"` alone matches no parameter, and recent brms
versions reject it outright.

`lkj(4)` for the joint model's correlation matrix, against `lkj(2)`
elsewhere, because fifteen correlations from 86 participants need more
regularisation than three do. How strong that is depends on the dimension,
which `lkj_marginal()` makes checkable:

```{r lkj}
tibble::tibble(
  eta = c(2, 4),
  `marginal SD on a 6x6 matrix` = vapply(
    c(2, 4), \(eta) stats::sd(lkj_marginal(1e5, eta, 6)), numeric(1))
) |>
  knitr::kable(digits = 3)
```

Every correlation reported on the [joint model](joint-model.html) page
carries a `moved` flag for whether its posterior is narrower than that
marginal. A correlation that did not move off its prior is not a finding.
The [performance](performance.html) page carries the same flag against the
`lkj(2)` marginal, which is the weaker comparison of the two.

## 4. ROPE conventions {#rope}

A region of practical equivalence needs a scale. The convention here is a
tenth of the outcome's own standard deviation, applied through
`report_rope()`, so the region moves with the outcome rather than being a
fixed number that means different things on different scales.

That matters most on the compositional coordinates, whose standard
deviations are near 0.09: a ROPE specified in raw units would either
swallow every effect or none.

## 5. Response families, per feature {#families}

Three features, three families, each following from what the boundary of
that scale means rather than from convenience.

```{r families}
tibble::tribble(
  ~feature, ~family, ~why,
  "Word", "Zero-one-inflated Beta",
  "91% at exactly 1 and 3% at exactly 0 are real point masses",
  "Orientation", "Beta", "No boundary mass once non-responders are removed",
  "Colour", "Beta after a squeeze",
  "Boundary mass is pixel resolution on a continuous wheel, not behaviour"
) |>
  knitr::kable()
```

The squeeze is Smithson and Verkuilen's, implemented in
`squeeze_boundaries()`. It moves every value by about one ten-thousandth at
this sample size, which is the point: enough to make a Beta likelihood
defined, small enough not to be a modelling decision in itself.

The [model diagnostics](model-diagnostics.html) page checks that all three
describe the data.

## 6. The MARS knot pre-check {#mars}

Before fitting any segmented model, `mars_knots()` asks whether a hinge is
identifiable at all, so that a model which would only report its own prior
is not fitted.

Two things it has to get right, both of which were got wrong first.

**Knots come off the pruned model**, not off `$cuts`, which also holds
candidate terms the backward pass discarded. Reading `$cuts` whole reported
five knots where the fitted model had one.

**It runs on one row per participant.** The question is the shape of a
between-person relationship. Run on items, each participant contributes
about 63 rows carrying the same vividness value, which inflates the sample
63-fold and places knots in trial-level noise while the cross-validated fit
stays near zero.

Its silence means no knot survives GCV pruning, not that the predictor is
unrelated to the outcome. Pruning is not a significance test, and on pure
noise MARS still retains hinge terms: what separates signal from noise is
the cross-validated fit. The gate is therefore **GRSq > 0**: a knot is only
used to seed a segmented model if the generalised R-squared is positive,
since a knot with GRSq at or below zero describes a fit that does not
generalise beyond the sample it was found in.

## 7. What is saved, and what is not {#save-pars}

Fitted models are cached to `inst/models/` and loaded by the pages. Three
of them are saved with `save_pars(group = FALSE)`, which discards the
per-participant random-effect draws: `joint-gates`
(`inst/scripts/09`), `perf-joint-orientation` (`10`) and
`comp-trial-multilevel` (`11`).

Those draws are about 89% of the parameters in a fitted model and nothing
on any page reads them: the pages report fixed effects, the correlation
matrix, and `posterior_epred()` at the population level. The `sd` and `cor`
hyperparameters that the correlations and the intraclass correlation come
from are hyperparameters, not group-level draws, and they survive.

One model is deliberately excluded from that: the per-feature performance
model, because the shrinkage figure on the
[performance](performance.html) page calls `brms::posterior_epred()` on it
at the default `re_formula`, which needs exactly the group-level draws
that would be discarded.

## 8. Two recurring hazards {#hazards}

Both bit more than once, and both are silent.

**Column masking inside `tibble()`.** Later expressions see earlier
columns. Writing `participants = nrow(participants)` and then referring to
`participants` gives the count, not the frame. This produced a table of
p-values that were all exactly 1, because a Wilcoxon test was comparing two
group means to each other rather than two samples. It never errors when the
masked value happens to be a valid input.

**Unanchored column selection.** A pattern like `score_(word|angle|color)`
also matches `score_color_raw`, which silently poisoned a set of
participant means to `NA`. The selectors in `compose_features()` are
anchored for that reason.

---

**Continuing through the Extended Online Report:** this page follows [model diagnostics](model-diagnostics.html). To keep reading in order, continue to [what was tried and withdrawn](lessons.html) next. Or return to [the analysis](analysis-strategy.html) for what the models were built to do.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
