---
title: "Implementation notes"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Implementation notes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

This page gathers the conventions behind every model on this site.

## 1. Fitting defaults {#impl-defaults}

Every model is fitted through `fit_brms_model()`, a thin wrapper over
`brms::brm()` that fixes the conventions below so they cannot drift between
scripts.

Four chains, 2000 **post-warmup** draws each after 1000 warmup iterations,
so `brm()` receives `iter = 3000, warmup = 1000` and keeps 8000 draws in
total. The argument below is named `iterations` instead of `iter` for
that reason: it is draws per chain, not a total, and reading it as
brms' `iter` gives 4000. The joint model additionally uses
`adapt_delta = 0.99` and `max_treedepth = 15`, set from the start: 
a model with fifteen correlated random intercepts from 86 participants 
is expected to need them.

```{r wrapper}
args(fit_brms_model)
```

## 2. Models are objects, not code {#impl-formulas}

Formulas and priors are **exported functions**, not written inline in the
scripts that fit them.

```{r formula}
joint_formula()
```

Documentation pages load fitted models with `file_refit = "never"`, and brms 
then returns the cached fit **without checking** that the formula it was handed 
matches the one the model was
fitted with. A page that re-typed its formula could display something that
is not what produced its numbers, silently and permanently. Calling the
same builder the script called makes that impossible.

## 3. Priors are set per coefficient {#impl-priors}

```{r priors}
response_priors("scoreangle")
```

... Not per class. On the logit scale a vividness slope acts over a 64-point
range and a binary group offset does not, so a single `class = "b"` prior
would treat them as the same quantity. In a multivariate model `resp` is
required as well: `class = "b"` alone matches no parameter, and recent brms
versions reject it outright.

`lkj(4)` is used for the joint model's correlation matrix, against `lkj(2)`
elsewhere, because fifteen correlations from 86 participants need more
regularisation than three do. How strong that is depends on the dimension,
which `lkj_marginal()` makes checkable:

```{r lkj}
tibble::tibble(
  eta = c(2, 4),
  `marginal SD on a 6x6 matrix` = vapply(
    c(2, 4), \(eta) stats::sd(lkj_marginal(1e5, eta, 6)), numeric(1))
) |>
  knitr::kable(digits = 3)
```

Every correlation reported on the [joint model](joint-model.html) page
carries a `moved` flag for whether its posterior is narrower than that
marginal. A correlation that did not move off its prior is not a finding.
The [performance](performance.html) page carries the same flag against the
`lkj(2)` marginal.

## 4. ROPE conventions {#impl-rope}

A region of practical equivalence needs a scale. The convention here is a
tenth of the outcome's own standard deviation, applied through
`report_rope()`, so the region moves with the outcome rather than being a
fixed number that means different things on different scales.

That matters most on the compositional coordinates, whose standard
deviations are near 0.09: a ROPE specified in raw units would either
swallow every effect or none.

## 5. Response families, per feature {#impl-families}

Three features, three families, each following from what the boundary of
that scale means rather than from convenience.

```{r families}
tibble::tribble(
  ~feature, ~family, ~why,
  "Word", "Zero-one-inflated Beta",
  "91% at exactly 1 and 3% at exactly 0 are real point masses",
  "Orientation", "Beta", "No boundary mass once non-responders are removed",
  "Colour", "Beta after a squeeze",
  "Boundary mass is pixel resolution on a continuous wheel, not behaviour"
) |>
  knitr::kable()
```

The squeeze is Smithson and Verkuilen's [-@smithsonBetterLemonSqueezer2006], implemented in
`squeeze_boundaries()`. It moves every value by about one ten-thousandth at
this sample size, which is the point: enough to make a Beta likelihood
defined, small enough not to be a modelling decision in itself.

The [model diagnostics](model-diagnostics.html) page checks that all three
describe the data.

## 6. The MARS knot pre-check {#impl-mars}

Before fitting any segmented model, `mars_knots()` asks whether a hinge is
identifiable at all, so that a model which would only report its own prior
is not fitted.

Two things it has to get right:

**Knots come off the pruned model**, not off `$cuts`, which also holds
candidate terms the backward pass discarded. Reading `$cuts` whole reported
five knots where the fitted model had one.

**It runs on one row per participant.** The question is the shape of a
between-person relationship. Run on items, each participant contributes
about 63 rows carrying the same vividness value, which inflates the sample
63-fold and places knots in trial-level noise while the cross-validated fit
stays near zero.

Its silence means no knot survives GCV pruning, not that the predictor is
unrelated to the outcome. Pruning is not a significance test, and on pure
noise MARS still retains hinge terms: what separates signal from noise is
the cross-validated fit. The gate is therefore **GRSq > 0**: a knot is only
used to seed a segmented model if the generalised R-squared is positive,
since a knot with GRSq at or below zero describes a fit that does not
generalise beyond the sample it was found in.

## 7. What is saved, and what is not {#impl-save-pars}

Fitted models are cached to `inst/models/` and loaded by the pages. Three
of them are saved with `save_pars(group = FALSE)`, which discards the
per-participant random-effect draws: `joint-gates`
(`inst/scripts/09`), `perf-joint-orientation` (`10`) and
`comp-trial-multilevel` (`11`).

Those draws are about 89% of the parameters in a fitted model and nothing
on any page reads them: the pages report fixed effects, the correlation
matrix, and `posterior_epred()` at the population level. The `sd` and `cor`
hyperparameters that the correlations and the intraclass correlation come
from are hyperparameters, not group-level draws, and they survive.

One model is deliberately excluded from that: the per-feature performance
model, because the shrinkage figure on the
[performance](performance.html) page calls `brms::posterior_epred()` on it
at the default `re_formula`, which needs exactly the group-level draws
that would be discarded.

---

**Continuing through the Extended Online Report:** this page follows [model diagnostics](model-diagnostics.html). To keep reading in order, continue to [what was tried and withdrawn](lessons.html) next. Or return to [the analysis](analysis-strategy.html) for what the models were built to do.

---

## References {#impl-refs}

::: {#refs}
:::

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
