---
title: "The joint model: reporting and accuracy together"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{The joint model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5.5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
library(patchwork)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)

pkg <- "aphantasiaWMStrats"
refit <- "never"
```

The [composition](composition.html) page reports this study's primary
result, and it computes it from **responded trials only**. Whether that
conditioning was innocent cannot be checked from inside a responders-only
analysis, because the information needed to check it has been removed.

This model checks it, and answers two questions of its own that no other
page can reach: whether willingness to report is itself a stable
individual trait, and whether doing well on one feature costs another
between people.

Six responses. Three **gates**, one per feature, for whether it was
reported at all. Three **accuracies**, for how well it was reported when it
was. All six share a participant-level correlation matrix, and those
fifteen correlations are the point.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

model_data <- v1 |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq)) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    score_color = squeeze_boundaries(score_color, n = sum(responded_color))
  )

joint_model <- fit_brms_model(
  formula = joint_formula(),
  data    = model_data,
  prior   = joint_priors(),
  file    = system.file("models", "joint-full.rds", package = pkg),
  file_refit = refit
)
```

## 1. The sample is larger here, deliberately

```{r sample}
engaged <- engaged_ids(v1)

# `model_data` already excludes participants with no vividness score, so
# this is "clears the thresholds AND has a VVIQ" -- the sample the
# composition and performance pages model.
comparison_sample <- dplyr::n_distinct(
  dplyr::filter(model_data, id %in% engaged)$id
)
```

`r length(engaged)` participants clear the engagement thresholds, and
`r comparison_sample` of those also have a vividness score: that is the
sample every other modelling page uses. This one uses all
`r dplyr::n_distinct(model_data$id)` v1 participants with a vividness
score, thresholds or no. [The analysis](analysis-strategy.html#samples)
sets out the whole chain in one table.

The thresholds exist to stop imprecise participant *means* being treated as
measurements, and this model forms no participant means. More to the point,
the participants they exclude are the ones with the most to say about
abstention: one answered **zero** orientation trials, which makes them
invisible to every responders-only analysis and highly informative here.

```{r specification}
joint_formula()
```

The `subset()` terms are what let one trial carry six responses with
independent non-response, and what let a participant contribute to a gate
whether or not they ever answered that feature. `parity_rate` is on the
accuracy arms only: it is the dual-task load a participant actually
incurred, and the [task engagement](engagement.html) page shows it does
not predict whether they respond.

The fifteen correlations are the object of interest, in four families:

| Family | Question | Reported in |
|---|---|---|
| Response and accuracy, within a feature | Was conditioning on responding legitimate, and is the effect feature-specific or a general engagement trait? | §2 |
| Accuracy and accuracy, across features | Does doing well on one feature cost another, between people? | §3 |
| Response and response, across features | Do people who skip one feature skip others? | §5 |
| The six cross terms | Everything else | §5 |

Which of them are foregrounded was fixed in advance, in three tiers, so
that choosing among fifteen after seeing them is not possible. The six in
the first two families were pre-specified; the nine in §5 were not, and
are labelled exploratory there.

## 2. Was analysing responded trials only legitimate?

Three correlations answer that, one per feature. The rule for reading them
was fixed before they existed: near zero everywhere would mean
responders-only analyses were fine; large everywhere would mean a general
engagement trait and a caveat on everything conditional.

```{r correlations}
labels <- c(respondedword = "Word gate", respondedangle = "Orientation gate",
            respondedcolor = "Colour gate", scoreword = "Word accuracy",
            scoreangle = "Orientation accuracy", scorecolor = "Colour accuracy")

correlations <- posterior_correlations(
  brms::as_draws_df(joint_model), dimension = 6) |>
  dplyr::mutate(
    feature_a = sub("^responded|^score", "", response_a),
    feature_b = sub("^responded|^score", "", response_b),
    kind_a = ifelse(grepl("^responded", response_a), "gate", "accuracy"),
    kind_b = ifelse(grepl("^responded", response_b), "gate", "accuracy"),
    family = dplyr::case_when(
      feature_a == feature_b ~ "Selection",
      kind_a == "accuracy" & kind_b == "accuracy" ~ "Trade-off",
      kind_a == "gate" & kind_b == "gate" ~ "Reporting, across features",
      TRUE ~ "Cross terms"
    ),
    pair = paste(labels[response_a], "and", labels[response_b])
  )

correlations |>
  dplyr::filter(family == "Selection") |>
  dplyr::select(pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "Reporting against accuracy, within a feature")
```

Neither. It is **feature-specific**: substantial for orientation, absent for
word and colour.

The `moved` column is the prior check. These correlations are estimated
under `lkj(4)`, which is informative on a 6×6 matrix, so a posterior no
narrower than the marginal that prior implies has not learned anything
from the data. `TRUE` means it did. The [implementation
notes](implementation-notes.html#priors) give the marginal standard
deviations the comparison uses.

That is a third case the rule did not anticipate, and it is recorded as one
rather than filed under whichever branch it resembles. What follows from it
is narrower and more useful than either branch would have been:
**responders-only analysis is sound for word and colour, and conditions on
something real for orientation.** Pages that mix all three features, such
as the [composition](composition.html) page, inherit a weakened version of
the caveat rather than a blanket one.

Two things make the orientation result more credible rather than less.
It replicates: estimated from orientation alone it is 0.52, and with all
six responses in the model 0.56, so it is not an artifact of the smaller
specification. Both numbers in the table below are read from the fitted
models rather than transcribed. (The [analysis
strategy](analysis-strategy.html) page gives the orientation-only figure as
0.51: `posterior_correlations()` returns posterior **medians** and
`report_rope()` returns **means**, and on this posterior they differ in the
second digit. Same parameter, two summaries.) And orientation is the one feature whose gate is
*inferred* rather than observed: an untouched widget returns 90 degrees, so
genuine 90-degree responses are miscoded as non-responses. That miscoding
removes guaranteed-wrong trials from the accuracy of exactly the
participants with low measured response rates, which pushes the correlation
**negative**. Finding it positive despite that makes it conservative.

## 3. Does doing well on one feature cost another?

```{r tradeoff}
correlations |>
  dplyr::filter(family == "Trade-off") |>
  dplyr::select(pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "Accuracy against accuracy, across features")
```

No. Orientation and colour correlate **positively**, clearly so, and the
other two pairs are uninformative. Between people, shared ability dominates:
participants good at one non-verbal feature tend to be good at the other.

This settles a question the earlier analyses could not. Working from raw
proportions, orientation and colour appeared to trade off against each
other, and the [scoring](scoring.html) page shows why that reading was
wrong: closure induces negative correlation whatever the data says, and the
apparent asymmetry was six low-engagement participants. Here the parts are
not closed, so a negative correlation would have meant something. It is
positive.

**The trade-off the task imposes is a within-trial constraint, not a
between-person one.**

```{r tier1-figure, fig.alt="Posterior distributions of the six pre-specified correlations."}
correlations |>
  dplyr::filter(family %in% c("Selection", "Trade-off")) |>
  dplyr::mutate(pair = stats::reorder(pair, median)) |>
  ggplot(aes(x = median, y = pair, colour = family)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.4) +
  facet_grid(family ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_colour_manual(values = unname(palette.colors()[c(6, 4)]),
                      guide = "none") +
  labs(x = "Participant-level correlation", y = NULL,
       caption = "Medians with 95% credible intervals.")
```

## 4. What complete aphantasia changes

The model estimates a floor-group offset on all six responses, so the same
question can be asked of willingness and of accuracy.

```{r floor-effects}
fixed <- as.data.frame(brms::fixef(joint_model))
fixed$parameter <- rownames(fixed)

fixed |>
  dplyr::filter(grepl("complete_aphantfloor$", parameter)) |>
  dplyr::mutate(
    response = sub("_complete_aphantfloor$", "", parameter),
    kind = ifelse(grepl("^responded", response),
                  "Gate: willing to report", "Accuracy when reporting"),
    feature = c(word = "Word", angle = "Orientation",
                color = "Colour")[sub("^responded|^score", "", response)]
  ) |>
  dplyr::select(kind, feature, Estimate, Q2.5, Q97.5) |>
  knitr::kable(digits = 3, row.names = FALSE,
               caption = "Floor-group offsets, log-odds")
```

**On willingness to report, there is no result.** Not a null: no result. On
the probability scale the orientation gate's interval runs from about 0.84
to 0.99, which is consistent with a large difference and with none.
Non-response is rare, between 7% and 14% of trials, and the floor group is
20 people. This is the analysis that could in principle have shown
aphantasic participants declining to report differently, and it does not
have the precision to.

That is worth stating as a design lesson rather than a finding: to measure
abstention you need a task in which abstention is common.

```{r floor-figure, fig.alt="Reporting probability for orientation against imagery vividness, with the floor group shown separately.", fig.height = 6}
grid <- tibble::tibble(
  vviq = seq(16, 80, length.out = 200),
  complete_aphant = factor("above_floor", levels = c("above_floor", "floor")),
  parity_rate = stats::median(model_data$parity_rate),
  responded_word = TRUE, responded_angle = TRUE, responded_color = TRUE
)

above_floor <- brms::posterior_epred(
  joint_model, newdata = grid, resp = "respondedangle", re_formula = NA)

floor_row <- grid[1, ]
floor_row$vviq <- 16
floor_row$complete_aphant <- factor("floor",
                                    levels = c("above_floor", "floor"))
floor_draws <- as.vector(brms::posterior_epred(
  joint_model, newdata = floor_row, resp = "respondedangle", re_formula = NA))

observed <- model_data |>
  dplyr::summarise(y = mean(responded_angle), vviq = dplyr::first(vviq),
                   .by = id)

gate_effect <- brms::fixef(joint_model)["respondedangle_complete_aphantfloor", ]

main_panel <- plot_floor_group(
  observed = dplyr::transmute(dplyr::filter(observed, vviq > 16),
                              x = vviq, y = y),
  fitted = tibble::tibble(
    x = grid$vviq,
    estimate = apply(above_floor, 2, stats::median),
    lower = apply(above_floor, 2, stats::quantile, 0.025),
    upper = apply(above_floor, 2, stats::quantile, 0.975)
  ),
  floor_draws = floor_draws,
  floor_observed = dplyr::filter(observed, vviq == 16)$y,
  effect_label = sprintf("%.2f\n[%.2f, %.2f]", gate_effect[["Estimate"]],
                         gate_effect[["Q2.5"]], gate_effect[["Q97.5"]]),
  y_lab = "P(reporting the orientation)",
  caption = "Offset in log-odds. Points coloured by vividness.",
  base_size = 16, point_size = 2.2
)

plot_vviq_histogram(observed$vviq, base_size = 16) / main_panel +
  patchwork::plot_layout(heights = c(1, 4))
```

The histogram above the panel is the data that motivates the model on its
own: vividness is a spike at the floor plus an irregular remainder, not a
smooth continuum. In the panel, the half violin is the floor group's own
values, the cross is where the above-floor relationship predicts they
would be, and the arrow is the gap between them. Here there is barely a
gap, which is the visual form of the row in the table above.

**On accuracy the picture is the one the other pages found.** Word's offset
is positive and excludes zero, colour's is negative and touches it, and
orientation's is centred near zero. Word's accuracy row is excluded from
individual-differences interpretation on reliability grounds ([task
validity](task-validity.html)), and its **gate** row is not: whether
someone answers at all is not a low-reliability quantity.

## 5. The strongest thing in the matrix was not pre-specified

The nine remaining correlations were reported for completeness. One family
of them is hard to ignore.

```{r tier2}
correlations |>
  dplyr::filter(!family %in% c("Selection", "Trade-off")) |>
  # the gate-to-gate family first: it is the one the text singles out
  dplyr::mutate(family = factor(
    family, levels = c("Reporting, across features", "Cross terms"))) |>
  dplyr::arrange(family, -abs(median)) |>
  dplyr::select(family, pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "The nine remaining correlations")
```

The three gate-to-gate correlations are the largest in the whole model, all
with the posterior entirely on one side of zero. **Willingness to report is
a coherent participant-level trait**, far more consistent across features
than accuracy is, and the two connect only on orientation.

This was **not** in the pre-specified set, so it is exploratory and is
labelled that way. It is also the clearest evidence available that
abstention deserved to be modelled rather than filtered: a behaviour this
consistent within participants is not noise.

## 6. Checks

The simpler models, reported in full on the
[performance](performance.html) page, are the same data through a simpler
lens, so agreement is a robustness check and not independent evidence. They
also run on `r comparison_sample` participants against this model's
`r dplyr::n_distinct(model_data$id)`, so the comparison below is
approximate by construction. It is matched on specification: both carry
`parity_rate` on the accuracy arms and neither carries it on a gate.

```{r sanity}
# Loaded rather than transcribed: the comparison stays true if either
# model is refitted. These two were fitted on the engaged sample, not on
# this page's larger one, so the frame is rebuilt here to match.
performance_data <- dplyr::filter(model_data, id %in% engaged)

performance_model <- fit_brms_model(
  formula = performance_formula(),
  data    = performance_data,
  prior   = c(
    do.call(c, lapply(
      c("scoreword", "scoreangle", "scorecolor"),
      response_priors, terms = c("vviq", "complete_aphant", "parity_rate")
    )),
    brms::prior("lkj(2)", class = "cor")
  ),
  file    = system.file("models", "perf-c-prime.rds", package = pkg),
  file_refit = refit
)
orientation_only <- fit_brms_model(
  formula =
    brms::bf(responded_angle ~ vviq + complete_aphant + (1 | p | id),
             family = brms::bernoulli()) +
    brms::bf(score_angle | subset(responded_angle) ~
               vviq + complete_aphant + parity_rate + (1 | p | id),
             family = brms::Beta()) +
    brms::set_rescor(FALSE),
  data    = performance_data,
  prior   = c(
    response_priors("respondedangle", terms = c("vviq", "complete_aphant")),
    response_priors("scoreangle",
                    terms = c("vviq", "complete_aphant", "parity_rate")),
    brms::prior("lkj(2)", class = "cor")
  ),
  save_pars = brms::save_pars(group = FALSE),
  file    = system.file("models", "perf-joint-orientation.rds", package = pkg),
  file_refit = refit
)

offset_of <- function(model, response) {
  brms::fixef(model)[paste0(response, "_complete_aphantfloor"), "Estimate"]
}
correlation_of <- function(model, a, b) {
  pairs <- posterior_correlations(brms::as_draws_df(model), lkj = NULL)
  match <- pairs$response_a %in% c(a, b) & pairs$response_b %in% c(a, b)
  pairs$median[match]
}

tibble::tibble(
  quantity = c("Word accuracy floor offset",
               "Orientation accuracy floor offset",
               "Colour accuracy floor offset",
               "cor(orientation, colour accuracy)",
               "cor(orientation gate, accuracy)"),
  simpler_model = c(
    offset_of(performance_model, "scoreword"),
    offset_of(performance_model, "scoreangle"),
    offset_of(performance_model, "scorecolor"),
    correlation_of(performance_model, "scoreangle", "scorecolor"),
    correlation_of(orientation_only, "respondedangle", "scoreangle")
  ),
  joint_model = c(
    offset_of(joint_model, "scoreword"),
    offset_of(joint_model, "scoreangle"),
    offset_of(joint_model, "scorecolor"),
    correlation_of(joint_model, "scoreangle", "scorecolor"),
    correlation_of(joint_model, "respondedangle", "scoreangle")
  )
) |>
  dplyr::mutate(difference = joint_model - simpler_model) |>
  knitr::kable(digits = 3, caption = "Against the performance models")
```

Everything lands close except the orientation-colour accuracy correlation,
which drops from 0.586 to 0.456. That is the expected direction: with the
gates in the model, variance previously absorbed into accuracy is
attributed to willingness instead.

Fitting the three gates without the accuracy arms moves the gate
coefficients by almost nothing, so the accuracy arms are not pulling them.

```{r diagnostics}
tibble::tibble(
  `max R-hat` = max(brms::rhat(joint_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "divergent__")$Value),
  `min bulk ESS ratio` = min(brms::neff_ratio(joint_model), na.rm = TRUE),
  draws = brms::ndraws(joint_model)
) |>
  knitr::kable(digits = 4, caption = "Sampler diagnostics")
```

---

**Continuing through the Extended Online Report:** this page follows [performance](performance.html). To keep reading in order, continue to [beyond vividness](beyond-vividness.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
