---
title: "The joint model: propensity and accuracy together"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{The joint model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
library(patchwork)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)

pkg <- "aphantasiaWMStrats"
refit <- "never"
```

The [composition](composition.html) page reports this study's primary result,
and it computes it from **responded items only**. Whether that conditioning was
innocent cannot be checked from inside a responders-only analysis, because the
information needed to check it has been removed.

This model checks it, and answers two questions of its own that no other page
can reach: whether willingness to report is itself a stable individual trait,
and whether doing well on one feature costs another between people.

Six responses. Three **gates**, one per feature, for whether it was reported at
all. Three **accuracies**, for how well it was reported when it was. All six
share a participant-level correlation matrix, and those fifteen correlations are
the point.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

model_data <- v1 |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq)) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    score_color = squeeze_boundaries(score_color, n = sum(responded_color))
  )

joint_model <- fit_brms_model(
  formula = joint_formula(),
  data    = model_data,
  prior   = joint_priors(),
  file    = system.file("models", "joint-full.rds", package = pkg),
  file_refit = refit
)
```

## 1. The sample is larger here, deliberately {#joint-sample}

```{r sample}
engaged <- engaged_ids(v1)

# `model_data` already excludes participants with no vividness score, so
# this is "clears the thresholds AND has a VVIQ" -- the sample the
# composition and performance pages model.
comparison_sample <- dplyr::n_distinct(
  dplyr::filter(model_data, id %in% engaged)$id
)
```

```{r comparison-models}
# The two simpler models. Loaded here rather than in section 6 because
# both sections read from them, and a number quoted four sections before
# the object that produces it is a transcription waiting to go stale.
# Both were fitted on the engaged sample, not on this page's larger one,
# so the frame is rebuilt to match.
performance_data <- dplyr::filter(model_data, id %in% engaged)

performance_model <- fit_brms_model(
  formula = performance_formula(),
  data    = performance_data,
  prior   = c(
    do.call(c, lapply(
      c("scoreword", "scoreangle", "scorecolor"),
      response_priors, terms = c("vviq", "complete_aphant", "parity_rate")
    )),
    brms::prior("lkj(2)", class = "cor")
  ),
  file    = system.file("models", "perf-c-prime.rds", package = pkg),
  file_refit = refit
)

orientation_only <- fit_brms_model(
  formula =
    brms::bf(responded_angle ~ vviq + complete_aphant + (1 | p | id),
             family = brms::bernoulli()) +
    brms::bf(score_angle | subset(responded_angle) ~
               vviq + complete_aphant + parity_rate + (1 | p | id),
             family = brms::Beta()) +
    brms::set_rescor(FALSE),
  data    = performance_data,
  prior   = c(
    response_priors("respondedangle", terms = c("vviq", "complete_aphant")),
    response_priors("scoreangle",
                    terms = c("vviq", "complete_aphant", "parity_rate")),
    brms::prior("lkj(2)", class = "cor")
  ),
  save_pars = brms::save_pars(group = FALSE),
  file    = system.file("models", "perf-joint-orientation.rds", package = pkg),
  file_refit = refit
)

offset_of <- function(model, response) {
  brms::fixef(model)[paste0(response, "_complete_aphantfloor"), "Estimate"]
}
correlation_of <- function(model, a, b) {
  pairs <- posterior_correlations(brms::as_draws_df(model), lkj = NULL)
  match <- pairs$response_a %in% c(a, b) & pairs$response_b %in% c(a, b)
  pairs$median[match]
}

selection_replication <- c(
  orientation_alone = correlation_of(orientation_only,
                                     "respondedangle", "scoreangle"),
  six_responses     = correlation_of(joint_model,
                                     "respondedangle", "scoreangle")
)
```

`r length(engaged)` participants clear the engagement thresholds, and
`r comparison_sample` of those also have a vividness score: that is the sample
every other modelling page uses. This one uses all
`r dplyr::n_distinct(model_data$id)` v1 participants with a vividness score,
thresholds or no. [The analysis](analysis-strategy.html#strat-samples) sets out the
whole decision chain in one table.

The thresholds exist to stop imprecise participant *means* being treated as
measurements, and this model forms no participant means. More to the point, the
participants they exclude are the ones with the most to say about abstention:
one answered **zero** orientation items, which makes them invisible to every
responders-only analysis and highly informative here.

```{r specification}
joint_formula()
```

The `subset()` terms are what let one item carry six responses with independent
non-response, and what let a participant contribute to a gate whether or not
they ever answered that feature. `parity_rate` is on the accuracy arms only: it
is the dual-task load a participant actually incurred, and the [task
engagement](engagement.html) page shows it does not predict whether they
respond.

The fifteen correlations are the object of interest, in four families:

| Family | Question | Reported in |
|----|----|----|
| Response and accuracy, within a feature | Was conditioning on responding legitimate, and is the effect feature-specific or a general engagement trait? | §2 |
| Accuracy and accuracy, across features | Does doing well on one feature cost another, between people? | §3 |
| Response and response, across features | Do people who skip one feature skip others? | §5 |
| The six cross terms | Everything else | §5 |

Which of them are foregrounded was fixed in advance, in three tiers, so that
choosing among fifteen after seeing them is not possible. The six in the first
two families were pre-specified; the nine in §5 were not, and are labelled
exploratory there.

## 2. Was analysing responded items only legitimate? {#joint-selection}

Three correlations answer that, one per feature. The rule for reading them was
fixed before they existed: near zero everywhere would mean responders-only
analyses were fine; large everywhere would mean a general engagement trait and a
caveat on everything conditional.

```{r correlations}
labels <- c(respondedword = "Word gate", respondedangle = "Orientation gate",
            respondedcolor = "Colour gate", scoreword = "Word accuracy",
            scoreangle = "Orientation accuracy", scorecolor = "Colour accuracy")

correlations <- posterior_correlations(
  brms::as_draws_df(joint_model), dimension = 6) |>
  dplyr::mutate(
    feature_a = sub("^responded|^score", "", response_a),
    feature_b = sub("^responded|^score", "", response_b),
    kind_a = ifelse(grepl("^responded", response_a), "gate", "accuracy"),
    kind_b = ifelse(grepl("^responded", response_b), "gate", "accuracy"),
    family = dplyr::case_when(
      feature_a == feature_b ~ "Selection",
      kind_a == "accuracy" & kind_b == "accuracy" ~ "Trade-off",
      kind_a == "gate" & kind_b == "gate" ~ "Reporting, across features",
      TRUE ~ "Cross terms"
    ),
    pair = paste(labels[response_a], "and", labels[response_b])
  )

correlations |>
  dplyr::filter(family == "Selection") |>
  dplyr::select(pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "Reporting against accuracy, within a feature")
```

... It is neither. It is **feature-specific**: substantial for orientation,
absent for word and colour.

The `moved` column is the prior check. These correlations are estimated under
`lkj(4)`, which is informative on a 6×6 matrix, so a posterior no narrower than
the marginal that prior implies has not learned anything from the data. `TRUE`
means it did. The [implementation notes](implementation-notes.html#impl-priors) give
the marginal standard deviations the comparison uses.

That is a third case the rule did not anticipate. What follows from it is
narrower: **responders-only analysis is sound for word and colour, and
conditions on something real for orientation.** Pages that mix all three
features, such as the [composition](composition.html) page, inherit a weakened
version of the caveat and can legitimately be considered fine.

This orientation result above
(`r fmt(selection_replication[["six_responses"]], 2)`), with all six responses
in the model, replicates the one estimated from a orientation-only model
(`r fmt(selection_replication[["orientation_alone"]], 2)`), so it is not an
artifact of the smaller specification. Both figures are read from the fitted
models.

## 3. Does doing well on one feature cost another? {#joint-tradeoff}

```{r tradeoff}
correlations |>
  dplyr::filter(family == "Trade-off") |>
  dplyr::select(pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "Accuracy against accuracy, across features")
```

No. Orientation and colour correlate **positively**, clearly so, and the other
two pairs are uninformative. Between people, shared ability dominates:
participants good at one non-verbal feature tend to be good at the other.

```{r tier1-figure, fig.alt="Posterior distributions of the six pre-specified correlations."}
correlations |>
  dplyr::filter(family %in% c("Selection", "Trade-off")) |>
  dplyr::mutate(pair = stats::reorder(pair, median)) |>
  ggplot(aes(x = median, y = pair, colour = family)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_pointrange(aes(xmin = lower, xmax = upper), size = 0.4) +
  facet_grid(family ~ ., scales = "free_y", space = "free_y") +
  scale_x_continuous(limits = c(-1, 1)) +
  scale_colour_manual(values = unname(palette.colors()[c(6, 4)]),
                      guide = "none") +
  labs(x = "Participant-level correlation", y = NULL,
       caption = "Medians with 95% credible intervals.")
```

## 4. What complete aphantasia changes {#joint-floor}

The model estimates a floor-group offset on all six responses, so the same
question can be asked of willingness and of accuracy.

```{r floor-effects}
fixed <- as.data.frame(brms::fixef(joint_model))
fixed$parameter <- rownames(fixed)

fixed |>
  dplyr::filter(grepl("complete_aphantfloor$", parameter)) |>
  dplyr::mutate(
    response = sub("_complete_aphantfloor$", "", parameter),
    kind = ifelse(grepl("^responded", response),
                  "Gate: willing to report", "Accuracy when reporting"),
    feature = c(word = "Word", angle = "Orientation",
                color = "Colour")[sub("^responded|^score", "", response)]
  ) |>
  dplyr::select(kind, feature, Estimate, Q2.5, Q97.5) |>
  knitr::kable(digits = 3, row.names = FALSE,
               caption = "Floor-group offsets, log-odds")
```

**On willingness to report, there is no result** (not evidence for the null). On
the probability scale the orientation gate's interval runs from about 0.84 to
0.99, which is consistent with a large difference and with none. Non-response is
rare, between 7% and 14% of items, and the floor group is 20 people. This is the
analysis that could in principle have shown aphantasic participants declining to
report differently, and it does not have the precision to.

This is more a design lesson than a finding: to measure abstention you need a
task in which abstention is common.

**On accuracy, the picture is the one the other analyses (compositional and
performance-based) found.** Word's offset is positive and excludes zero,
colour's is negative and touches it, and orientation's is centred near zero.
Word's accuracy row is excluded from individual-differences interpretation on
reliability grounds ([task validity](task-validity.html)), and its **gate** row
is not: whether someone answers at all is a reliable quantity.

## 5. The largest correlations were not pre-specified {#joint-exploratory}

The nine remaining correlations were reported for completeness. One family of
them is hard to ignore.

```{r tier2}
correlations |>
  dplyr::filter(!family %in% c("Selection", "Trade-off")) |>
  # the gate-to-gate family first: it is the one the text singles out
  dplyr::mutate(family = factor(
    family, levels = c("Reporting, across features", "Cross terms"))) |>
  dplyr::arrange(family, -abs(median)) |>
  dplyr::select(family, pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "The nine remaining correlations")
```

The three gate-to-gate correlations are the largest in the whole model, all with
the posterior entirely on one side of zero. **Willingness to report is a
coherent participant-level trait**, far more consistent across features than
accuracy is, and the two connect only on orientation.

This was **not** in the pre-specified set, so it is exploratory. It is also the
clearest evidence available that abstention deserved to be modelled rather than
filtered: a behaviour this consistent within participants is not noise.

## 6. Checks {#joint-checks}

The simpler models, reported in full on the [performance](performance.html)
page, are the same data through a simpler lens, so agreement with these models
can be seen as robustness checks. They run on `r comparison_sample` participants
against this model's `r dplyr::n_distinct(model_data$id)`, so the comparison
below is approximate by construction. It is matched on specification: both carry
`parity_rate` on the accuracy arms and neither carries it on a gate.

```{r sanity}
comparison <- tibble::tibble(
  quantity = c("Word accuracy floor offset",
               "Orientation accuracy floor offset",
               "Colour accuracy floor offset",
               "cor(orientation, colour accuracy)",
               "cor(orientation gate, accuracy)"),
  simpler_model = c(
    offset_of(performance_model, "scoreword"),
    offset_of(performance_model, "scoreangle"),
    offset_of(performance_model, "scorecolor"),
    correlation_of(performance_model, "scoreangle", "scorecolor"),
    correlation_of(orientation_only, "respondedangle", "scoreangle")
  ),
  joint_model = c(
    offset_of(joint_model, "scoreword"),
    offset_of(joint_model, "scoreangle"),
    offset_of(joint_model, "scorecolor"),
    correlation_of(joint_model, "scoreangle", "scorecolor"),
    correlation_of(joint_model, "respondedangle", "scoreangle")
  )
) |>
  dplyr::mutate(difference = joint_model - simpler_model)

knitr::kable(comparison, digits = 3,
             caption = "Against the performance models")
```

Everything lands close except the orientation-colour accuracy correlation, which
moves from
`r fmt(comparison$simpler_model[comparison$quantity == "cor(orientation, colour accuracy)"])`
to
`r fmt(comparison$joint_model[comparison$quantity == "cor(orientation, colour accuracy)"])`.
That is the expected direction: with the gates in the model, variance previously
absorbed into accuracy is attributed to willingness instead.

The gates can also be fitted on their own, without the accuracy arms, to check
that the accuracy side is not pulling them:

```{r gates-only}
gates_only <- fit_brms_model(
  formula = joint_formula(accuracy_features = character(0)),
  data    = model_data,
  prior   = joint_priors(accuracy_features = character(0), lkj = 4),
  save_pars = brms::save_pars(group = FALSE),
  adapt_delta = 0.99,
  max_treedepth = 15,
  file    = system.file("models", "joint-gates.rds", package = pkg),
  file_refit = refit
)

gate_terms <- c("vviq", "complete_aphantfloor")
purrr::map(c("respondedword", "respondedangle", "respondedcolor"), \(r) {
  rows <- paste0(r, "_", gate_terms)
  tibble::tibble(
    gate = r,
    term = gate_terms,
    gates_only = brms::fixef(gates_only)[rows, "Estimate"],
    full_model = brms::fixef(joint_model)[rows, "Estimate"]
  )
}) |>
  purrr::list_rbind() |>
  dplyr::mutate(difference = full_model - gates_only) |>
  knitr::kable(digits = 3, caption = "Gate coefficients, with and without the accuracy arms")
```

The convergence and posterior predictive checks appear alongside the rest of the
models on the [model diagnostics](model-diagnostics.html#diag-convergence) page. They
are repeated here for completeness:

```{r diagnostics}
tibble::tibble(
  `max R-hat` = max(brms::rhat(joint_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "divergent__")$Value),
  `min bulk ESS ratio` = min(brms::neff_ratio(joint_model), na.rm = TRUE),
  draws = brms::ndraws(joint_model)
) |>
  knitr::kable(digits = 4, caption = "Sampler diagnostics")
```

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows
[performance](performance.html). To keep reading in order, continue to [beyond
vividness](beyond-vividness.html) next. Or see [model
diagnostics](model-diagnostics.html) and [implementation
notes](implementation-notes.html) for the technical detail behind these models.

--------------------------------------------------------------------------------

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
