---
title: "What was tried and withdrawn"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{What was tried and withdrawn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

Several findings on this site were reported at some point and are no longer
claimed. This page says what they were and what removed them, because an
Extended Online Report that only shows what survived is a longer methods
section rather than a more honest one.

Three of them were caught by the same move: **check whether a pooled result
holds within each stratum before reporting it.** The other two are
different in kind, and are included because a page of near-misses that all
share one cause would misrepresent how the errors actually arrived.

## 1. A two-way trade-off between colour and orientation

**Claimed:** the task was designed to force a three-way trade-off but
achieves a two-way one, because word is easy enough that participants can
max it and then divide effort between colour and orientation. Supported by
correlations between features after removing each participant's overall
level.

**Withdrawn** when the same correlations were computed on participants who
clear the engagement thresholds:

```{r tradeoff}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

all_parts <- compose_features(v1, id)
all_parts <- all_parts[stats::complete.cases(all_parts), ]
engaged_parts <- dplyr::filter(all_parts, id %in% engaged_ids(v1))

feature_labels <- c(word = "Word", angle = "Orientation", color = "Colour")

knitr::kable(
  centred_correlations(all_parts, labels = feature_labels), digits = 3,
  caption = paste0("All participants with three parts (n = ",
                   nrow(all_parts), ")"))

knitr::kable(
  centred_correlations(engaged_parts, labels = feature_labels), digits = 3,
  caption = paste0("Clearing the engagement thresholds (n = ",
                   nrow(engaged_parts), ")"))
```

Residualising three parts on their own mean induces a correlation near
-0.5 by construction, so that is the reference and not zero. On the first
table orientation and colour appear to trade off more strongly than either
does against word. On the second all three pairs sit close to the closure
baseline.

**The difference between the two tables is six participants**, all of whom
answered very few orientation items. The apparent trade-off described
them, not the task.

The [joint model](joint-model.html) later settled the question from the
other direction: between people, orientation and colour accuracy correlate
**positively**. There is no between-person trade-off for closure to have
been hiding.

## 2. Reporting propensity is related to imagery vividness

**Claimed:** participants with lower imagery vividness decline to report
features more often, with a pooled correlation around -0.30.

**Withdrawn** when stratified by task version:

```{r propensity}
by_participant <- get_data() |>
  dplyr::filter(grepl("^expe_block", expe_phase)) |>
  dplyr::summarise(
    non_response = 1 - mean(responded_angle),
    vviq = dplyr::first(vviq_total_score),
    .by = c(id, version)
  ) |>
  dplyr::filter(!is.na(vviq))

dplyr::bind_rows(
  dplyr::mutate(
    correlation_test(by_participant$non_response, by_participant$vviq),
    sample = "Pooled"),
  by_participant |>
    dplyr::group_split(version) |>
    purrr::map(\(data) dplyr::mutate(
      correlation_test(data$non_response, data$vviq),
      sample = dplyr::first(data$version))) |>
    purrr::list_rbind()
) |>
  dplyr::select(sample, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "Orientation non-response against vividness")
```

v1 is `r round(100 * mean(by_participant$version == "v1"))`% of the
participants with a vividness score and shows nothing. The pooled estimate
is driven by v3, which is both heavily aphantasic — 17 of its 21
participants — and by far the highest in non-response, so pooling
manufactured an association that neither version contains.

This one did more than get withdrawn: it is the concrete reason the
[version scope](version-scope.html) decision is not bureaucratic.

## 3. Parity accuracy predicts imagery vividness

**Claimed**, on an earlier version of the [task
engagement](engagement.html) page: parity engagement is not independent of
imagery, and this is what stops parity accuracy being usable as a nuisance
covariate.

**Withdrawn** for two reasons at once. The correlation was rho = -0.154 at
p = .157, reported as a finding when the interval comfortably contains
zero. And the variable was not what its name said: `parity_*_acc` scores an
unanswered probe as 0, and 92% of its zeros in v1 are unanswered rather
than wrong, so a participant mean of it correlates 0.99 with the proportion
of probes answered and 0.22 with accuracy among those answered.

Corrected, parity engagement does not differ by imagery group at all
(p = .93). The fix added `responded_parity_1` and `responded_parity_2` to
the data, and the engagement page now shows the conflation rather than
inheriting it.

## 4. Word recall as an individual-differences measure

Not withdrawn so much as **demoted**, and it constrains every page that
reports a word coefficient.

Word's split-half reliability is 0.45, against 0.82 and 0.83 for the other
two features. The metric is not at fault: single-word recall at this
exposure is simply too easy to discriminate between people. The
[performance](performance.html) page shows what that looks like from the
model's side, with every participant's estimate collapsing to nearly the
same value.

Word's coefficients are still reported, because a group contrast is a
different estimand and measurement error widens intervals rather than
biasing them. But no individual-differences claim about word is made
anywhere, whatever its coefficients say.

## 5. Two claims about this project's own history

**A published clustering study was described here as circular in design.**
It was not: it included a genuine held-out validation attempt that returned
a null. What went wrong was the framing, which oversold a descriptive
result under co-author pressure. Recording it accurately matters more than
recording it harshly.

**A claim about a mislabelled sample was itself wrong.** An earlier version
of the parity analysis asserted that the engagement page computed a pooled
correlation while labelling it v1. It did not: the page filtered to v1
correctly, and the claim came from reading one function call without
reading the object it consumed. The planning documents keep the retraction
rather than deleting it.

## What these have in common

Four of the five were produced by the same thing: **a statistic computed on
one sample and interpreted as though it applied to another.** Pooled across
versions, pooled across engagement levels, or pooled across a distinction
the variable's name concealed.

The habit that catches it is cheap. Before reporting an association,
compute it again within each stratum you have, and see whether it survives.

---

**Continuing through the Extended Online Report:** this page follows [implementation notes](implementation-notes.html). That is the end of the ordered sequence; the [codebook](codebook.html) documents every column in the data. Or return to [the analysis](analysis-strategy.html) for what the models were built to do.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
