---
title: "What was tried and withdrawn"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{What was tried and withdrawn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
```

Several findings on this site were reported at some point and are no longer
claimed. This page says what they were and what removed them.

Three of them were caught by the same move: **check whether a pooled result
holds within each version before reporting it.**

## 1. A two-way trade-off between colour and orientation {#less-tradeoff}

**Claimed:** the task was designed to force a three-way trade-off but achieves a
two-way one, because word is easy enough that participants can max it and then
divide effort between colour and orientation. Supported by correlations between
features after removing each participant's overall level.

**Withdrawn** when the same correlations were computed on participants who clear
the engagement thresholds:

```{r tradeoff}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

all_parts <- compose_features(v1, id)
all_parts <- all_parts[stats::complete.cases(all_parts), ]
engaged_parts <- dplyr::filter(all_parts, id %in% engaged_ids(v1))

feature_labels <- c(word = "Word", angle = "Orientation", color = "Colour")

knitr::kable(
  centred_correlations(all_parts, labels = feature_labels), digits = 3,
  caption = paste0("All participants with three parts (n = ",
                   nrow(all_parts), ")"))

knitr::kable(
  centred_correlations(engaged_parts, labels = feature_labels), digits = 3,
  caption = paste0("Clearing the engagement thresholds (n = ",
                   nrow(engaged_parts), ")"))
```

Residualising three parts on their own mean induces a correlation near -0.5 by
construction, so that is the reference and not zero. On the first table
orientation and colour appear to trade off more strongly than either does
against word. On the second all three pairs sit close to the closure baseline.

**The difference between the two tables is six participants**, all of whom
answered very few orientation items. The apparent trade-off described them, not
the task.

The [joint model](joint-model.html) later settled the question from the other
direction: between people, orientation and colour accuracy correlate
**positively**. There is no between-person trade-off for closure to have been
hiding.

## 2. Reporting propensity is related to imagery vividness {#less-propensity}

**Claimed:** participants with lower imagery vividness decline to report
features more often, with a pooled correlation around -0.30.

**Withdrawn** when stratified by task version:

```{r propensity}
by_participant <- get_data() |>
  dplyr::filter(grepl("^expe_block", expe_phase)) |>
  dplyr::summarise(
    non_response = 1 - mean(responded_angle),
    vviq = dplyr::first(vviq_total_score),
    .by = c(id, version)
  ) |>
  dplyr::filter(!is.na(vviq))

dplyr::bind_rows(
  dplyr::mutate(
    correlation_test(by_participant$non_response, by_participant$vviq),
    sample = "Pooled"),
  by_participant |>
    dplyr::group_split(version) |>
    purrr::map(\(data) dplyr::mutate(
      correlation_test(data$non_response, data$vviq),
      sample = dplyr::first(data$version))) |>
    purrr::list_rbind()
) |>
  dplyr::select(sample, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "Orientation non-response against vividness")
```

v1 is `r round(100 * mean(by_participant$version == "v1"))`% of the participants
with a vividness score and shows nothing. The pooled estimate is driven by v3,
which is both heavily aphantasic — 17 of its 21 participants — and by far the
highest in non-response, so pooling manufactured an association that neither
version contains.

## 3. Parity accuracy predicts imagery vividness {#less-parity}

**Claimed**, on an earlier version of the [task engagement](engagement.html)
page: parity engagement is not independent of imagery, and this is what stops
parity accuracy being usable as a nuisance covariate.

**Withdrawn** for two reasons at once. The correlation was rho = -0.154 at p =
.157, reported as a finding when the interval comfortably contains zero. And the
variable was not what its name said: `parity_*_acc` scores an unanswered probe
as 0, and 92% of its zeros in v1 are unanswered rather than wrong, so a
participant mean of it correlates 0.99 with the proportion of probes answered
and 0.22 with accuracy among those answered.

Corrected, parity engagement does not differ by imagery group at all (p = .93).
The fix added `responded_parity_1` and `responded_parity_2` to the data, and the
engagement page now shows the conflation rather than inheriting it.

## 4. Word recall as an individual-differences measure {#less-word}

Not withdrawn so much as **demoted**, and it constrains every page that reports
a word coefficient.

Word's split-half reliability is far below the other two features', which the
[task validity](task-validity.html) page computes. The metric is not at fault:
single-word recall at this exposure is simply too easy to discriminate between
people. The [performance](performance.html) page shows what that looks like from
the model's side, with every participant's estimate collapsing to nearly the
same value.

Word's coefficients are still reported, because a group contrast is a different
estimand and measurement error widens intervals rather than biasing them. But no
strong individual-differences claim about word should be made, whatever its
coefficients say.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows
[implementation notes](implementation-notes.html). That is the end of the
ordered sequence; the [codebook](codebook.html) documents every column in the
data. Or return to [the analysis](analysis-strategy.html) for what the models
were built to do.

--------------------------------------------------------------------------------

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
