---
title: "Model diagnostics"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 4.5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
pkg <- "aphantasiaWMStrats"
refit <- "never"
```

The modelling pages report estimates. This page asks whether the models
producing them sampled properly and whether their response families
describe the data. Both questions are usually answered in a sentence; here
they get the space to be checked, because three of the family choices are
not obvious.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

model_data <- v1 |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq)) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    score_color = squeeze_boundaries(score_color, n = sum(responded_color))
  )

joint_model <- fit_brms_model(
  formula = joint_formula(),
  data    = model_data,
  prior   = joint_priors(),
  # Not the wrapper's defaults of 0.95 and 10: this model was fitted with
  # these from the start, for the reason section 1 gives.
  adapt_delta = 0.99,
  max_treedepth = 15,
  file    = system.file("models", "joint-full.rds", package = pkg),
  file_refit = refit
)
```

## 1. Convergence {#convergence}

Four chains, 3000 iterations each with the first 1000 discarded as warmup,
giving 2000 kept per chain and 8000 draws in total. `fit_brms_model()`'s
`iterations` argument is post-warmup draws per chain rather than a total,
which is why the call reads `iterations = 2000, warmup = 1000`.

Those draws cover every parameter in the fitted object, which is the count
the table below reports. Most of them are the per-participant deviations;
the model's *structural* size is the 54 parameters section 4 refers to —
the population-level coefficients, the standard deviations and the fifteen
correlations.

```{r convergence}
tibble::tibble(
  parameters = length(brms::rhat(joint_model)),
  `max R-hat` = max(brms::rhat(joint_model), na.rm = TRUE),
  `R-hat above 1.01` = sum(brms::rhat(joint_model) > 1.01, na.rm = TRUE),
  `min bulk ESS ratio` = min(brms::neff_ratio(joint_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "divergent__")$Value),
  `max treedepth hits` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "treedepth__")$Value >= 15)
) |>
  knitr::kable(digits = 4, caption = "The joint model")
```

Nothing to report, which is the point of reporting it. R-hat below 1.004
everywhere, no divergent transitions, no saturated treedepths, and the
worst-sampled parameter still has an effective sample size ratio above
0.12.

That was not guaranteed. A six-response model with fifteen correlated
random intercepts estimated from 86 participants is the kind of model that
diverges, which is why it was fitted with `adapt_delta = 0.99` and
`max_treedepth = 15` from the start.

```{r rhat-spread, fig.alt="Distribution of R-hat values across all parameters."}
tibble::tibble(rhat = brms::rhat(joint_model)) |>
  dplyr::filter(!is.na(rhat)) |>
  ggplot(aes(x = rhat)) +
  geom_histogram(bins = 40, fill = "grey55") +
  geom_vline(xintercept = 1.01, linetype = "dashed", colour = "#C44E52") +
  labs(x = "R-hat", y = "Parameters",
       caption = "Dashed line marks the conventional 1.01 threshold.")
```

## 2. Do the families fit? {#ppc}

The three recall features get three different response families, and the
[scoring](scoring.html) page argues for each from what the boundary of that
feature's scale means. An argument is not a check.

A posterior predictive check draws replicate datasets from the fitted model
and compares them to the real one. If the family is wrong, the replicates
look wrong in a visible way.

```{r ppc-orientation, fig.alt="Posterior predictive check for orientation accuracy."}
brms::pp_check(joint_model, resp = "scoreangle", ndraws = 60) +
  labs(title = "Orientation: plain Beta",
       caption = "Dark line is the observed density; light lines are replicates.") +
  theme_pdf(base_size = 16)
```

Orientation has no boundary mass at all once non-responders are removed,
so a plain Beta is unremarkable and the replicates track the observed
density closely.

```{r ppc-word, fig.alt="Posterior predictive check for word accuracy."}
brms::pp_check(joint_model, resp = "scoreword", ndraws = 60) +
  labs(title = "Word: zero-one-inflated Beta") +
  theme_pdf(base_size = 16)
```

Word is the interesting one. About 91% of responded word trials score
exactly 1 and 3% exactly 0, because a recalled word either matches the
target or does not. Those are genuine point masses, not rounding, and the
inflation components are what let the model put mass exactly there. A plain
Beta cannot: it has zero density at both boundaries.

```{r ppc-colour, fig.alt="Posterior predictive check for colour accuracy."}
brms::pp_check(joint_model, resp = "scorecolor", ndraws = 60) +
  labs(title = "Colour: Beta after a Smithson-Verkuilen squeeze") +
  theme_pdf(base_size = 16)
```

Colour is the case where the same reasoning gives the opposite answer. It
has a trace of mass at 1, but colour score is cosine similarity on a
continuous wheel, so an exact 1 means an error of exactly zero degrees:
pixel resolution, not behaviour. Modelling that with an inflation component
would estimate a process parameter for a rounding artifact on a couple of
dozen trials. The squeeze moves every value by about one ten-thousandth
instead, and `squeeze_boundaries()` documents the trade.

```{r ppc-gates, fig.alt="Posterior predictive check for the response gates."}
patchwork::wrap_plots(
  purrr::map(
    c("respondedword", "respondedangle", "respondedcolor"),
    \(response) brms::pp_check(joint_model, resp = response, ndraws = 40) +
      labs(title = response) + theme_pdf(base_size = 14)
  ),
  ncol = 3, guides = "collect"
)
```

The three gates are Bernoulli, where the check reduces to whether the model
reproduces the proportion of trials answered. It does.

## 3. The compositional model {#composition-diagnostics}

The [composition](composition.html) page fits both log-ratio coordinates
together with the residual correlation estimated, which is what makes the
omnibus test invariant to the choice of partition.

```{r composition-model}
composition_data <- compose_features(
  dplyr::filter(v1, id %in% engaged_ids(v1)), id
)
composition_data <- composition_data[stats::complete.cases(composition_data), ]
composition_data <- composition_data |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq))
composition_data <- dplyr::bind_cols(
  composition_data,
  ilr_coords(dplyr::select(composition_data, tidyselect::starts_with("part_")))
) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    )
  )

composition_model <- fit_brms_model(
  formula = composition_formula("vviq + complete_aphant + parity_rate"),
  data    = composition_data,
  prior   = composition_priors(),
  file    = system.file("models", "comp-floor.rds", package = pkg),
  file_refit = refit
)

tibble::tibble(
  `max R-hat` = max(brms::rhat(composition_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(composition_model),
           Parameter == "divergent__")$Value),
  `min bulk ESS ratio` = min(brms::neff_ratio(composition_model), na.rm = TRUE)
) |>
  knitr::kable(digits = 4, caption = "The compositional model")
```

```{r ppc-composition, fig.alt="Posterior predictive checks for both ILR coordinates."}
patchwork::wrap_plots(
  brms::pp_check(composition_model, resp = "ilr1", ndraws = 60) +
    labs(title = "ilr1: words vs non-verbal") + theme_pdf(base_size = 14),
  brms::pp_check(composition_model, resp = "ilr2", ndraws = 60) +
    labs(title = "ilr2: colour vs orientation") + theme_pdf(base_size = 14),
  ncol = 2, guides = "collect"
)
```

Gaussian is appropriate here for a reason worth stating: log-ratio
coordinates are unbounded by construction. That is the whole point of
leaving the simplex, and it is why the [composition](composition.html)
page can use an ordinary linear model where the accuracy pages cannot.

## 4. What is not checked here

**Prior sensitivity.** The joint model's priors were chosen per
coefficient and are documented on the [analysis
strategy](analysis-strategy.html) page, but no alternative-prior refit was
run. For the correlations there is a partial substitute: each posterior is
compared against the marginal implied by the `lkj(4)` prior, and a
posterior no narrower than its prior is flagged rather than reported.

**Cross-validation of the joint model.** Model comparison was run on
functional form for one feature ([performance](performance.html) §4) and
on the compositional coordinates ([composition](composition.html) §5 and
§8), not on the joint model as a whole. With 54 parameters and 86
participants, a LOO comparison against alternatives would mostly report its
own uncertainty.

Both are limitations of scope rather than oversights, and both are the kind
of thing a larger sample would make worth doing.

---

**Continuing through the Extended Online Report:** this page follows [beyond vividness](beyond-vividness.html). To keep reading in order, continue to [implementation notes](implementation-notes.html) next. Or return to [the analysis](analysis-strategy.html) for what the models were built to do.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
