---
title: "Model diagnostics"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
pkg <- "aphantasiaWMStrats"
refit <- "never"
```

The modelling pages report estimates. This page asks whether the models
producing them sampled properly and whether their response families describe the
data. Both questions get the space to be checked, because three of the family
choices are not obvious. The most complex model in the project, the [joint model](joint-model.html), is checked first.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

model_data <- v1 |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq)) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    score_color = squeeze_boundaries(score_color, n = sum(responded_color))
  )

joint_model <- fit_brms_model(
  formula = joint_formula(),
  data    = model_data,
  prior   = joint_priors(),
  # Not the wrapper's defaults of 0.95 and 10: this model was fitted with
  # these from the start, for the reason section 1 gives.
  adapt_delta = 0.99,
  max_treedepth = 15,
  file    = system.file("models", "joint-full.rds", package = pkg),
  file_refit = refit
)
```

## 1. Convergence {#diag-convergence}

The model was fitted with four chains, 3000 iterations each with the first 1000 discarded as warmup,
giving 2000 kept per chain and 8000 draws in total. `fit_brms_model()`'s
`iterations` argument is post-warmup draws per chain rather than a total, which
is why the call reads `iterations = 2000, warmup = 1000`.

Those draws cover every parameter in the fitted object, which is the count the
table below reports. Most of them are the per-participant deviations; the joint
model's *structural* size is the 54 main parameters: the population-level
coefficients, the standard deviations and the fifteen correlations.

```{r convergence}
tibble::tibble(
  parameters = length(brms::rhat(joint_model)),
  `max R-hat` = max(brms::rhat(joint_model), na.rm = TRUE),
  `R-hat above 1.01` = sum(brms::rhat(joint_model) > 1.01, na.rm = TRUE),
  `min bulk ESS ratio` = min(brms::neff_ratio(joint_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "divergent__")$Value),
  `max treedepth hits` = sum(
    subset(brms::nuts_params(joint_model),
           Parameter == "treedepth__")$Value >= 15)
) |>
  knitr::kable(digits = 4, caption = "The joint model")
```

R-hat below 1.004 everywhere, no divergent transitions, no saturated treedepths,
and the worst-sampled parameter still has an effective sample size ratio above
0.12.

That was not guaranteed. A six-response model with fifteen correlated random
intercepts estimated from 86 participants is the kind of model that diverges,
which is why it was fitted with `adapt_delta = 0.99` and `max_treedepth = 15`
from the start.

```{r rhat-spread, fig.alt="Distribution of R-hat values across all parameters."}
tibble::tibble(rhat = brms::rhat(joint_model)) |>
  dplyr::filter(!is.na(rhat)) |>
  ggplot(aes(x = rhat)) +
  geom_histogram(bins = 40, fill = "grey55") +
  geom_vline(xintercept = 1.01, linetype = "dashed", colour = "#C44E52") +
  labs(x = "R-hat", y = "Parameters",
       caption = "Dashed line marks the conventional 1.01 threshold.")
```

## 2. Do the families fit? {#diag-ppc}

The three recall features get three different response families, and the
[scoring](scoring.html) page argues for each from what the boundary of that
feature's scale means.

A posterior predictive check draws replicate datasets from the fitted model and
compares them to the real one. If the chosen family is wrong, the replicates look wrong
in a visible way.

```{r ppc-orientation, fig.alt="Posterior predictive check for orientation accuracy."}
brms::pp_check(joint_model, resp = "scoreangle", ndraws = 60) +
  labs(title = "Orientation: plain Beta",
       caption = "Dark line is the observed density; light lines are replicates.") +
  theme_pdf(base_size = 16)
```

Orientation has no boundary mass at all once non-responders are removed, so a
plain Beta is unremarkable and the replicates track the observed density
closely.

```{r ppc-word, fig.alt="Posterior predictive check for word accuracy."}
brms::pp_check(joint_model, resp = "scoreword", ndraws = 60) +
  labs(title = "Word: zero-one-inflated Beta") +
  theme_pdf(base_size = 16)
```

Word is the interesting one. About 91% of responded word items score exactly 1
and 3% exactly 0, because a recalled word either matches the target or does not.
Those are genuine point masses, not rounding, and the inflation components are
what let the model put mass exactly there. A plain Beta cannot: it has zero
density at both boundaries.

```{r ppc-colour, fig.alt="Posterior predictive check for colour accuracy."}
brms::pp_check(joint_model, resp = "scorecolor", ndraws = 60) +
  labs(title = "Colour: Beta after a Smithson-Verkuilen squeeze") +
  theme_pdf(base_size = 16)
```

Colour is the case where the same reasoning gives the opposite answer. It has a
trace of mass at 1, but colour score is cosine similarity on a continuous wheel,
so an exact 1 means an error of exactly zero degrees: pixel resolution, not
behaviour. Modelling that with an inflation component would estimate a process
parameter for a rounding artifact on a couple of dozen items. The squeeze moves
every value by about one ten-thousandth instead, and `squeeze_boundaries()`
documents the trade.

```{r ppc-gates, fig.alt="Posterior predictive check for the response gates."}
patchwork::wrap_plots(
  purrr::map(
    c("respondedword", "respondedangle", "respondedcolor"),
    \(response) brms::pp_check(joint_model, resp = response, ndraws = 40) +
      labs(title = response) + theme_pdf(base_size = 14)
  ),
  ncol = 3, guides = "collect"
)
```

The three gates are Bernoulli, where the check reduces to whether the model
reproduces the proportion of items answered. It does.

## 3. The compositional model {#diag-composition}

The [composition](composition.html) page fits both log-ratio coordinates
together with the residual correlation estimated, which is what makes the
omnibus test invariant to the choice of partition.

```{r composition-model}
composition_data <- compose_features(
  dplyr::filter(v1, id %in% engaged_ids(v1)), id
)
composition_data <- composition_data[stats::complete.cases(composition_data), ]
composition_data <- composition_data |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq))
composition_data <- dplyr::bind_cols(
  composition_data,
  ilr_coords(dplyr::select(composition_data, tidyselect::starts_with("part_")))
) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    )
  )

composition_model <- fit_brms_model(
  formula = composition_formula("vviq + complete_aphant + parity_rate"),
  data    = composition_data,
  prior   = composition_priors(),
  file    = system.file("models", "comp-floor.rds", package = pkg),
  file_refit = refit
)

tibble::tibble(
  `max R-hat` = max(brms::rhat(composition_model), na.rm = TRUE),
  `divergent transitions` = sum(
    subset(brms::nuts_params(composition_model),
           Parameter == "divergent__")$Value),
  `min bulk ESS ratio` = min(brms::neff_ratio(composition_model), na.rm = TRUE)
) |>
  knitr::kable(digits = 4, caption = "The compositional model")
```

```{r ppc-composition, fig.alt="Posterior predictive checks for both ILR coordinates."}
patchwork::wrap_plots(
  brms::pp_check(composition_model, resp = "ilr1", ndraws = 60) +
    labs(title = "ilr1: words vs non-verbal") + theme_pdf(base_size = 14),
  brms::pp_check(composition_model, resp = "ilr2", ndraws = 60) +
    labs(title = "ilr2: colour vs orientation") + theme_pdf(base_size = 14),
  ncol = 2, guides = "collect"
)
```

Gaussian is appropriate here: log-ratio coordinates
are unbounded by construction. That is the whole point of leaving the simplex,
and it is why the [composition](composition.html) page can use an ordinary
linear model where the accuracy pages cannot.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows [beyond
vividness](beyond-vividness.html). To keep reading in order, continue to
[implementation notes](implementation-notes.html) next. Or return to [the
analysis](analysis-strategy.html) for what the models were built to do.

--------------------------------------------------------------------------------

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
