---
title: "The participants: who took part"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{The participants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 4.5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
```

Who these participants are, and one thing about how they were recruited
that constrains what any analysis here can claim.

This page describes **all `r dplyr::n_distinct(all_data$id)` people who
took part across the three task versions**. The behavioural analyses use
v1 alone and the questionnaire analyses pool all three; section 3 is where
that split comes from.

One caveat about counting, because two totals circulate on this site and
both are correct. **One participant completed the task twice, once in v1
and once in v3**, so there are `r dplyr::n_distinct(all_data$id)` people
and `r nrow(dplyr::distinct(all_data, id, version))` person-sessions.
Anything about a *person* — age, gender, education, imagery group — is
counted per person below. Anything about a *version* is counted per
session, because that participant did two different versions of the task
and each one is a separate observation of the design.

```{r data}
# One row per person, for the demographic and imagery-group tables. The
# repeat participant's v1 session is the one kept.
participants <- all_data |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::mutate(
    imagery_group = factor(
      vviq_group_4,
      levels = c("aphantasia", "hypophantasia", "typical", "hyperphantasia")
    ),
    # Free-text gender, harmonised. Seven raw values across two languages
    # describe three categories; the raw column is kept in `all_data`.
    gender = dplyr::case_when(
      gender == "f" ~ "Women",
      gender == "m" ~ "Men",
      TRUE ~ "Non-binary or self-described"
    )
  )

# One row per person-session, for anything counted by version.
sessions <- dplyr::distinct(all_data, id, version, .keep_all = TRUE)

extra <- dplyr::bind_rows(participants$extra_demographics)
```

## 1. Overall

```{r overall}
# Named `n_participants`, not `participants`: inside tibble() later
# expressions see earlier columns, so a column called `participants` would
# mask the data frame of the same name and every line below it would be
# operating on a count.
tibble::tibble(
  n_participants = nrow(participants),
  median_age = stats::median(participants$age),
  age_range = paste(range(participants$age), collapse = " to "),
  french_speaking = sum(participants$language == "fr"),
  in_France = sum(extra$country == "fra", na.rm = TRUE)
) |>
  knitr::kable(caption = "The pooled sample, across all three task versions")
```

```{r gender}
participants |>
  dplyr::count(gender) |>
  dplyr::arrange(dplyr::desc(n)) |>
  knitr::kable(caption = "Gender")
```

A predominantly French, predominantly women sample, skewed toward higher
education: `r sum(grepl("master|doctorat", extra$education))` of
`r nrow(extra)` hold a master's degree or a doctorate. That matters mainly
for generalisation, and it is the usual profile for online volunteer
studies of this kind.

```{r education}
extra |>
  dplyr::filter(!is.na(education)) |>
  dplyr::mutate(
    level = factor(sub("^isced_[0-9]+_", "", education))
  ) |>
  dplyr::count(level) |>
  dplyr::arrange(dplyr::desc(n)) |>
  knitr::kable(caption = "Highest completed education (ISCED)")
```

## 2. By imagery group

Imagery vividness is measured by the VVIQ [@marksVisualImageryDifferences1973],
whose sixteen items give a total between 16 and 80. **Three different
groupings of that score appear on this site**, and they are not
interchangeable, so they are set out together here.

| Grouping | Column | Definition |
|---|---|---|
| Four-way, used below | `vviq_group_4` | aphantasia = 16 (the scale floor), hypophantasia 17–32, typical 33–74, hyperphantasia ≥ 75 |
| Two-way, descriptive | `vviq_group_2` | aphantasia ≤ 32, typical ≥ 33 |
| Floor split, used by every model | not stored | at the floor (16) against everyone above it |

The four-way split divides the low end of the scale at the floor itself,
following @reederNonvisualSpatialStrategies2024. That is **narrower than
the convention a reader may assume**: the widely used aphantasia cutoff of
32 is what `vviq_group_2` applies, and it corresponds here to
aphantasia *plus* hypophantasia. The two columns therefore both have a
level called `aphantasia` and the two levels are not the same people.

The modelling pages use the third grouping, between participants at the
scale floor and everyone above it, and the [analysis
strategy](analysis-strategy.html) page explains why. It coincides with
`vviq_group_4`'s aphantasia level and not with `vviq_group_2`'s.

```{r groups}
participants |>
  dplyr::filter(!is.na(imagery_group)) |>
  dplyr::summarise(
    n = dplyr::n(),
    median_age = stats::median(age),
    women = sum(gender == "Women"),
    .by = imagery_group
  ) |>
  knitr::kable(caption = "By VVIQ group")
```

That table counts `r sum(!is.na(participants$vviq_total_score))` people
rather than `r nrow(participants)`:
`r sum(is.na(participants$vviq_total_score))` did not complete the VVIQ.
They are absent from every model on this site as well, since every model
reports a floor-group offset and neither can be placed on either side of
the floor.

```{r vviq-distribution, fig.alt="Distribution of VVIQ scores, with the floor group highlighted."}
plot_vviq_histogram(participants$vviq_total_score, base_size = 16) +
  labs(x = "VVIQ total score",
       caption = "The floor group is coloured separately.") +
  theme(axis.text.x = element_text(), axis.ticks.x = element_line(),
        axis.line.x = element_line())
```

The distribution is the reason the modelling pages treat vividness the way
they do. It is not a smooth continuum with a low tail: it is a sharp,
isolated spike at the scale minimum plus an irregular remainder above it.
A model fitting one slope through all of it would describe neither part.

## 3. Recruitment, and why the versions are not interchangeable

This is the constraint worth carrying into every other page.

```{r recruitment}
sessions |>
  dplyr::summarise(
    n = dplyr::n(),
    at_vviq_floor = sum(vviq_total_score == 16, na.rm = TRUE),
    percent_floor = round(100 * mean(vviq_total_score == 16, na.rm = TRUE)),
    typical_imagers = sum(vviq_group_2 == "typical", na.rm = TRUE),
    median_age = stats::median(age),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(caption = "By task version, counted per person-session")
```

The three versions recruited **different populations**. v1 is roughly a
quarter floor group; v2 and v3 are around two thirds, because later
recruitment deliberately targeted participants with no visual imagery,
partly through aphantasia community channels.

Two consequences run through the whole site.

**The behavioural analyses use v1 only.** Not because v2 and v3 are worse
data, but because a group comparison needs both groups, and the
`typical_imagers` column above is where that fails: v2 and v3 have single
figures. The [version scope](version-scope.html) page sets out the full
argument, including a case where pooling the three produced an association
that stratification removed.

**Most questionnaire analyses pool all three**, because the instruments
are identical and were administered identically in every version, so
version is a property of the task rather than of the scales. That decision
is checked rather than assumed on the [beyond
vividness](beyond-vividness.html) page. The exception is
[psychometrics](psychometrics.html), which stays on v1 so that its
reliabilities describe the sample the modelling pages use.

## 4. What was administered

Every participant completed the same battery in the same session: the
WM-FTT itself, then the VVIQ, the OSIVQ and the NIEQ. The
[psychometrics](psychometrics.html) page reports how those instruments
behave in this sample, including one subscale that does not reach its
reliability threshold and one scoring inconsistency between versions.

Field of study, occupation and recruitment route are recorded in
`extra_demographics` and are not analysed anywhere: they were collected for
description rather than as variables.

```{r fields}
extra |>
  dplyr::filter(!is.na(field)) |>
  dplyr::mutate(field = sub("^isced_f_[0-9]+_", "", field)) |>
  dplyr::count(field) |>
  dplyr::arrange(dplyr::desc(n)) |>
  head(6) |>
  knitr::kable(caption = "Field of study or work, six most common")
```

---

**Continuing through the Extended Online Report:** this is the first page of the report. To keep reading in order, continue to [task design](task-design.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
