---
title: "Performance: how well each feature is recalled"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
library(patchwork)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)

pkg <- "aphantasiaWMStrats"
refit <- "never"
```

How accurately is each feature recalled, and does imagery vividness predict
it? This is the question the study started from, and it is no longer the
main event: the [composition](composition.html) page reports the primary
result, because *where* effort goes is what the task was built to measure
and absolute accuracy cannot separate a deliberate choice from an
inability.

This page is the same data read the other way. Like the composition it
conditions on responded items, and whether that is innocent is tested on
the [joint model](joint-model.html) page.

Same data through a simpler lens, so agreement with the joint model is not
independent evidence. It is still worth having, because a result that
appears only under the more complex model deserves suspicion.

```{r data}
v1 <- get_data("v1") |>
  dplyr::filter(grepl("^expe_block", expe_phase))

participant_info <- v1 |>
  dplyr::summarise(
    vviq = dplyr::first(vviq_total_score),
    imagery_group = dplyr::first(vviq_group_2),
    # The proportion of parity probes ANSWERED, not a mean of
    # `parity_*_acc`: that column scores an unanswered probe as 0, so a
    # mean of it is a response rate wearing an accuracy name.
    parity_rate = (sum(responded_parity_1) + sum(responded_parity_2)) /
      (2 * dplyr::n()),
    .by = id
  )

model_data <- v1 |>
  dplyr::filter(id %in% engaged_ids(v1)) |>
  dplyr::left_join(participant_info, by = "id") |>
  dplyr::filter(!is.na(vviq)) |>
  dplyr::mutate(
    complete_aphant = factor(
      dplyr::if_else(vviq == 16, "floor", "above_floor"),
      levels = c("above_floor", "floor")
    ),
    score_color = squeeze_boundaries(score_color, n = sum(responded_color))
  )

feature_labels <- c(word = "Word", angle = "Orientation", color = "Colour")
```

This page uses the `r dplyr::n_distinct(model_data$id)` participants who
clear the engagement thresholds of the [task
validity](task-validity.html) page **and** have a vividness score, not the
86 the [joint model](joint-model.html) uses. The
`r dplyr::n_distinct(engaged_ids(v1))` who clear the thresholds include two
with no VVIQ, and every page that reports a floor-group offset needs one;
[the analysis](analysis-strategy.html#samples) sets out the whole chain. The
thresholds are there to stop imprecise participant means being treated as
measurements, and unlike the joint model this analysis does form them.

## 1. Three features, three response families

Nothing about these scores makes a single response family appropriate, and
the [scoring](scoring.html) page shows why: on responded items, word piles
at 1, orientation has no boundary mass at all, and colour has a trace of
mass at 1 that is pixel resolution rather than behaviour.

```{r families}
performance_formula()
```

Every accuracy model on this site carries `parity_rate`: it is the
dual-task load a participant actually incurred, which [task
engagement](engagement.html) §5 shows is a strategy variable in its own
right rather than a control for something it does not predict. The [joint
model](joint-model.html) puts it on its accuracy arms and not on its
gates, for the same reason, so the two pages estimate the same quantity.

Word takes a zero-one-inflated Beta because its ceiling is a real point
mass, orientation a plain Beta, and colour a Beta after a
Smithson-Verkuilen squeeze. `subset()` restricts each arm to the items
that feature was reported on, which is what lets one row carry three
responses with independent non-response.

## 2. Modelling the features separately is the wrong thing to do

Start with what looks obvious: fit each feature on its own.

```{r option-a}
# Built from the same exported `response_priors()` the script calls, rather
# than left to brms' defaults. `resp` is omitted for a univariate model,
# where coefficients belong to no response.
univariate_priors <- function(rhs) {
  terms <- c("vviq", "complete_aphant", "imagery_group", "parity_rate")
  response_priors(NULL, terms = terms[vapply(terms, grepl, logical(1), rhs)])
}

fit_feature <- function(feature, family, file) {
  rhs <- "vviq + complete_aphant + parity_rate + (1 | id)"
  fit_brms_model(
    formula = brms::bf(
      stats::as.formula(paste0(
        "score_", feature, " | subset(responded_", feature, ") ~ ", rhs)),
      family = family
    ),
    data = model_data,
    prior = univariate_priors(rhs),
    file = system.file("models", file, package = pkg),
    file_refit = refit
  )
}

independent <- list(
  Word = fit_feature("word", brms::zero_one_inflated_beta(), "perf-a-word.rds"),
  Orientation = fit_feature("angle", brms::Beta(), "perf-a-angle.rds"),
  Colour = fit_feature("color", brms::Beta(), "perf-a-color.rds")
)
```

The task makes that false by construction. Points are traded across
features within a trial, so a participant who does well on one has spent
effort they could not spend elsewhere. Three independent models cannot
represent that, and more importantly they cannot **measure** it.

```{r c-prime}
# lkj(2) mildly favours the identity, so the cross-feature correlations
# have to be earned from the data rather than assumed. It is passed here
# rather than left to brms' default lkj(1), because section 3 reads the
# posteriors against this prior and a different one would make that read
# wrong.
c_prime_priors <- c(
  do.call(c, lapply(
    c("scoreword", "scoreangle", "scorecolor"),
    response_priors, terms = c("vviq", "complete_aphant", "parity_rate")
  )),
  brms::prior("lkj(2)", class = "cor")
)

correlated <- fit_brms_model(
  formula = performance_formula(),
  data    = model_data,
  prior   = c_prime_priors,
  file    = system.file("models", "perf-c-prime.rds", package = pkg),
  file_refit = refit
)
```

The correlated model adds `(1 | p | id)`, which gives the three features a
shared participant-level correlation matrix. That is the whole difference,
and it buys two things: a dependency structure the design implies, and
three named correlations that are themselves the quantity of interest.

```{r contrast, fig.alt="Floor-group offsets from independent and correlated models."}
offsets <- dplyr::bind_rows(
  purrr::imap(independent, \(model, feature) {
    values <- brms::fixef(model)["complete_aphantfloor", ]
    tibble::tibble(feature = feature, model = "Independent",
                   estimate = values[["Estimate"]],
                   lower = values[["Q2.5"]], upper = values[["Q97.5"]])
  }) |> purrr::list_rbind(),
  purrr::map(names(feature_labels), \(feature) {
    row <- paste0("score", feature, "_complete_aphantfloor")
    values <- brms::fixef(correlated)[row, ]
    tibble::tibble(feature = feature_labels[[feature]],
                   model = "Correlated",
                   estimate = values[["Estimate"]],
                   lower = values[["Q2.5"]], upper = values[["Q97.5"]])
  }) |> purrr::list_rbind()
) |>
  dplyr::mutate(feature = factor(feature, levels = unname(feature_labels)))

ggplot(offsets, aes(x = estimate, y = feature, colour = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_pointrange(aes(xmin = lower, xmax = upper),
                  position = position_dodge(width = 0.5), size = 0.4) +
  scale_colour_manual(values = unname(palette.colors()[c(6, 2)]), name = NULL) +
  labs(x = "Floor-group offset, log-odds", y = NULL,
       caption = "The same quantity, estimated two ways.")
```

**The coefficients barely move.** That is worth saying plainly rather than
implying the simpler model was misleading: on these data, respecting the
dependency structure changes the floor-group estimates very little. What it
adds is not a correction but a measurement.

## 3. What the correlations say

```{r correlations}
posterior_correlations(brms::as_draws_df(correlated), lkj = 2, dimension = 3) |>
  dplyr::mutate(
    pair = paste(feature_labels[sub("^score", "", response_a)], "and",
                 feature_labels[sub("^score", "", response_b)])
  ) |>
  dplyr::select(pair, median, lower, upper, pd, moved) |>
  knitr::kable(digits = 3, caption = "Participant-level correlations between features")
```

Orientation and colour correlate **positively**, and clearly. Between
people, doing well on one non-verbal feature goes with doing well on the
other: shared ability, not a trade-off.

This matters because the raw data suggests the opposite. Working from
proportions, orientation and colour appear to trade off, and the
[scoring](scoring.html) page shows why that reading is an artifact:
residualising three parts on their own mean induces negative correlation
whatever the data says. Here the parts are not closed, so a negative
correlation would have meant something. It is positive.

**The trade-off the task imposes is a within-trial constraint, not a
between-person one.** The [joint model](joint-model.html) reaches the same
conclusion with the gates in place.

## 4. How imagery enters

Vividness is not a smoothly continuous predictor here, so rather than
assume a form, several were compared on orientation, which is the
best-measured feature.

```{r comparison}
form_models <- list(
  "Intercept only" = fit_brms_model(
    formula = brms::bf(
      score_angle | subset(responded_angle) ~ parity_rate + (1 | id),
      family = brms::Beta()),
    data = model_data,
    prior = univariate_priors("parity_rate + (1 | id)"),
    file = system.file("models", "perf-form-null.rds", package = pkg),
    file_refit = refit),
  "Two-group split" = fit_brms_model(
    formula = brms::bf(
      score_angle | subset(responded_angle) ~
        imagery_group + parity_rate + (1 | id),
      family = brms::Beta()),
    data = model_data,
    prior = univariate_priors("imagery_group + parity_rate + (1 | id)"),
    file = system.file("models", "perf-form-group.rds", package = pkg),
    file_refit = refit),
  "Linear vividness" = fit_brms_model(
    formula = brms::bf(
      score_angle | subset(responded_angle) ~ vviq + parity_rate + (1 | id),
      family = brms::Beta()),
    data = model_data,
    prior = univariate_priors("vviq + parity_rate + (1 | id)"),
    file = system.file("models", "perf-form-linear.rds", package = pkg),
    file_refit = refit),
  "Floor-group additive" = independent$Orientation
)

form_models <- lapply(form_models, brms::add_criterion, "loo")
comparison <- loo::loo_compare(lapply(form_models, \(fit) fit$criteria$loo))
if (!"model" %in% names(comparison)) {
  comparison <- tibble::rownames_to_column(as.data.frame(comparison), "model")
}
comparison |>
  dplyr::select(model, elpd_diff, se_diff) |>
  knitr::kable(digits = 2, caption = "Functional form, compared on orientation")
```

The differences sit inside their standard errors, so **the data do not
separate these forms**. That was predicted in advance, and it is reported
rather than converted into a claim: at this sample size, with this outcome,
a comparison of four models was never going to be decisive.

The floor-group form remains primary because it was declared so before
fitting, not because it won. Its justification is structural: participants
at the scale floor have no vividness variance among themselves, so a slope
cannot be estimated for them, while an offset can.

```{r floor-figure, fig.alt="Orientation accuracy against imagery vividness, with the floor group shown separately.", fig.height = 6}
orientation <- independent$Orientation

grid <- tibble::tibble(
  vviq = seq(16, 80, length.out = 200),
  complete_aphant = factor("above_floor", levels = c("above_floor", "floor")),
  # held at the sample median: the model carries parity_rate, so a grid
  # without it cannot be predicted from at all
  parity_rate = stats::median(model_data$parity_rate),
  responded_angle = TRUE
)
above_floor <- brms::posterior_epred(orientation, newdata = grid,
                                     re_formula = NA)

floor_row <- grid[1, ]
floor_row$complete_aphant <- factor("floor",
                                    levels = c("above_floor", "floor"))
floor_draws <- as.vector(
  brms::posterior_epred(orientation, newdata = floor_row, re_formula = NA))

observed <- model_data |>
  dplyr::filter(responded_angle) |>
  dplyr::summarise(y = mean(score_angle), vviq = dplyr::first(vviq), .by = id)

effect <- brms::fixef(orientation)["complete_aphantfloor", ]

main_panel <- plot_floor_group(
  observed = dplyr::transmute(dplyr::filter(observed, vviq > 16),
                              x = vviq, y = y),
  fitted = tibble::tibble(
    x = grid$vviq,
    estimate = apply(above_floor, 2, stats::median),
    lower = apply(above_floor, 2, stats::quantile, 0.025),
    upper = apply(above_floor, 2, stats::quantile, 0.975)
  ),
  floor_draws = floor_draws,
  floor_observed = dplyr::filter(observed, vviq == 16)$y,
  effect_label = sprintf("%.2f\n[%.2f, %.2f]", effect[["Estimate"]],
                         effect[["Q2.5"]], effect[["Q97.5"]]),
  y_lab = "Orientation recall accuracy",
  caption = "Offset in log-odds. Points coloured by vividness.",
  base_size = 16, point_size = 2.2, left_expansion = 0.15,
  arrow_nudge = -5.5
)

plot_vviq_histogram(observed$vviq, base_size = 16, left_expansion = 0.15) /
  main_panel + patchwork::plot_layout(heights = c(1, 4))
```

The histogram is the argument for the model: vividness is a spike at the
floor plus an irregular remainder, not a continuum. In the panel the arrow
has almost nothing to span, which is the result.

## 5. Where the floor group does and does not differ

```{r floor-table}
purrr::map(names(feature_labels), \(feature) {
  row <- paste0("score", feature, "_complete_aphantfloor")
  values <- brms::fixef(correlated)[row, ]
  tibble::tibble(feature = feature_labels[[feature]],
                 estimate = values[["Estimate"]],
                 lower = values[["Q2.5"]], upper = values[["Q97.5"]])
}) |>
  purrr::list_rbind() |>
  knitr::kable(digits = 3, caption = "Floor-group offsets, log-odds")
```

Word is positive and excludes zero, colour is negative and touches it,
orientation is centred near nothing.

**Word's row must not be read as an individual-differences result.** Its
split-half reliability is 0.45, and the [task
validity](task-validity.html) page sets out why: single-word recall at this
exposure is too easy to discriminate between people. A group contrast is a
different estimand from an individual-differences claim, and measurement
error widens intervals rather than biasing estimates, so the offset is not
*wrong*. But this page is not where that claim should be made, and the
[composition](composition.html) page reaches the same conclusion through a
quantity that does clear its reliability gate.

## 6. Partial pooling, made visible

A multilevel model does not take each participant's mean at face value: it
pulls imprecise estimates toward the group. Worth seeing rather than
asserting, because on one feature it is dramatic.

```{r shrinkage, fig.alt="Observed against model-implied participant means for the three features."}
observed_means <- compose_features(model_data, id) |>
  dplyr::select(id, tidyselect::starts_with("mean_")) |>
  tidyr::pivot_longer(tidyselect::starts_with("mean_"), names_to = "feature",
                      values_to = "observed", names_prefix = "mean_")

fitted_means <- purrr::map(names(feature_labels), \(feature) {
  responded <- model_data[[paste0("responded_", feature)]]
  epred <- brms::posterior_epred(correlated, resp = paste0("score", feature))
  tibble::tibble(
    id = model_data$id[responded],
    feature = feature,
    modelled = apply(epred, 2, mean)
  ) |>
    dplyr::summarise(modelled = mean(modelled), .by = c(id, feature))
}) |>
  purrr::list_rbind()

dplyr::inner_join(observed_means, fitted_means, by = c("id", "feature")) |>
  dplyr::mutate(feature = factor(feature_labels[feature],
                                 levels = unname(feature_labels))) |>
  ggplot(aes(x = observed, y = modelled, colour = feature)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.3) +
  geom_point(size = 1.6, alpha = 0.7, show.legend = FALSE) +
  facet_wrap(~feature) +
  scale_discrete_feature(aesthetics = "colour") +
  labs(x = "Observed mean, responded items", y = "Model-implied mean",
       caption = "Points off the diagonal have been pulled toward the group.")
```

Orientation and colour track the diagonal with modest shrinkage, which is
what partial pooling normally looks like.

**Word does not track it at all.** Observed means run from about 0.78 to
1.00, and the model returns essentially the same estimate for everybody.
That is not a bug: it is the model saying it cannot distinguish these
participants, and it is the same fact as word's split-half reliability of
0.45 seen from another angle. A figure of this shape is what an unusable
individual-differences measure looks like, and it is the clearest reason on
this page not to read word's coefficient as one.

---

**Continuing through the Extended Online Report:** this page follows [composition](composition.html). To keep reading in order, continue to [the joint model](joint-model.html) next. Or see [model diagnostics](model-diagnostics.html) and [implementation notes](implementation-notes.html) for the technical detail behind these models.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
