---
title: "Psychometrics: how the questionnaires behave in this sample"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Psychometrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_minimal(base_size = 16))
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

Three questionnaires accompany the WM-FTT: the VVIQ [@marksVisualImageryDifferences1973],
the OSIVQ [@blazhenkovaNewObjectSpatialVerbal2009] and the NIEQ
[@heaveyMeasuringFrequencyInnerExperience2019]. All three feed downstream
analyses, and two of them turned out to need caveats. This page reports how
each behaves in this sample rather than citing the reliabilities from their
manuals, because one of them does not reproduce.

The task's own reliabilities are on the [Scoring](scoring.html) page, where
the split-half machinery lives. Everything below is computed when this page
is built.

```{r data}
participants <- get_data("v1") |> dplyr::distinct(id, .keep_all = TRUE)

# Cronbach's alpha from raw items. Written out rather than pulled from a
# package so the formula is visible: k/(k-1) * (1 - sum(item variances) /
# variance of the total).
cronbach_alpha <- function(items) {
  items <- items[stats::complete.cases(items), , drop = FALSE]
  n_items <- ncol(items)
  n_items / (n_items - 1) *
    (1 - sum(apply(items, 2, stats::var)) / stats::var(rowSums(items)))
}

# Corrected item-total correlation: each item against the sum of the others.
item_total <- function(items) {
  items <- items[stats::complete.cases(items), , drop = FALSE]
  total <- rowSums(items)
  purrr::map_dbl(names(items), \(item) stats::cor(items[[item]], total - items[[item]]))
}
```

## 1. VVIQ

Sixteen items, no reverse-keyed items, single dimension.

```{r vviq}
vviq_items <- purrr::list_rbind(participants$vviq_items)

tibble::tibble(
  instrument = "VVIQ",
  items = ncol(vviq_items),
  alpha = cronbach_alpha(vviq_items),
  min_item_total = min(item_total(vviq_items)),
  negative_item_totals = sum(item_total(vviq_items) < 0)
) |>
  knitr::kable(digits = 3, caption = "VVIQ internal consistency, v1")
```

Nothing to report, which is the useful result: alpha is high, every item
correlates positively with the rest, and the weakest item still exceeds
0.75. The VVIQ is doing its job here.

The distribution is another matter, and matters more for modelling than
the reliability does:

```{r vviq-distribution}
participants |>
  dplyr::filter(!is.na(vviq_total_score)) |>
  ggplot(aes(vviq_total_score)) +
  geom_histogram(binwidth = 4, boundary = 16) +
  labs(x = "VVIQ total score", y = "Participants",
       title = "VVIQ is close to bimodal in v1")
```

```{r vviq-floor}
participants |>
  dplyr::filter(!is.na(vviq_total_score)) |>
  dplyr::count(band = cut(vviq_total_score, c(15, 16, 25, 40, 60, 80),
                          include.lowest = TRUE)) |>
  knitr::kable(caption = "VVIQ distribution: a floor spike and a sparse middle")
```

Roughly a quarter of the sample sits at exactly 16, the scale minimum, and
the middle of the range is nearly empty. That is why VVIQ is not treated as
a plain continuous predictor everywhere downstream; see the floor-group
model in the planning documentation.

## 2. OSIVQ, and a scoring inconsistency worth knowing about

Three subscales: object, spatial, verbal. **Four items are reverse-keyed,
and they are stored inconsistently across task versions.** v1 and v2 store
`osivq_q02v`, `osivq_q09v`, `osivq_q41v` and `osivq_q42s` unreversed; v3
stores them already reversed. The `object_mean`, `spatial_mean` and
`verbal_mean` columns are correct in both cases. Only the raw items in
`osivq_items` differ.

The consequence is silent, so it is worth showing:

```{r osivq-reversal}
reverse_keyed <- c("osivq_q02v", "osivq_q09v", "osivq_q41v", "osivq_q42s")
osivq_raw <- purrr::list_rbind(participants$osivq_items)
osivq_fixed <- osivq_raw
for (item in reverse_keyed) osivq_fixed[[item]] <- 6 - osivq_fixed[[item]]

subscale_items <- function(items, suffix) {
  items[, grepl(paste0(suffix, "$"), names(items)), drop = FALSE]
}

tibble::tibble(
  subscale = c("object", "spatial", "verbal"),
  suffix = c("o", "s", "v")
) |>
  dplyr::mutate(
    items = purrr::map_int(suffix, \(s) ncol(subscale_items(osivq_raw, s))),
    alpha_as_stored = purrr::map_dbl(suffix, \(s) cronbach_alpha(subscale_items(osivq_raw, s))),
    alpha_corrected = purrr::map_dbl(suffix, \(s) cronbach_alpha(subscale_items(osivq_fixed, s)))
  ) |>
  dplyr::select(-suffix) |>
  knitr::kable(digits = 3,
               caption = "OSIVQ alpha in v1, before and after reversing the reverse-keyed items")
```

Computed naively from the stored items, the verbal subscale looks broken.
Corrected, all three are healthy. The three items driving it are visible in
their item-total correlations:

```{r osivq-item-total}
verbal_raw <- subscale_items(osivq_raw, "v")
tibble::tibble(item = names(verbal_raw),
               item_total_as_stored = item_total(verbal_raw)) |>
  dplyr::arrange(item_total_as_stored) |>
  knitr::kable(digits = 2,
               caption = "OSIVQ verbal items: negative correlations mark the unreversed ones")
```

Anyone recomputing OSIVQ scores from the item columns needs to reverse
those four for v1 and v2, and not for v3. The subscale mean columns already
handle it.

## 3. NIEQ

Ten items: five dimensions, each measured by a "frequently" item and a
"generally" item, scored as the pair's mean.

**A total across all ten items is meaningless**, since the dimensions are
not facets of one construct. Someone with frequent inner speech need not
have frequent imagery. So Cronbach's alpha over the full instrument is not
the right statistic, and computing it would produce a number that looks
like a defect but is not one. The right check is whether each pair agrees.

```{r nieq}
nieq_items <- purrr::list_rbind(participants$nieq_items)

# Two-item reliability is the Spearman-Brown correction of the pair's
# correlation, not alpha.
spearman_brown <- function(r) 2 * r / (1 + r)

nieq_pairs <- tibble::tibble(
  dimension = c("Mental imagery", "Inner voice", "Emotions",
                "Sensory focus", "Unsymbolised thinking"),
  stem = c("mental_imagery", "inner_voice", "emotions",
           "sensory_focus", "unsymbolised")
) |>
  dplyr::mutate(
    pair_r = purrr::map_dbl(stem, \(s) stats::cor(
      nieq_items[[paste0("nieq_freq_", s)]],
      nieq_items[[paste0("nieq_prop_", s)]],
      use = "complete.obs")),
    reliability = spearman_brown(pair_r)
  ) |>
  dplyr::select(-stem)

knitr::kable(nieq_pairs, digits = 3,
             caption = "NIEQ two-item subscale reliability, v1")
```

```{r nieq-values, include = FALSE}
reliability_of <- function(d) nieq_pairs$reliability[nieq_pairs$dimension == d]
```

Four of the five dimensions are adequate. **Sensory focus is not**, at
`r fmt(reliability_of("Sensory focus"))`, well below the conventional 0.70
floor: its two items barely agree, so the subscale mean is not measuring
one thing. That is consistent with sensory awareness being the weakest
dimension in the instrument's own validation, and it means sensory focus
should not carry the same weight as the others wherever NIEQ subscales are
used as inputs.

Unsymbolised thinking, at `r fmt(reliability_of("Unsymbolised thinking"))`,
is adequate. That matters more than it might appear: it is the dimension
with the clearest theoretical stake in this project, and a two-item
subscale is exactly the kind of measure that can quietly fail to support
the weight placed on it.

## 4. Summary

```{r summary}
dplyr::bind_rows(
  tibble::tibble(instrument = "VVIQ", scale = "total",
                 reliability = cronbach_alpha(vviq_items), method = "alpha, 16 items"),
  tibble::tibble(
    instrument = "OSIVQ",
    scale = c("object", "spatial", "verbal"),
    reliability = purrr::map_dbl(c("o", "s", "v"),
                                 \(s) cronbach_alpha(subscale_items(osivq_fixed, s))),
    method = "alpha, reverse-keyed items corrected"),
  nieq_pairs |>
    dplyr::transmute(instrument = "NIEQ", scale = dimension,
                     reliability, method = "Spearman-Brown, 2 items")
) |>
  dplyr::mutate(adequate = dplyr::if_else(reliability >= 0.70, "yes", "NO")) |>
  knitr::kable(digits = 3, caption = "All questionnaire scales, v1")
```

One scale of nine falls below the threshold. The OSIVQ reverse-coding
inconsistency does not affect any stored score, but does affect anyone
recomputing from items. Neither problem is visible without checking, which
is the argument for this page existing.

## References

<div id="refs"></div>

---

**Continuing through the Extended Online Report:** this page follows [task validity](task-validity.html). To keep reading in order, continue to [the analysis](analysis-strategy.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
