---
title: "Psychometrics: how the questionnaires behave in this sample"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Psychometrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_minimal(base_size = 16))
fmt <- function(x, digits = 2) formatC(x, format = "f", digits = digits)
```

Three questionnaires accompany the WM-FTT: the VVIQ
[@marksVisualImageryDifferences1973], the OSIVQ
[@blazhenkovaNewObjectSpatialVerbal2009] and the NIEQ
[@heaveyMeasuringFrequencyInnerExperience2019]. All three feed downstream
analyses, and two of them turned out to need caveats. This page reports how each
behaves in this sample.

The task's own reliabilities are on the [task validity](task-validity.html)
page, where the split-half machinery lives. Everything below is computed when
this page is built.

**This page stays on v1**, where the [beyond vividness](beyond-vividness.html)
page pools all three versions. The two are asking different questions. Pooling
is right when the aim is to describe the instruments in as many people as
possible; here the aim is to say whether the scales are trustworthy *in the
sample the modelling pages use*, and that sample is v1. Where a claim below
rests on the other versions, it says so.

```{r data}
participants <- get_data("v1") |> dplyr::distinct(id, .keep_all = TRUE)

# Cronbach's alpha from raw items. Written out rather than pulled from a
# package so the formula is visible: k/(k-1) * (1 - sum(item variances) /
# variance of the total).
cronbach_alpha <- function(items) {
  items <- items[stats::complete.cases(items), , drop = FALSE]
  n_items <- ncol(items)
  n_items / (n_items - 1) *
    (1 - sum(apply(items, 2, stats::var)) / stats::var(rowSums(items)))
}

# Corrected item-total correlation: each item against the sum of the others.
item_total <- function(items) {
  items <- items[stats::complete.cases(items), , drop = FALSE]
  total <- rowSums(items)
  purrr::map_dbl(names(items), \(item) stats::cor(items[[item]], total - items[[item]]))
}
```

## 1. VVIQ {#psy-vviq}

Sixteen items, no reverse-keyed items, single dimension.

```{r vviq}
vviq_items <- purrr::list_rbind(participants$vviq_items)

tibble::tibble(
  instrument = "VVIQ",
  items = ncol(vviq_items),
  alpha = cronbach_alpha(vviq_items),
  min_item_total = min(item_total(vviq_items)),
  negative_item_totals = sum(item_total(vviq_items) < 0)
) |>
  knitr::kable(digits = 3, caption = "VVIQ internal consistency, v1")
```

Nothing to report, which is the useful result: alpha is high, every item
correlates positively with the rest, and the weakest item still exceeds 0.75.
The VVIQ is doing its job here.

The distribution is another matter, and matters more for modelling than the
reliability does:

```{r vviq-distribution}
participants |>
  dplyr::filter(!is.na(vviq_total_score)) |>
  ggplot(aes(vviq_total_score)) +
  geom_histogram(binwidth = 4, boundary = 16) +
  labs(x = "VVIQ total score", y = "Participants",
       title = "VVIQ is close to bimodal in v1")
```

```{r vviq-floor}
participants |>
  dplyr::filter(!is.na(vviq_total_score)) |>
  dplyr::count(band = cut(vviq_total_score, c(15, 16, 25, 40, 60, 80),
                          include.lowest = TRUE)) |>
  knitr::kable(caption = "VVIQ distribution: a floor spike and a sparse middle")
```

Roughly a quarter of the sample sits at exactly 16, the scale minimum, and the
middle of the range is sparsely populated. That is why VVIQ is not treated as a
plain continuous predictor everywhere downstream; [the analysis
strategy](analysis-strategy.html) page §3 sets out the floor-group form every
model uses and the caution that comes with it.

## 2. OSIVQ {#psy-osivq}

Three subscales: object, spatial, verbal. **Four items are reverse-keyed, and
they are stored inconsistently across task versions.** v1 and v2 store
`osivq_q02v`, `osivq_q09v`, `osivq_q41v` and `osivq_q42s` un-reversed; v3 stores
them already reversed. The `object_mean`, `spatial_mean` and `verbal_mean`
columns are correct in both cases. Only the raw items in `osivq_items` differ.

That claim spans versions while the rest of this page is v1, so it is checked
directly rather than asserted: if v3 really does store them reversed, its
item-total correlations for those items should be **positive** where v1's are
negative.

```{r reversal-across-versions}
reverse_keyed <- c("osivq_q02v", "osivq_q09v", "osivq_q41v", "osivq_q42s")

item_total_by_version <- function(v) {
  people <- get_data(v) |> dplyr::distinct(id, .keep_all = TRUE)
  items <- purrr::list_rbind(people$osivq_items)
  suffix <- sub("^.*q[0-9]+", "", reverse_keyed)
  purrr::map2_dbl(reverse_keyed, suffix, function(item, s) {
    block <- items[, grepl(paste0(s, "$"), names(items)), drop = FALSE]
    block <- block[stats::complete.cases(block), , drop = FALSE]
    stats::cor(block[[item]], rowSums(block) - block[[item]])
  })
}

tibble::tibble(item = reverse_keyed) |>
  dplyr::mutate(
    v1 = item_total_by_version("v1"),
    v3 = item_total_by_version("v3")
  ) |>
  knitr::kable(digits = 2,
               caption = "Corrected item-total correlations for the four reverse-keyed items")
```

Negative in v1, positive in v3. The items are the same; only the storage
convention differs.

The consequence within v1:

```{r osivq-reversal}
osivq_raw <- purrr::list_rbind(participants$osivq_items)
osivq_fixed <- osivq_raw
for (item in reverse_keyed) osivq_fixed[[item]] <- 6 - osivq_fixed[[item]]

subscale_items <- function(items, suffix) {
  items[, grepl(paste0(suffix, "$"), names(items)), drop = FALSE]
}

tibble::tibble(
  subscale = c("object", "spatial", "verbal"),
  suffix = c("o", "s", "v")
) |>
  dplyr::mutate(
    items = purrr::map_int(suffix, \(s) ncol(subscale_items(osivq_raw, s))),
    alpha_as_stored = purrr::map_dbl(suffix, \(s) cronbach_alpha(subscale_items(osivq_raw, s))),
    alpha_corrected = purrr::map_dbl(suffix, \(s) cronbach_alpha(subscale_items(osivq_fixed, s)))
  ) |>
  dplyr::select(-suffix) |>
  knitr::kable(digits = 3,
               caption = "OSIVQ alpha in v1, before and after reversing the reverse-keyed items")
```

Computed naively from the stored items, the verbal subscale looks broken.
Corrected, all three are healthy. The three items driving it are visible in
their item-total correlations:

```{r osivq-item-total}
verbal_raw <- subscale_items(osivq_raw, "v")
tibble::tibble(item = names(verbal_raw),
               item_total_as_stored = item_total(verbal_raw)) |>
  dplyr::arrange(item_total_as_stored) |>
  knitr::kable(digits = 2,
               caption = "OSIVQ verbal items: negative correlations mark the unreversed ones")
```

Anyone recomputing OSIVQ scores from the item columns needs to reverse those
four for v1 and v2, and not for v3. The subscale mean columns already handle it.

## 3. NIEQ {#psy-nieq}

Ten items: five dimensions, each measured by a "frequently" item and a
"generally" item, scored as the pair's mean.

**A total across all ten items is meaningless**, since the dimensions are
different constructs, not facets of a single construct. Someone with frequent
inner speech need not have frequent imagery. So Cronbach's alpha over the full
instrument is not the right statistic, and computing it would produce a number
that looks like a defect but is not one. The right check is whether each pair
agrees.

```{r nieq}
nieq_items <- purrr::list_rbind(participants$nieq_items)

# Two-item reliability is the Spearman-Brown correction of the pair's
# correlation, not alpha. Written out for the same reason as
# `cronbach_alpha()` above: this page's job is to show the arithmetic, so
# both formulas are visible rather than called from the package. The
# exported `spearman_brown()` is identical and is what the scripts use.
spearman_brown <- function(r) 2 * r / (1 + r)

nieq_pairs <- tibble::tibble(
  dimension = c("Mental imagery", "Inner voice", "Emotions",
                "Sensory focus", "Unsymbolised thinking"),
  stem = c("mental_imagery", "inner_voice", "emotions",
           "sensory_focus", "unsymbolised")
) |>
  dplyr::mutate(
    pair_r = purrr::map_dbl(stem, \(s) stats::cor(
      nieq_items[[paste0("nieq_freq_", s)]],
      nieq_items[[paste0("nieq_prop_", s)]],
      use = "complete.obs")),
    reliability = spearman_brown(pair_r)
  ) |>
  dplyr::select(-stem)

knitr::kable(nieq_pairs, digits = 3,
             caption = "NIEQ two-item subscale reliability, v1")
```

```{r nieq-values, include = FALSE}
reliability_of <- function(d) nieq_pairs$reliability[nieq_pairs$dimension == d]
```

Four of the five dimensions are adequate. **Sensory focus is not**, at
`r fmt(reliability_of("Sensory focus"))`, well below the conventional 0.70
floor: its two items barely agree, so the subscale mean is not measuring one
thing. That is consistent with sensory awareness being the weakest dimension in
the instrument's own validation, and it means sensory focus should not carry the
same weight as the others wherever NIEQ subscales are used as inputs.

Unsymbolised thinking, at `r fmt(reliability_of("Unsymbolised thinking"))`, is
adequate. It is the dimension with the clearest theoretical stake in this
project, so it is important to assess whether it can legitimately support the
weight placed on it.

## 4. Summary {#psy-summary}

```{r summary}
dplyr::bind_rows(
  tibble::tibble(instrument = "VVIQ", scale = "total",
                 reliability = cronbach_alpha(vviq_items), method = "alpha, 16 items"),
  tibble::tibble(
    instrument = "OSIVQ",
    scale = c("object", "spatial", "verbal"),
    reliability = purrr::map_dbl(c("o", "s", "v"),
                                 \(s) cronbach_alpha(subscale_items(osivq_fixed, s))),
    method = "alpha, reverse-keyed items corrected"),
  nieq_pairs |>
    dplyr::transmute(instrument = "NIEQ", scale = dimension,
                     reliability, method = "Spearman-Brown, 2 items")
) |>
  dplyr::mutate(adequate = dplyr::if_else(reliability >= 0.70, "yes", "NO")) |>
  knitr::kable(digits = 3, caption = "All questionnaire scales, v1")
```

One scale of nine falls below the threshold. Whether the others enter the models
and how was decided based on our hypotheses on each scale, which is detailed 
further in the [composition](composition.html) analyses page.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows [task
validity](task-validity.html). To keep reading in order, continue to [the
analysis](analysis-strategy.html) next, which lays out the bases of all further
analyses in this project.

--------------------------------------------------------------------------------

## References {#psy-refs}

::: {#refs}
:::

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
