---
title: "The sample: who took part"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Sample description}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 4.5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
```

Who these participants are, and one thing about how they were recruited
that constrains what any analysis here can claim.

This page describes **all `r nrow(all_data |> dplyr::distinct(id))`
participants across the three task versions**. The behavioural analyses
use v1 alone and the questionnaire analyses pool all three; section 3 is
where that split comes from.

```{r data}
participants <- all_data |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::mutate(
    imagery_group = factor(
      vviq_group_4,
      levels = c("aphantasia", "hypophantasia", "typical", "hyperphantasia")
    ),
    # Free-text gender, harmonised. Seven raw values across two languages
    # describe three categories; the raw column is kept in `all_data`.
    gender = dplyr::case_when(
      gender == "f" ~ "Women",
      gender == "m" ~ "Men",
      TRUE ~ "Non-binary or self-described"
    )
  )

extra <- dplyr::bind_rows(participants$extra_demographics)
```

## 1. Overall

```{r overall}
# Named `n_participants`, not `participants`: inside tibble() later
# expressions see earlier columns, so a column called `participants` would
# mask the data frame of the same name and every line below it would be
# operating on a count.
tibble::tibble(
  n_participants = nrow(participants),
  median_age = stats::median(participants$age),
  age_range = paste(range(participants$age), collapse = " to "),
  french_speaking = sum(participants$language == "fr"),
  in_France = sum(extra$country == "fra", na.rm = TRUE)
) |>
  knitr::kable(caption = "The pooled sample, across all three task versions")
```

```{r gender}
participants |>
  dplyr::count(gender) |>
  dplyr::arrange(dplyr::desc(n)) |>
  knitr::kable(caption = "Gender")
```

A predominantly French, predominantly women sample, skewed toward higher
education: `r sum(grepl("master|doctorat", extra$education))` of
`r nrow(extra)` hold a master's degree or a doctorate. That matters mainly
for generalisation, and it is the usual profile for online volunteer
studies of this kind.

```{r education}
extra |>
  dplyr::filter(!is.na(education)) |>
  dplyr::mutate(
    level = factor(sub("^isced_[0-9]+_", "", education))
  ) |>
  dplyr::count(level) |>
  dplyr::arrange(dplyr::desc(n)) |>
  knitr::kable(caption = "Highest completed education (ISCED)")
```

## 2. By imagery group

Imagery vividness is measured by the VVIQ, and the four-way split below is
the conventional one. The modelling pages use a **two-way** split instead,
between participants at the scale floor and everyone above it, and the
[analysis strategy](analysis-strategy.html) page explains why.

```{r groups}
participants |>
  dplyr::filter(!is.na(imagery_group)) |>
  dplyr::summarise(
    n = dplyr::n(),
    median_age = stats::median(age),
    women = sum(gender == "Women"),
    .by = imagery_group
  ) |>
  knitr::kable(caption = "By VVIQ group")
```

```{r vviq-distribution, fig.alt="Distribution of VVIQ scores, with the floor group highlighted."}
plot_vviq_histogram(participants$vviq_total_score, base_size = 16) +
  labs(x = "VVIQ total score",
       caption = "The floor group is coloured separately.") +
  theme(axis.text.x = element_text(), axis.ticks.x = element_line(),
        axis.line.x = element_line())
```

The distribution is the reason the modelling pages treat vividness the way
they do. It is not a smooth continuum with a low tail: it is a sharp,
isolated spike at the scale minimum plus an irregular remainder above it.
A model fitting one slope through all of it would describe neither part.

## 3. Recruitment, and why the versions are not interchangeable

This is the constraint worth carrying into every other page.

```{r recruitment}
participants |>
  dplyr::summarise(
    n = dplyr::n(),
    at_vviq_floor = sum(vviq_total_score == 16, na.rm = TRUE),
    percent_floor = round(100 * mean(vviq_total_score == 16, na.rm = TRUE)),
    median_age = stats::median(age),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(caption = "By task version")
```

The three versions recruited **different populations**. v1 is roughly a
quarter floor group; v2 and v3 are around two thirds, because later
recruitment deliberately targeted participants with no visual imagery,
partly through aphantasia community channels.

Two consequences run through the whole site.

**The behavioural analyses use v1 only.** Not because v2 and v3 are worse
data, but because a group comparison needs both groups: v2 has nine
participants and v3 has twenty, with four and one typical imagers
respectively. The [version scope](version-scope.html) page sets out the
full argument, including a case where pooling the three produced an
association that stratification removed.

**The questionnaire analyses pool all three**, because the instruments are
identical and were administered identically in every version, so version
is a property of the task rather than of the scales. That decision is
checked rather than assumed on the [beyond
vividness](beyond-vividness.html) page.

## 4. What was administered

Every participant completed the same battery in the same session: the
WM-FTT itself, then the VVIQ, the OSIVQ and the NIEQ. The
[psychometrics](psychometrics.html) page reports how those instruments
behave in this sample, including one subscale that does not reach its
reliability threshold and one scoring inconsistency between versions.

Field of study, occupation and recruitment route are recorded in
`extra_demographics` and are not analysed anywhere: they were collected for
description rather than as variables.

```{r fields}
extra |>
  dplyr::filter(!is.na(field)) |>
  dplyr::mutate(field = sub("^isced_f_[0-9]+_", "", field)) |>
  dplyr::count(field) |>
  dplyr::arrange(dplyr::desc(n)) |>
  head(6) |>
  knitr::kable(caption = "Field of study or work, six most common")
```

---

**Continuing through the Extended Online Report:** this is the first page of the report. To keep reading in order, continue to [task design](task-design.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
