---
title: "Scoring the WM-FTT: what the numbers are, and what they support"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Scoring the WM-FTT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_minimal(base_size = 16))
set.seed(20260820)

n_splits <- 1000
trials_per_feature <- 63L

as_pct <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

The Working Memory Feature Trade-off Task asks participants to hold three
features of each stimulus (a word, the orientation of the rectangle
containing it, and its colour) and to report all three at recall. Scoring
it is less obvious than it looks. This page documents how the scores in
`all_data` are computed, what the distributions look like, how reliable
they are, and which claims they will and will not support.

Every number below is computed from `all_data` when this page is built,
not transcribed from an analysis run elsewhere.

## The data this page works from

`all_data` is stimulus-level: one row per presented stimulus, with all
three recalled features side by side in separate columns. Almost every
question below is about *features*, so the first step is to reshape to one
row per stimulus per feature. Three frames are built once here and reused
throughout.

```{r data}
# 1. Experimental-block rows only. Tutorial rows carry placeholder values,
#    and training rows precede the blocks by design.
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase)) |>
  dplyr::mutate(
    participant = paste(id, version, sep = "_"),  # one id spans two versions
    trial_id = paste(expe_phase, trial_number, sep = "_"),
    position_in_trial = (item_number - 1L) %% 3L + 1L
  )

# 2. One row per stimulus per feature: score, whether it was answered.
trials_by_feature <- block_trials |>
  dplyr::select(
    id, participant, version, trial_id, position_in_trial,
    vviq = vviq_total_score,
    score_word, score_angle, score_color,
    responded_word, responded_angle, responded_color
  ) |>
  tidyr::pivot_longer(
    cols = c(tidyselect::starts_with("score_"),
             tidyselect::starts_with("responded_")),
    names_to = c(".value", "feature"),
    names_pattern = "(score|responded)_(word|angle|color)"
  ) |>
  dplyr::mutate(
    feature = factor(feature, levels = c("word", "angle", "color"),
                     labels = c("Word", "Orientation", "Colour"))
  )

# 3. v1 alone is the primary analysis sample; see section 4.
v1_trials <- dplyr::filter(trials_by_feature, version == "v1")

# 4. One mean per participant per feature, over answered trials only.
v1_participant_means <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(
    mean_score = mean(score),
    n_answered = dplyr::n(),
    vviq = dplyr::first(vviq),
    .by = c(id, feature)
  )

dplyr::slice_head(trials_by_feature, n = 6) |>
  knitr::kable(digits = 3, caption = "trials_by_feature: first six rows")
```

## 1. Why the task's own scores are not the analysis scores

WM-FTT computes scores in the browser, to display feedback between trials.
Those columns survive in `all_data` under a `feedback_` prefix, alongside
the intermediate dissimilarities they derive from under `live_diff_`.
Neither should be analysed. The reason is not that in-task feedback is
coarse, though it is. It is that one of the three metrics is wrong.

```{r utils-bug, eval = FALSE}
# js/jspsych/utils.js, lines 20-22
for (let i = 0; (i = 0); i--) {
  dp[i][0] = i;
}
```

The loop *condition* is an assignment. `i = 0` evaluates to `0`, which is
falsy, so the body never runs and the first column of the
dynamic-programming matrix is never initialised. It keeps its `fill(0)`
value, which makes deleting any prefix of the target free. The function is
named `levenshteinDistance` and does not compute Levenshtein distance; it
returns something closer to an approximate-substring distance,
systematically smaller and never larger. The paired loop initialising the
first *row* is correct, so only one boundary is broken. That is why exact
matches still score 0, and why the error survived three versions of the
task undetected.

The claim is testable. Below is a faithful R port of the broken function,
dead loop included, compared against the correct distance.

```{r bug-repro}
broken_levenshtein <- function(target, response) {
  n_target <- nchar(target)
  n_response <- nchar(response)
  if (n_target == 0 || n_response == 0) return(0)

  distances <- matrix(0L, n_target + 1L, n_response + 1L)
  distances[1L, seq_len(n_response) + 1L] <- seq_len(n_response)
  # The first column is never initialised: this is the bug.

  target_chars <- strsplit(target, "")[[1]]
  response_chars <- strsplit(response, "")[[1]]
  for (i in seq_len(n_target)) for (j in seq_len(n_response)) {
    distances[i + 1L, j + 1L] <-
      if (target_chars[i] == response_chars[j]) distances[i, j] else
        min(distances[i, j + 1L], distances[i + 1L, j], distances[i, j]) + 1L
  }
  1 - distances[n_target + 1L, n_response + 1L] / max(n_target, n_response)
}

# JavaScript's Math.round() rounds half away from zero; R's round() does not.
js_round_2dp <- function(x) floor(x * 100 + 0.5) / 100

word_reproduction <- all_data |>
  dplyr::filter(expe_phase != "tutorial") |>
  dplyr::transmute(
    stored_value = live_diff_word,
    target = normalise_word(target_word),
    response = normalise_word(response_word),
    both_present = nzchar(target) & nzchar(response),
    from_broken_code = js_round_2dp(
      1 - purrr::map2_dbl(target, response, \(tgt, rsp)
        if (nzchar(tgt) && nzchar(rsp)) broken_levenshtein(tgt, rsp) else 0)
    ),
    from_correct_code = js_round_2dp(dplyr::if_else(
      both_present,
      purrr::map2_int(target, response, \(tgt, rsp) as.integer(utils::adist(tgt, rsp))) /
        pmax(nchar(target), nchar(response)),
      1
    ))
  )

reproduction_summary <- word_reproduction |>
  dplyr::summarise(
    trials = dplyr::n(),
    matched_by_broken_code = sum(from_broken_code == stored_value),
    matched_by_correct_code = sum(from_correct_code == stored_value)
  )

knitr::kable(reproduction_summary,
             caption = "Which implementation reproduces the stored column")
```

The broken port reproduces the stored column for every one of the
`r reproduction_summary$trials` trials. Correct Levenshtein reproduces
`r reproduction_summary$matched_by_correct_code`. A worked example:

```{r bug-example}
tibble::tibble(
  target = "cadeau",
  response = "tout",
  correct_distance = as.integer(utils::adist(target, response)),
  broken_distance = round(
    (1 - broken_levenshtein(target, response)) * nchar(target)),
  stored_live_diff_word = all_data |>
    dplyr::filter(target_word == "cadeau", response_word == "tout") |>
    dplyr::pull(live_diff_word) |>
    dplyr::first()
) |>
  knitr::kable(caption = "One trial, scored three ways")
```

The practical consequence is small: in-task feedback was more generous on
word recall than intended. The methodological one is not. It is why every
score in this package is computed from raw target and response values in R
rather than transformed from the front end's output.

## 2. The three metrics

**Word** uses the edit-distance scoring of @gonthierEasyWayImprove2022:
Damerau-Levenshtein distance, capped at the target's length, subtracted
from that length and divided by it. Bounded on [0, 1] and floored at 0 by
construction.

Decomposing that formula against a naive normalised distance, one design
choice at a time:

```{r decompose}
word_score_variants <- block_trials |>
  dplyr::filter(version == "v1", responded_word) |>
  dplyr::transmute(
    target = normalise_word(target_word),
    response = normalise_word(response_word),
    target_length = nchar(target),
    longer_length = pmax(nchar(target), nchar(response)),
    damerau = purrr::map2_dbl(target, response, damerau_levenshtein),
    levenshtein = purrr::map2_dbl(target, response,
                                  \(tgt, rsp) as.integer(utils::adist(tgt, rsp)))
  ) |>
  dplyr::mutate(
    as_adopted = (target_length - pmin(damerau, target_length)) / target_length,
    with_plain_metric = (target_length - pmin(levenshtein, target_length)) / target_length,
    with_longer_denominator = (target_length - pmin(damerau, target_length)) / longer_length,
    without_the_cap = (target_length - damerau) / target_length
  )

word_formula_decomposition <- tibble::tribble(
  ~design_choice,                                    ~column,
  "Metric: Damerau-Levenshtein vs plain Levenshtein", "with_plain_metric",
  "Denominator: target length vs longer string",      "with_longer_denominator",
  "The cap that floors the score at 0",               "without_the_cap"
) |>
  dplyr::mutate(
    trials_changed = purrr::map_int(column, \(variant)
      sum(abs(word_score_variants$as_adopted - word_score_variants[[variant]]) > 1e-9)),
    of_trials = nrow(word_score_variants)
  ) |>
  dplyr::select(design_choice, trials_changed, of_trials)

knitr::kable(word_formula_decomposition,
             caption = "Trials whose score changes if one choice is reversed")
```

```{r decompose-values, include = FALSE}
changed_by <- function(choice) {
  word_formula_decomposition$trials_changed[
    grepl(choice, word_formula_decomposition$design_choice)]
}
```

The transposition rule that gives the metric its name changes
`r changed_by("Metric")` trials; the denominator changes
`r changed_by("Denominator")`; the cap changes `r changed_by("cap")`, and
without it scores fall as low as
`r fmt(min(word_score_variants$without_the_cap), 2)`. The least famous of
the three decisions does the most work.

**Colour and orientation** both use a cosine of the angular error, with
**different periods**. Colour is a hue wheel: period 360. Orientation is
the tilt of a plain rectangle, which has 180° rotational symmetry. A
rectangle at +80° and one at −80° differ by 160° arithmetically but are
20° apart as orientations, and both look near-horizontal. The response
widget clamps to [−90, 90], exactly one period.

```{r period}
orientation_period_comparison <- block_trials |>
  dplyr::filter(version == "v1") |>
  dplyr::mutate(with_period_360 = dplyr::if_else(
    responded_angle,
    score_angular(target_angle, response_angle, period = 360),
    0
  )) |>
  dplyr::summarise(
    with_period_180 = mean(score_angle),
    with_period_360 = mean(with_period_360),
    .by = participant
  ) |>
  dplyr::mutate(rank_shift = abs(rank(with_period_180) - rank(with_period_360)))

orientation_period_comparison |>
  dplyr::summarise(participants = dplyr::n(),
                   largest_rank_shift = max(rank_shift)) |>
  knitr::kable(caption = "How far participants move between the two periods")
```

Using 360 for orientation would score a perpendicular error, the most
wrong an orientation can be, as 0.5 rather than 0, and no response could
score 0 at all. The choice is structural rather than empirical, but it is
not cosmetic: participant rankings move by up to
`r max(orientation_period_comparison$rank_shift)` places out of
`r nrow(orientation_period_comparison)` between the two conventions.

## 3. What counts as a response

The task encodes non-response with sentinels rather than missing values: an
empty word box, a colour wheel never touched (999, rendered grey), and an
orientation widget left at its starting vertical position (90°).

```{r nonresponse}
non_response_rates <- v1_trials |>
  dplyr::summarise(non_response_rate = 1 - mean(responded), .by = feature)

knitr::kable(non_response_rates, digits = 3,
             caption = "Share of v1 trials with no response, by feature")
```

Between `r as_pct(min(non_response_rates$non_response_rate))` and
`r as_pct(max(non_response_rates$non_response_rate))` of v1 trials are
non-responses. These are scored 0 and flagged, rather than set to `NA`.

`NA` would be the more cautious-looking choice and is the worse one.
Non-response here is unlikely to be missing-at-random, so listwise
deletion would silently drop a non-random subset. Scoring 0 produces a
visible distortion in the distribution instead of an invisible one in the
inference. The `responded_*` flags then let any analysis separate the two
quantities the score column combines: whether a participant reported a
feature at all, and how accurately they did so given that they tried.

```{r circularity}
vviq_association <- v1_trials |>
  dplyr::summarise(
    counting_non_responses_as_zero = mean(score),
    responded_trials_only = mean(score[responded]),
    vviq = dplyr::first(vviq),
    .by = c(id, feature)
  ) |>
  dplyr::summarise(
    dplyr::across(c(counting_non_responses_as_zero, responded_trials_only),
                  \(x) cor(x, vviq, method = "spearman", use = "complete.obs")),
    .by = feature
  )

knitr::kable(vviq_association, digits = 3,
             caption = "Spearman correlation with VVIQ, scored two ways")
```

The gap between those two columns is why the distinction matters:
including non-responses as zeros inflates the apparent association between
score and imagery vividness. Every analysis downstream states which of the
two quantities it means.

## 4. Which sample

v1 is the primary analysis sample. v2 and v3 are described here but do not
enter the inferential analyses.

```{r versions}
block_trials |>
  dplyr::distinct(participant, version, vviq_group_2) |>
  dplyr::count(version, vviq_group_2) |>
  tidyr::pivot_wider(names_from = vviq_group_2, values_from = n,
                     values_fill = 0) |>
  knitr::kable(caption = "Participants per version and imagery group")
```

v1 is the only version with a usable group split. More importantly,
pooling the three actively created a false result: an apparent association
between non-response and imagery vividness dissolves within version,
because v3 is simultaneously aphantasia-heavy and the version with by far
the highest non-response.

```{r pooling-trap}
non_response_per_participant <- trials_by_feature |>
  dplyr::filter(feature != "Word") |>
  dplyr::summarise(non_response_rate = 1 - mean(responded),
                   vviq = dplyr::first(vviq),
                   .by = c(participant, version, feature))

dplyr::bind_rows(
  dplyr::mutate(non_response_per_participant, sample = "All versions pooled"),
  non_response_per_participant |>
    dplyr::filter(version == "v1") |>
    dplyr::mutate(sample = "v1 alone")
) |>
  dplyr::summarise(
    rho = cor(non_response_rate, vviq, method = "spearman",
              use = "complete.obs"),
    .by = c(sample, feature)
  ) |>
  tidyr::pivot_wider(names_from = feature, values_from = rho) |>
  knitr::kable(digits = 3,
               caption = "Non-response against VVIQ, pooled versus within v1")
```

v3 retains one job it uniquely enables. v1 has *fixed* recall order (word,
then orientation, then colour), so feature differences are confounded with
output position. v3 randomised the order and can estimate that effect.

Two costs are worth stating. v1 carries no parity penalty, so the parity
distractor had no scoring consequence there; and v1 has the lowest
non-response of the three versions.

## 5. Distributions

```{r distributions}
v1_trials |>
  dplyr::filter(responded) |>
  ggplot(aes(score)) +
  geom_histogram(bins = 40) +
  facet_wrap(~feature, scales = "free_y") +
  labs(x = "Similarity score, responded trials", y = "Trials")
```

```{r boundaries}
boundary_mass <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(share_at_zero = mean(score == 0),
                   share_at_one = mean(score == 1),
                   .by = feature)

knitr::kable(boundary_mass, digits = 3,
             caption = "Mass at the boundaries, responded trials only")
```

Conditional on responding, orientation and colour are well spread with no
mass at either boundary. Word is not:
`r as_pct(boundary_mass$share_at_one[boundary_mass$feature == "Word"])` of
responded word trials are exact matches.

All the mass at zero in the unfiltered columns is non-response; the mass
at one is genuine ceiling. Since those are different generative processes,
a zero-one-inflated model would be the wrong shape. The performance model
uses a hurdle instead, gating on whether a response occurred before
modelling its accuracy.

## 6. Reliability

Splits are drawn over trials rather than items, because items within a
trial share an encoding episode and there is a within-trial primacy
effect:

```{r primacy}
v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(mean_score = mean(score),
                   .by = c(feature, position_in_trial)) |>
  tidyr::pivot_wider(names_from = position_in_trial, values_from = mean_score,
                     names_prefix = "position_") |>
  dplyr::mutate(primacy_drop = position_1 - position_3) |>
  knitr::kable(digits = 3,
               caption = "Mean score by serial position within the trial")
```

Each trial's three items therefore stay together, and the split is
repeated `r n_splits` times rather than relying on one arbitrary odd/even
partition.

The engagement threshold is derived rather than chosen: a participant
contributes to a feature only if their mean has a standard error of at
most half the between-person SD.

```{r thresholds}
engagement_thresholds <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(participant_mean = mean(score),
                   sd_within_participant = sd(score),
                   .by = c(feature, id)) |>
  dplyr::summarise(
    sd_within_participant = median(sd_within_participant, na.rm = TRUE),
    sd_between_participants = sd(participant_mean),
    .by = feature
  ) |>
  dplyr::mutate(
    trials_needed = ceiling(
      (sd_within_participant / (0.5 * sd_between_participants))^2),
    trials_available = trials_per_feature,
    # Fall back to half the trials where the target is unreachable.
    threshold_used = pmin(trials_needed, 32L)
  )

knitr::kable(engagement_thresholds, digits = 3,
             caption = "Trials needed for a participant mean worth comparing")
```

```{r threshold-values, include = FALSE}
word_row <- dplyr::filter(engagement_thresholds, feature == "Word")
```

**Word's requirement cannot be met.** It asks for
`r word_row$trials_needed` trials of the `r trials_per_feature` that exist,
because its between-person SD among responders
(`r fmt(word_row$sd_between_participants)`) is small relative to
within-person trial noise (`r fmt(word_row$sd_within_participant)`). A
floor of 32 trials is used instead, recorded as a data-sufficiency rule
and not a precision guarantee.

```{r reliability}
eligible_participants <- v1_participant_means |>
  dplyr::left_join(dplyr::select(engagement_thresholds, feature, threshold_used),
                   by = "feature") |>
  dplyr::filter(n_answered >= threshold_used) |>
  dplyr::select(id, feature)

trial_ids <- unique(v1_trials$trial_id)
half_size <- length(trial_ids) %/% 2

# Mean score per participant per feature, over one half of the trials.
mean_scores_over <- function(chosen_trials) {
  v1_trials |>
    dplyr::filter(responded, trial_id %in% chosen_trials) |>
    dplyr::semi_join(eligible_participants, by = c("id", "feature")) |>
    dplyr::summarise(mean_score = mean(score), .by = c(feature, id))
}

correlate_one_split <- function(...) {
  first_half <- sample(trial_ids, half_size)
  dplyr::inner_join(
    mean_scores_over(first_half),
    mean_scores_over(setdiff(trial_ids, first_half)),
    by = c("feature", "id"), suffix = c("_first", "_second")
  ) |>
    dplyr::summarise(r = cor(mean_score_first, mean_score_second), .by = feature)
}

# Corrects a half-length test's correlation up to full-test reliability.
spearman_brown <- function(r) 2 * r / (1 + r)

split_half_reliability <- purrr::map(seq_len(n_splits), correlate_one_split) |>
  purrr::list_rbind() |>
  dplyr::summarise(
    reliability = spearman_brown(median(r)),
    lower_95 = spearman_brown(quantile(r, 0.025)),
    upper_95 = spearman_brown(quantile(r, 0.975)),
    .by = feature
  ) |>
  dplyr::left_join(dplyr::count(eligible_participants, feature,
                                name = "participants"),
                   by = "feature") |>
  dplyr::select(feature, participants, reliability, lower_95, upper_95)

knitr::kable(split_half_reliability, digits = 3,
             caption = "Split-half reliability, Spearman-Brown corrected")
```

```{r reliability-values, include = FALSE}
reliability_of <- function(which_feature) {
  split_half_reliability$reliability[
    split_half_reliability$feature == which_feature]
}
```

Orientation (`r fmt(reliability_of("Orientation"), 2)`) and colour
(`r fmt(reliability_of("Colour"), 2)`) clear the conventional 0.80 mark
for individual-level interpretation. Word
(`r fmt(reliability_of("Word"), 2)`) does not clear the 0.70 floor for
group-level use. That is the same fact as the threshold failure, seen from
the other side.

**Word does not function as an individual-differences measure in this
sample.** The metric is not at fault; Gonthier's formula works exactly as
designed. Single-word recall at this exposure is simply too easy to
discriminate between people. No downstream model should make an
individual-differences claim about word, whatever its coefficients say.

```{r composition-stability}
ids_clearing_all_features <- eligible_participants |>
  dplyr::count(id) |>
  dplyr::filter(n == 3) |>
  dplyr::pull(id)

# Isometric log-ratio coordinates, computed from one half of the trials.
ilr_coordinates_over <- function(chosen_trials) {
  v1_trials |>
    dplyr::filter(responded, id %in% ids_clearing_all_features,
                  trial_id %in% chosen_trials) |>
    dplyr::summarise(mean_score = mean(score), .by = c(id, feature)) |>
    tidyr::pivot_wider(names_from = feature, values_from = mean_score) |>
    tidyr::drop_na() |>
    dplyr::mutate(
      total = Word + Orientation + Colour,
      ilr1 = sqrt(2 / 3) * log((Word / total) /
                                 sqrt((Colour / total) * (Orientation / total))),
      ilr2 = sqrt(1 / 2) * log((Colour / total) / (Orientation / total))
    ) |>
    dplyr::select(id, ilr1, ilr2)
}

composition_stability <- purrr::map(seq_len(n_splits), \(iteration) {
  first_half <- sample(trial_ids, half_size)
  dplyr::inner_join(
    ilr_coordinates_over(first_half),
    ilr_coordinates_over(setdiff(trial_ids, first_half)),
    by = "id", suffix = c("_first", "_second")
  ) |>
    dplyr::summarise(ilr1 = cor(ilr1_first, ilr1_second),
                     ilr2 = cor(ilr2_first, ilr2_second))
}) |>
  purrr::list_rbind() |>
  dplyr::summarise(dplyr::across(
    c(ilr1, ilr2), \(x) spearman_brown(median(x, na.rm = TRUE)))) |>
  tidyr::pivot_longer(everything(), names_to = "coordinate",
                      values_to = "stability") |>
  dplyr::mutate(contrast = c("Word vs (colour and orientation)",
                             "Colour vs orientation"))

knitr::kable(composition_stability, digits = 3,
             caption = "Split-half stability of the compositional profile")
```

The compositional profile fares better than its parts would suggest, on
the `r length(ids_clearing_all_features)` participants clearing all three
thresholds. A ratio can carry information a level does not, though the
first coordinate rests partly on word, and its interpretation should not
lean on word as a well-measured quantity.

## 7. What this supports

Orientation and colour scores support individual-differences analysis.
Word scores do not, and should be reported with the measurement caveat
attached. The compositional profile is stable enough to interpret, with
the same caveat on its verbal coordinate.

Self-reported strategy and behavioural allocation are not the same
construct:

```{r self-report}
self_reported_priority <- block_trials |>
  dplyr::filter(version == "v1") |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, strategy_items) |>
  tidyr::unnest(strategy_items) |>
  dplyr::select(id, tidyselect::contains("scoring_strat")) |>
  tidyr::pivot_longer(-id, values_to = "named_feature",
                      values_transform = as.character) |>
  dplyr::filter(named_feature %in% c("words", "colours", "orientations")) |>
  dplyr::summarise(
    priority = dplyr::if_else(dplyr::n() > 1, "multiple",
                              dplyr::first(named_feature)),
    .by = id
  )

v1_compositions <- v1_participant_means |>
  dplyr::filter(id %in% ids_clearing_all_features) |>
  dplyr::mutate(proportion = mean_score / sum(mean_score), .by = id) |>
  dplyr::select(id, feature, proportion) |>
  tidyr::pivot_wider(names_from = feature, values_from = proportion)

v1_compositions |>
  dplyr::left_join(self_reported_priority, by = "id") |>
  tidyr::replace_na(list(priority = "none/other")) |>
  dplyr::summarise(dplyr::across(c(Word, Orientation, Colour), mean),
                   participants = dplyr::n(),
                   .by = priority) |>
  knitr::kable(digits = 3,
               caption = "Behavioural composition by self-reported priority")
```

Behavioural compositions barely differ between participants who named
different priorities. This is a dissociation to report, not a validity
failure. But "preference" language should not be used as though the two
measures were interchangeable.

## 8. For a future version of the task

The task was designed to force a three-way trade-off. It achieves a
two-way one, because word is cheap enough that participants can max it and
then divide effort between colour and orientation:

```{r tradeoff}
v1_scores_wide <- v1_participant_means |>
  dplyr::select(id, feature, mean_score) |>
  tidyr::pivot_wider(names_from = feature, values_from = mean_score) |>
  tidyr::drop_na()

# Residualise each feature on overall level, then correlate the residuals.
v1_scores_wide |>
  dplyr::mutate(overall_level = (Word + Orientation + Colour) / 3) |>
  dplyr::mutate(dplyr::across(
    c(Word, Orientation, Colour),
    \(feature_score) residuals(lm(feature_score ~ overall_level)))) |>
  dplyr::select(Word, Orientation, Colour) |>
  cor(method = "spearman") |>
  knitr::kable(digits = 3,
               caption = "Feature correlations after removing overall level")
```

Controlling overall level, orientation and colour trade off against each
other more strongly than either does against word. (Residualising three
variables on their own mean induces negative correlation of roughly −0.5
by construction, so read these relative to that baseline rather than as
raw evidence of a trade-off.)

Three changes would address what this analysis could not:

- **Make word harder**, with longer words, more items per trial, or
  shorter exposure, so the verbal feature enters the trade-off and
  discriminates.
- **Distinguish abstention from a default**, via an explicit "I don't know"
  control, separate from an untouched widget, and a confidence rating.
  Abstention is currently far faster than responding, which rules out
  effortful-but-failed search but cannot separate "nothing to report" from
  "not worth the effort".
- **Vary the points per feature**, since strategic abstention should
  respond to incentive and representational absence should not.

## References
