---
title: "Scoring the WM-FTT: what the numbers are, and what they support"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Scoring the WM-FTT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_minimal(base_size = 16))
set.seed(20260820)

n_splits <- 1000
trials_per_feature <- 63L

as_pct <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

The Working Memory Feature Trade-off Task asks participants to hold three
features of each stimulus (a word, the orientation of the rectangle
containing it, and its colour) and to report all three at recall. Scoring
it is less obvious than it looks. This page documents how the scores in
`all_data` are computed, what the distributions look like, how reliable
they are, and which claims they will and will not support.

Every number below is computed from `all_data` when this page is built,
not transcribed from an analysis run elsewhere.

## The data this page works from

`all_data` is stimulus-level: one row per presented stimulus, with all
three recalled features side by side in separate columns. Almost every
question below is about *features*, so the first step is to reshape to one
row per stimulus per feature. Three frames are built once here and reused
throughout.

```{r data}
# 1. Experimental-block rows only. Tutorial rows carry placeholder values,
#    and training rows precede the blocks by design.
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase)) |>
  dplyr::mutate(
    participant = paste(id, version, sep = "_"),  # one id spans two versions
    trial_id = paste(expe_phase, trial_number, sep = "_"),
    position_in_trial = (item_number - 1L) %% 3L + 1L
  )

# 2. One row per stimulus per feature: score, whether it was answered.
trials_by_feature <- block_trials |>
  dplyr::select(
    id, participant, version, trial_id, position_in_trial,
    vviq = vviq_total_score,
    score_word, score_angle, score_color,
    responded_word, responded_angle, responded_color
  ) |>
  tidyr::pivot_longer(
    cols = c(tidyselect::starts_with("score_"),
             tidyselect::starts_with("responded_")),
    names_to = c(".value", "feature"),
    names_pattern = "(score|responded)_(word|angle|color)"
  ) |>
  dplyr::mutate(
    feature = factor(feature, levels = c("word", "angle", "color"),
                     labels = c("Word", "Orientation", "Colour"))
  )

# 3. v1 alone is the primary analysis sample; see section 4.
v1_trials <- dplyr::filter(trials_by_feature, version == "v1")

# 4. One mean per participant per feature, over answered items only.
v1_participant_means <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(
    mean_score = mean(score),
    n_answered = dplyr::n(),
    vviq = dplyr::first(vviq),
    .by = c(id, feature)
  )

dplyr::slice_head(trials_by_feature, n = 6) |>
  knitr::kable(digits = 3, caption = "trials_by_feature: first six rows")
```

## 1. Why the task's own scores are not the analysis scores

WM-FTT computes scores in the browser, to display feedback between trials.
Those columns survive in `all_data` under a `feedback_` prefix, alongside
the intermediate dissimilarities they derive from under `live_diff_`.
Neither should be analysed. The reason is not that in-task feedback is
coarse, though it is. It is that one of the three metrics is wrong.

```{r utils-bug, eval = FALSE}
# js/jspsych/utils.js, lines 20-22
for (let i = 0; (i = 0); i--) {
  dp[i][0] = i;
}
```

The loop *condition* is an assignment. `i = 0` evaluates to `0`, which is
falsy, so the body never runs and the first column of the
dynamic-programming matrix is never initialised. It keeps its `fill(0)`
value, which makes deleting any prefix of the target free. The function is
named `levenshteinDistance` and does not compute Levenshtein distance; it
returns something closer to an approximate-substring distance,
systematically smaller and never larger. The paired loop initialising the
first *row* is correct, so only one boundary is broken. That is why exact
matches still score 0, and why the error survived three versions of the
task undetected.

The claim is testable. Below is a faithful R port of the broken function,
dead loop included, compared against the correct distance.

```{r bug-repro}
broken_levenshtein <- function(target, response) {
  n_target <- nchar(target)
  n_response <- nchar(response)
  if (n_target == 0 || n_response == 0) return(0)

  distances <- matrix(0L, n_target + 1L, n_response + 1L)
  distances[1L, seq_len(n_response) + 1L] <- seq_len(n_response)
  # The first column is never initialised: this is the bug.

  target_chars <- strsplit(target, "")[[1]]
  response_chars <- strsplit(response, "")[[1]]
  for (i in seq_len(n_target)) for (j in seq_len(n_response)) {
    distances[i + 1L, j + 1L] <-
      if (target_chars[i] == response_chars[j]) distances[i, j] else
        min(distances[i, j + 1L], distances[i + 1L, j], distances[i, j]) + 1L
  }
  1 - distances[n_target + 1L, n_response + 1L] / max(n_target, n_response)
}

# JavaScript's Math.round() rounds half away from zero; R's round() does not.
js_round_2dp <- function(x) floor(x * 100 + 0.5) / 100

word_reproduction <- all_data |>
  dplyr::filter(expe_phase != "tutorial") |>
  dplyr::transmute(
    stored_value = live_diff_word,
    target = normalise_word(target_word),
    response = normalise_word(response_word),
    both_present = nzchar(target) & nzchar(response),
    from_broken_code = js_round_2dp(
      1 - purrr::map2_dbl(target, response, \(tgt, rsp)
        if (nzchar(tgt) && nzchar(rsp)) broken_levenshtein(tgt, rsp) else 0)
    ),
    from_correct_code = js_round_2dp(dplyr::if_else(
      both_present,
      purrr::map2_int(target, response, \(tgt, rsp) as.integer(utils::adist(tgt, rsp))) /
        pmax(nchar(target), nchar(response)),
      1
    ))
  )

reproduction_summary <- word_reproduction |>
  dplyr::summarise(
    items = dplyr::n(),
    matched_by_broken_code = sum(from_broken_code == stored_value),
    matched_by_correct_code = sum(from_correct_code == stored_value)
  )

knitr::kable(reproduction_summary,
             caption = "Which implementation reproduces the stored column")
```

The broken port reproduces the stored column for every one of the
`r reproduction_summary$items` items. Correct Levenshtein reproduces
`r reproduction_summary$matched_by_correct_code`. A worked example:

```{r bug-example}
tibble::tibble(
  target = "cadeau",
  response = "tout",
  correct_distance = as.integer(utils::adist(target, response)),
  broken_distance = round(
    (1 - broken_levenshtein(target, response)) * nchar(target)),
  stored_live_diff_word = all_data |>
    dplyr::filter(target_word == "cadeau", response_word == "tout") |>
    dplyr::pull(live_diff_word) |>
    dplyr::first()
) |>
  knitr::kable(caption = "One item, scored three ways")
```

The practical consequence is small: in-task feedback was more generous on
word recall than intended. The methodological one is not. It is why every
score in this package is computed from raw target and response values in R
rather than transformed from the front end's output.

## 2. The three metrics

**Word** uses the edit-distance scoring of @gonthierEasyWayImprove2022:
Damerau-Levenshtein distance, capped at the target's length, subtracted
from that length and divided by it. Bounded on [0, 1] and floored at 0 by
construction.

Decomposing that formula against a naive normalised distance, one design
choice at a time:

```{r decompose}
word_score_variants <- block_trials |>
  dplyr::filter(version == "v1", responded_word) |>
  dplyr::transmute(
    target = normalise_word(target_word),
    response = normalise_word(response_word),
    target_length = nchar(target),
    longer_length = pmax(nchar(target), nchar(response)),
    damerau = purrr::map2_dbl(target, response, damerau_levenshtein),
    levenshtein = purrr::map2_dbl(target, response,
                                  \(tgt, rsp) as.integer(utils::adist(tgt, rsp)))
  ) |>
  dplyr::mutate(
    as_adopted = (target_length - pmin(damerau, target_length)) / target_length,
    with_plain_metric = (target_length - pmin(levenshtein, target_length)) / target_length,
    with_longer_denominator = (target_length - pmin(damerau, target_length)) / longer_length,
    without_the_cap = (target_length - damerau) / target_length
  )

word_formula_decomposition <- tibble::tribble(
  ~design_choice,                                    ~column,
  "Metric: Damerau-Levenshtein vs plain Levenshtein", "with_plain_metric",
  "Denominator: target length vs longer string",      "with_longer_denominator",
  "The cap that floors the score at 0",               "without_the_cap"
) |>
  dplyr::mutate(
    items_changed = purrr::map_int(column, \(variant)
      sum(abs(word_score_variants$as_adopted - word_score_variants[[variant]]) > 1e-9)),
    of_items = nrow(word_score_variants)
  ) |>
  dplyr::select(design_choice, items_changed, of_items)

knitr::kable(word_formula_decomposition,
             caption = "Items whose score changes if one choice is reversed")
```

```{r decompose-values, include = FALSE}
changed_by <- function(choice) {
  word_formula_decomposition$items_changed[
    grepl(choice, word_formula_decomposition$design_choice)]
}
```

The transposition rule that gives the metric its name changes
`r changed_by("Metric")` items; the denominator changes
`r changed_by("Denominator")`; the cap changes `r changed_by("cap")`, and
without it scores fall as low as
`r fmt(min(word_score_variants$without_the_cap), 2)`. The least famous of
the three decisions does the most work.

**Colour and orientation** both use a cosine of the angular error, with
**different periods**. Colour is a hue wheel: period 360. Orientation is
the tilt of a plain rectangle, which has 180° rotational symmetry. A
rectangle at +80° and one at −80° differ by 160° arithmetically but are
20° apart as orientations, and both look near-horizontal. The response
widget clamps to [−90, 90], exactly one period.

```{r period}
orientation_period_comparison <- block_trials |>
  dplyr::filter(version == "v1") |>
  dplyr::mutate(with_period_360 = dplyr::if_else(
    responded_angle,
    score_angular(target_angle, response_angle, period = 360),
    0
  )) |>
  dplyr::summarise(
    with_period_180 = mean(score_angle),
    with_period_360 = mean(with_period_360),
    .by = participant
  ) |>
  dplyr::mutate(rank_shift = abs(rank(with_period_180) - rank(with_period_360)))

orientation_period_comparison |>
  dplyr::summarise(participants = dplyr::n(),
                   largest_rank_shift = max(rank_shift)) |>
  knitr::kable(caption = "How far participants move between the two periods")
```

Using 360 for orientation would score a perpendicular error, the most
wrong an orientation can be, as 0.5 rather than 0, and no response could
score 0 at all. The choice is structural rather than empirical, but it is
not cosmetic: participant rankings move by up to
`r max(orientation_period_comparison$rank_shift)` places out of
`r nrow(orientation_period_comparison)` between the two conventions.

## 3. What counts as a response

The task encodes non-response with sentinels rather than missing values: an
empty word box, a colour wheel never touched (999, rendered grey), and an
orientation widget left at its starting vertical position (90°).

```{r nonresponse}
non_response_rates <- v1_trials |>
  dplyr::summarise(non_response_rate = 1 - mean(responded), .by = feature)

knitr::kable(non_response_rates, digits = 3,
             caption = "Share of v1 items with no response, by feature")
```

Between `r as_pct(min(non_response_rates$non_response_rate))` and
`r as_pct(max(non_response_rates$non_response_rate))` of v1 items are
non-responses. These are scored 0 and flagged, rather than set to `NA`.

`NA` would be the more cautious-looking choice and is the worse one.
Non-response here is unlikely to be missing-at-random, so listwise
deletion would silently drop a non-random subset. Scoring 0 produces a
visible distortion in the distribution instead of an invisible one in the
inference. The `responded_*` flags then let any analysis separate the two
quantities the score column combines: whether a participant reported a
feature at all, and how accurately they did so given that they tried.

```{r circularity}
vviq_association <- v1_trials |>
  dplyr::summarise(
    counting_non_responses_as_zero = mean(score),
    responded_items_only = mean(score[responded]),
    vviq = dplyr::first(vviq),
    .by = c(id, feature)
  ) |>
  dplyr::summarise(
    dplyr::across(c(counting_non_responses_as_zero, responded_items_only),
                  \(x) cor(x, vviq, method = "spearman", use = "complete.obs")),
    .by = feature
  )

knitr::kable(vviq_association, digits = 3,
             caption = "Spearman correlation with VVIQ, scored two ways")
```

The gap between those two columns is why the distinction matters:
including non-responses as zeros inflates the apparent association between
score and imagery vividness. Every analysis downstream states which of the
two quantities it means.

## 4. Which sample

v1 is the primary analysis sample. v2 and v3 are described but do not enter
the inferential analyses, and the [version scope](version-scope.html) page
sets out the procedure that resolved to that, including the case where
pooling the three versions produced an association that stratification
removed.

Everything below is v1 unless stated otherwise.

## 5. Distributions

```{r distributions}
v1_trials |>
  dplyr::filter(responded) |>
  ggplot(aes(score)) +
  geom_histogram(bins = 40) +
  facet_wrap(~feature, scales = "free_y") +
  labs(x = "Similarity score, responded items", y = "Items")
```

```{r boundaries}
boundary_mass <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(share_at_zero = mean(score == 0),
                   share_at_one = mean(score == 1),
                   .by = feature)

knitr::kable(boundary_mass, digits = 3,
             caption = "Mass at the boundaries, responded items only")
```

Conditional on responding, orientation is well spread with no mass at
either boundary at all. Word is not:
`r as_pct(boundary_mass$share_at_one[boundary_mass$feature == "Word"])` of
responded word items are exact matches, and
`r as_pct(boundary_mass$share_at_zero[boundary_mass$feature == "Word"])` are
exact misses. Colour has a trace of mass at one and none at zero.

**Those three patterns need three different treatments**, and each choice
follows from what the boundary means rather than from convenience.

- **Word's boundaries are real.** A recalled word either matches the target
  or it does not, so the pile at one is a genuine point mass, and the
  zeros here are wrong answers rather than absent ones: this table is
  already conditional on responding. Both earn an inflation component, so
  word is modelled with a zero-one-inflated Beta.
- **Orientation needs nothing.** Plain Beta.
- **Colour's mass at one is an artifact.** Colour score is cosine
  similarity on a continuous wheel, so an exact 1 is an error of exactly
  zero degrees, which is pixel resolution rather than a behaviour.
  Modelling it as an inflation component would estimate a process
  parameter for a rounding effect on a couple of dozen items, so it is
  handled with a Smithson-Verkuilen squeeze instead
  (`squeeze_boundaries()`), which moves every value by about one
  ten-thousandth.

Separately from all of that, the mass at zero in the **unfiltered** columns
is non-response, which is a different quantity again and is modelled as
such: see [task engagement](engagement.html).

## 6. Reliability, and what these scores support

Both questions have moved to their own page: [task
validity](task-validity.html). It covers the trial-level split-half
procedure, the derived engagement thresholds, the finding that word does
not function as an individual-differences measure in this sample, and the
stability of the compositional coordinates.

The short version: orientation and colour clear the conventional marks for
individual-level interpretation, word does not, and no downstream model
should make an individual-differences claim about word whatever its
coefficients say.

## 7. For a future version of the task

The design goal was a three-way trade-off. Correlations between the three
features, after removing each participant's overall level, once looked like
a **two-way** one: orientation and colour trading off against each other
while word sat out at ceiling. That reading is withdrawn. It was six
participants who answered very few orientation items, and the [what was
tried and withdrawn](lessons.html) page owns that withdrawal and shows both
tables side by side.

What survives is weaker and still worth acting on: word is far too easy,
with `r as_pct(boundary_mass$share_at_one[boundary_mass$feature == "Word"])`
of responded items at ceiling and a reliability that does not support
individual-differences use, so it cannot be carrying much of a trade-off
whatever the correlations say.

Three changes would address what this analysis could not:

- **Make word harder**, with longer words, more items per trial, or
  shorter exposure, so the verbal feature enters the trade-off and
  discriminates.
- **Distinguish abstention from a default**, via an explicit "I don't know"
  control, separate from an untouched widget, and a confidence rating.
  Abstention is currently far faster than responding, which rules out
  effortful-but-failed search but cannot separate "nothing to report" from
  "not worth the effort".
- **Vary the points per feature**, since strategic abstention should
  respond to incentive and representational absence should not.

## References

<div id="refs"></div>

---

**Continuing through the Extended Online Report:** this page follows [task design](task-design.html). To keep reading in order, continue to [task engagement](engagement.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
