---
title: "Task validity: what the WM-FTT measures well, and what it does not"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Task validity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
set.seed(20260820)

as_pct <- function(x, digits = 1) sprintf(paste0("%.", digits, "f%%"), 100 * x)
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

```{r constants}
# The three constants the chunks below use, set here rather than hidden in
# the setup chunk, so "repeated 1000 times" and "of the 63 that exist" can
# be tied to something.
n_splits <- 1000
trials_per_feature <- 63L
feature_labels <- c(word = "Word", angle = "Orientation", color = "Colour")
```

The [scoring](scoring.html) page defines the three recall scores. This page
asks the prior question every result on this site depends on: **are they
reliable enough to carry an individual-differences claim?**

For two of the three features the answer is yes. For the third it is no,
and that is a fact about the task rather than about the metric. The
[questionnaire psychometrics](psychometrics.html) page asks the same
question of the instruments.

Everything here is v1 only, per the [version scope](version-scope.html)
page.

## The data this page works from

The same frames the scoring page builds, reproduced so this page stands
alone.

```{r data}
# 1. Experimental-block rows only. Tutorial rows carry placeholder values,
#    and training rows precede the blocks by design.
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase)) |>
  dplyr::mutate(
    participant = paste(id, version, sep = "_"),  # one id spans two versions
    trial_id = paste(expe_phase, trial_number, sep = "_"),
    position_in_trial = (item_number - 1L) %% 3L + 1L
  )

# 2. One row per stimulus per feature: score, whether it was answered.
trials_by_feature <- block_trials |>
  dplyr::select(
    id, participant, version, trial_id, position_in_trial,
    vviq = vviq_total_score,
    score_word, score_angle, score_color,
    responded_word, responded_angle, responded_color
  ) |>
  tidyr::pivot_longer(
    cols = c(tidyselect::starts_with("score_"),
             tidyselect::starts_with("responded_")),
    names_to = c(".value", "feature"),
    names_pattern = "(score|responded)_(word|angle|color)"
  ) |>
  dplyr::mutate(
    feature = factor(feature, levels = c("word", "angle", "color"),
                     labels = c("Word", "Orientation", "Colour"))
  )

# 3. v1 alone is the primary analysis sample; see section 4.
v1_trials <- dplyr::filter(trials_by_feature, version == "v1")

# 4. One mean per participant per feature, over answered items only.
v1_participant_means <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(
    mean_score = mean(score),
    n_answered = dplyr::n(),
    vviq = dplyr::first(vviq),
    .by = c(id, feature)
  )

dplyr::slice_head(trials_by_feature, n = 6) |>
  knitr::kable(digits = 3, caption = "trials_by_feature: first six rows")
```

## 1. Reliability

Splits are drawn over trials rather than items, because items within a
trial share an encoding episode and there is a within-trial primacy
effect:

```{r primacy}
v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(mean_score = mean(score),
                   .by = c(feature, position_in_trial)) |>
  tidyr::pivot_wider(names_from = position_in_trial, values_from = mean_score,
                     names_prefix = "position_") |>
  dplyr::mutate(primacy_drop = position_1 - position_3) |>
  knitr::kable(digits = 3,
               caption = "Mean score by serial position within the trial")
```

Each trial's three items therefore stay together, and the split is
repeated `r n_splits` times rather than relying on one arbitrary odd/even
partition.

The engagement threshold is derived rather than chosen: a participant
contributes to a feature only if their mean has a standard error of at
most half the between-person SD.

```{r thresholds}
engagement_thresholds <- v1_trials |>
  dplyr::filter(responded) |>
  dplyr::summarise(participant_mean = mean(score),
                   sd_within_participant = sd(score),
                   .by = c(feature, id)) |>
  dplyr::summarise(
    sd_within_participant = median(sd_within_participant, na.rm = TRUE),
    sd_between_participants = sd(participant_mean),
    .by = feature
  ) |>
  dplyr::mutate(
    items_needed = ceiling(
      (sd_within_participant / (0.5 * sd_between_participants))^2),
    items_available = trials_per_feature,
    # Fall back to half the items where the target is unreachable.
    threshold_used = pmin(items_needed, 32L)
  )

knitr::kable(engagement_thresholds, digits = 3,
             caption = "Items needed for a participant mean worth comparing")
```

```{r threshold-values, include = FALSE}
word_row <- dplyr::filter(engagement_thresholds, feature == "Word")
```

**Word's requirement cannot be met.** It asks for
`r word_row$items_needed` items of the `r trials_per_feature` that exist,
because its between-person SD among responders
(`r fmt(word_row$sd_between_participants)`) is small relative to
within-person item noise (`r fmt(word_row$sd_within_participant)`). A
floor of 32 items is used instead, recorded as a data-sufficiency rule
and not a precision guarantee. The `pmin()` above applies it as a **cap on
the requirement**, not as a minimum: where the precision target asks for
more items than exist, 32 is substituted.

One thing about what happens downstream. The derivation above is recomputed
live on this page, but the filter every modelling page applies is
`engaged_ids()`, which reads the **hard-coded** `wm_thresholds()`:

```{r hard-coded-thresholds}
wm_thresholds()
```

They agree today. They are hard-coded deliberately, because the rule was
fixed before any contact with VVIQ and must not drift with the sample it is
applied to — so if this page's recomputation ever diverges from that vector,
the vector is what the analyses used, and the divergence is the finding.

```{r reliability}
eligible_participants <- v1_participant_means |>
  dplyr::left_join(dplyr::select(engagement_thresholds, feature, threshold_used),
                   by = "feature") |>
  dplyr::filter(n_answered >= threshold_used) |>
  dplyr::select(id, feature)

# split_half_reliability() and spearman_brown() are package functions, so
# this page and inst/scripts/06a-reliability.R compute the same quantity
# with the same code rather than two implementations of it.
v1_block <- block_trials |>
  dplyr::filter(version == "v1") |>
  dplyr::rename(trial_uid = trial_id)

# Splits are drawn over trials, so these are needed here and again by the
# compositional stability check below.
trial_ids <- unique(v1_block$trial_uid)
half_size <- length(trial_ids) %/% 2

split_half_draws <- lapply(
  c(word = "word", angle = "angle", color = "color"),
  \(feature) spearman_brown(
    split_half_reliability(v1_block, feature, n_splits = n_splits)
  )
)

reliability_table <-
  tibble::tibble(
    feature = feature_labels[names(split_half_draws)],
    reliability = vapply(split_half_draws, stats::median, numeric(1),
                         na.rm = TRUE),
    lower_95 = vapply(split_half_draws, stats::quantile, numeric(1),
                      probs = 0.025, na.rm = TRUE),
    upper_95 = vapply(split_half_draws, stats::quantile, numeric(1),
                      probs = 0.975, na.rm = TRUE)
  ) |>
  dplyr::left_join(dplyr::count(eligible_participants, feature,
                                name = "participants"),
                   by = "feature") |>
  dplyr::select(feature, participants, reliability, lower_95, upper_95)

knitr::kable(reliability_table, digits = 3,
             caption = "Split-half reliability, Spearman-Brown corrected")
```

```{r reliability-values, include = FALSE}
reliability_of <- function(which_feature) {
  reliability_table$reliability[reliability_table$feature == which_feature]
}
```

```{r reliability-figure, fig.alt="Split-half reliability distributions for the three recall features."}
plot_split_half(split_half_draws, base_size = 16)
```

Orientation (`r fmt(reliability_of("Orientation"), 2)`) and colour
(`r fmt(reliability_of("Colour"), 2)`) clear the conventional 0.80 mark
for individual-level interpretation. Word
(`r fmt(reliability_of("Word"), 2)`) does not clear the 0.70 floor for
group-level use. That is the same fact as the threshold failure, seen from
the other side.

**Word does not function as an individual-differences measure in this
sample.** The metric is not at fault; Gonthier's formula works exactly as
designed. Single-word recall at this exposure is simply too easy to
discriminate between people. No downstream model should make an
individual-differences claim about word, whatever its coefficients say.

```{r composition-stability}
ids_clearing_all_features <- eligible_participants |>
  dplyr::count(id) |>
  dplyr::filter(n == 3) |>
  dplyr::pull(id)

# Isometric log-ratio coordinates from one half of the trials, via the
# package functions rather than a local copy of the transform. The
# partition is word vs (colour and orientation), which is the one the
# modelling pages use, so this stability estimate applies to those
# coordinates and not to some other rotation of them.
ilr_coordinates_over <- function(chosen_trials) {
  parts <- v1_block |>
    dplyr::filter(id %in% ids_clearing_all_features,
                  trial_uid %in% chosen_trials) |>
    compose_features(id)
  parts <- parts[stats::complete.cases(parts), ]

  dplyr::bind_cols(
    dplyr::select(parts, id),
    ilr_coords(dplyr::select(parts, tidyselect::starts_with("part_")))
  )
}

composition_stability <- purrr::map(seq_len(n_splits), \(iteration) {
  first_half <- sample(trial_ids, half_size)
  dplyr::inner_join(
    ilr_coordinates_over(first_half),
    ilr_coordinates_over(setdiff(trial_ids, first_half)),
    by = "id", suffix = c("_first", "_second")
  ) |>
    dplyr::summarise(ilr1 = cor(ilr1_first, ilr1_second),
                     ilr2 = cor(ilr2_first, ilr2_second))
}) |>
  purrr::list_rbind() |>
  dplyr::summarise(dplyr::across(
    c(ilr1, ilr2), \(x) spearman_brown(median(x, na.rm = TRUE)))) |>
  tidyr::pivot_longer(everything(), names_to = "coordinate",
                      values_to = "stability") |>
  dplyr::mutate(contrast = c("Word vs (colour and orientation)",
                             "Colour vs orientation"))

knitr::kable(composition_stability, digits = 3,
             caption = "Split-half stability of the compositional profile")
```

The compositional profile fares better than its parts would suggest, on
the `r length(ids_clearing_all_features)` participants clearing all three
thresholds. A ratio can carry information a level does not, though the
first coordinate rests partly on word, and its interpretation should not
lean on word as a well-measured quantity.

## 2. What this supports

Orientation and colour scores support individual-differences analysis.
Word scores do not, and should be reported with the measurement caveat
attached. The compositional profile is stable enough to interpret, with
the same caveat on its verbal coordinate.

Self-reported strategy and behavioural allocation are not the same
construct:

```{r self-report}
self_reported_priority <- block_trials |>
  dplyr::filter(version == "v1") |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(id, strategy_items) |>
  tidyr::unnest(strategy_items) |>
  dplyr::select(id, tidyselect::contains("scoring_strat")) |>
  tidyr::pivot_longer(-id, values_to = "named_feature",
                      values_transform = as.character) |>
  dplyr::filter(named_feature %in% c("words", "colours", "orientations")) |>
  dplyr::summarise(
    priority = dplyr::if_else(dplyr::n() > 1, "multiple",
                              dplyr::first(named_feature)),
    .by = id
  )

v1_compositions <- v1_participant_means |>
  dplyr::filter(id %in% ids_clearing_all_features) |>
  dplyr::mutate(proportion = mean_score / sum(mean_score), .by = id) |>
  dplyr::select(id, feature, proportion) |>
  tidyr::pivot_wider(names_from = feature, values_from = proportion)

v1_compositions |>
  dplyr::left_join(self_reported_priority, by = "id") |>
  tidyr::replace_na(list(priority = "none/other")) |>
  dplyr::summarise(dplyr::across(c(Word, Orientation, Colour), mean),
                   participants = dplyr::n(),
                   .by = priority) |>
  knitr::kable(digits = 3,
               caption = "Behavioural composition by self-reported priority")
```

Behavioural compositions barely differ between participants who named
different priorities. This table is the evidence behind [task
design](task-design.html) §3's claim that the report and the behaviour are
not interchangeable per person. This is a dissociation to report, not a validity
failure. But "preference" language should not be used as though the two
measures were interchangeable.

---

**Continuing through the Extended Online Report:** this page follows [versions](version-scope.html). To keep reading in order, continue to [questionnaire psychometrics](psychometrics.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
