---
title: "Versions: which one the analyses use, and why"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Version scope}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5,
  warning = FALSE, 
  message = FALSE,
  error = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

The WM-FTT ran in three versions, and the [task design](task-design.html) page
explains why each replaced the last; the [participants](participants.html) page
describes who took part in each. This page answers a narrower question with a
consequence for everything after it: **which versions enter the inferential
analyses.**

As of 2026-09-03, the answer is **v1 only**. Every modelling page on this site
inherits that restriction, and that rule may change (e.g., if v3 ever recruits a
bigger sample) so the reasoning deserves its own page that can be checked from
anywhere. All versions are kept in `all_data`, described in the package, and
available to anyone. The pipeline built on v1 is a template to be re-applied
when the later versions have the sample to support it.

```{r data}
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase))

by_participant <- block_trials |>
  dplyr::summarise(
    imagery_group = dplyr::first(vviq_group_2),
    vviq = dplyr::first(vviq_total_score),
    non_response_word = 1 - mean(responded_word),
    non_response_orientation = 1 - mean(responded_angle),
    non_response_colour = 1 - mean(responded_color),
    .by = c(id, version)
  )
```

## 1. The versions are not interchangeable samples {#ver-samples}

Pooling would be reasonable if the three versions were the same task run three
times. They are not: each changed the task in a way that could plausibly change
behaviour, and they recruited different populations.

```{r group-balance}
by_participant |>
  dplyr::mutate(imagery_group = dplyr::coalesce(as.character(imagery_group),
                                                "no VVIQ")) |>
  dplyr::count(version, imagery_group) |>
  tidyr::pivot_wider(names_from = imagery_group, values_from = n,
                     values_fill = 0L) |>
  knitr::kable(caption = "Participants by version and imagery group")
```

**Group balance is the decisive number.** v1 is 31 aphantasic to 55 typical
imagers. v2 is 8 to 1, and v3 is 17 to 4 (as of 2026-09-03). In v2 and v3 a
group comparison is not merely underpowered: with one and four typical imagers
respectively, it is structurally impossible. No amount of modelling recovers a
contrast that the recruitment never sampled.

The later versions are also aphantasia-heavy *because* no recruitment in the
general population had been conducted yet, only targeted communication in
aphantasia communities.

## 2. Pooling can create spurious results {#ver-pooling}

Non-response differs sharply by version.

```{r nonresponse-by-version}
by_participant |>
  dplyr::summarise(
    dplyr::across(tidyselect::starts_with("non_response_"), mean),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(digits = 3, caption = "Mean non-response rate by version")
```

v3 has roughly three times v1's orientation non-response. It is also heavily
aphantasic — 17 of its 21 participants, against 31 of 88 in v1. Those two facts
together are enough to create an association between imagery and non-response in
a pooled sample where none exists within any version. (v2 is more aphantasic
still in proportion, at 8 of 9, but with nine participants it moves a pooled
estimate very little; v3 has both the composition and the weight.)

```{r simpsons}
propensity <- by_participant |>
  dplyr::filter(!is.na(vviq))

dplyr::bind_rows(
  dplyr::mutate(
    correlation_test(propensity$non_response_orientation, propensity$vviq),
    sample = "Pooled"),
  propensity |>
    dplyr::group_split(version) |>
    purrr::map(\(d) dplyr::mutate(
      correlation_test(d$non_response_orientation, d$vviq),
      sample = dplyr::first(d$version))) |>
    purrr::list_rbind()
) |>
  dplyr::select(sample, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "Orientation non-response against VVIQ, pooled and within version")
```

Pooled, the association looks clear and significant. Within v1, which is
`r round(100 * mean(propensity$version == "v1"))`% of the participants, it is
absent. The pooled estimate is a mixture of strata rather than an effect, and
the version that drives it is the one with 21 participants.

This is the concrete reason behind the scoping decision. Pooling here produced a
finding that stratification removed.

## 3. v1 is also the cleanest sample, not just the largest {#ver-v1}

```{r v1-associations}
v1 <- dplyr::filter(by_participant, version == "v1")

v1_scores <- block_trials |>
  dplyr::filter(version == "v1") |>
  compose_features(id) |>
  dplyr::left_join(dplyr::select(v1, id, vviq), by = "id")

dplyr::bind_rows(
  dplyr::mutate(correlation_test(v1_scores$mean_word, v1_scores$vviq),
                feature = "Word"),
  dplyr::mutate(correlation_test(v1_scores$mean_angle, v1_scores$vviq),
                feature = "Orientation"),
  dplyr::mutate(correlation_test(v1_scores$mean_color, v1_scores$vviq),
                feature = "Colour")
) |>
  dplyr::select(feature, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "v1 only: responders-only recall accuracy against VVIQ")
```

Orientation has one participant fewer than the other two features: one person
answered **zero** orientation items, so they have no responders-only mean to
correlate. They are not missing at random, and the [joint
model](joint-model.html) is the analysis that can see them.

Analysed on its own, v1 gives an interpretable pattern. The [task
validity](task-validity.html) page covers what those associations do and do not
support.

There is a second, quieter argument. Per-version standardisation needs each
version to be able to carry its own scale, and v2 cannot. Drop one participant
at a time, re-standardise, and see how far everyone else's z-score moves:

```{r loo-standardisation}
# How much does one participant's presence move everyone else's z-score?
leave_one_out_shift <- function(x) {
  x <- x[!is.na(x)]
  full <- as.numeric(scale(x))
  purrr::map_dbl(seq_along(x), function(i) {
    without <- as.numeric(scale(x[-i]))
    max(abs(without - full[-i]))
  })
}

block_trials |>
  compose_features(id, version) |>
  dplyr::filter(stats::complete.cases(dplyr::pick(tidyselect::starts_with("mean_")))) |>
  dplyr::summarise(
    participants = dplyr::n(),
    median_shift = stats::median(leave_one_out_shift(mean_angle)),
    max_shift = max(leave_one_out_shift(mean_angle)),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(digits = 3,
               caption = "Largest z-score shift caused by removing one participant, orientation")
```

v2's median shift is roughly five times v1's: on nine participants, one person's
presence routinely moves everyone else's z-score by an appreciable fraction of a
standard deviation. That is not a scale, and it is why standardising within
version was abandoned.

## 4. What this means for every other page {#ver-scope}

- **All modelling pages are v1 only.** v1 has `r nrow(v1)` participants; no
  model uses all of them, because each applies further filters for vividness,
  engagement or questionnaire completeness. [The
  analysis](analysis-strategy.html#strat-samples) lists every resulting sample in one
  table.
- **v2 and v3 are described, not modelled.** They appear in `all_data`, in the
  [codebook](codebook.html), and in descriptive comparisons.
- **Version is treated as a structural feature, not a nuisance covariate.**
  Adding a version term to a pooled model would assume the versions differ by an
  additive constant, which section 2 shows is exactly what they do not do.
- **The restriction is a sample size, not a methodological ceiling.** The same
  pipeline applies to v3 once it recruits, and it is written to be re-run rather
  than rewritten.

## 5. What the excluded versions are still for {#ver-excluded}

v3 retains one job it uniquely enables. **v1 has a fixed recall order**, word
then orientation then colour, so feature differences there are confounded with
output position. v3 randomised the order, and it is the only version that can
estimate that effect. Any claim about differences between features carries that
caveat, and v3 is where it can be checked.

The choice of v1 has two costs. **v1 has no parity penalty**, so the distractor
there was free, which shapes what the [task engagement](engagement.html) page
can conclude. And **v1 has the lowest non-response of the three versions**, so
it is the version where declining to report is least common, which is worth
remembering when reading how much weight that behaviour carries.

--------------------------------------------------------------------------------

**Continuing through the Extended Online Report:** this page follows [task
engagement](engagement.html). To keep reading in order, continue to [task
validity](task-validity.html) next. Or skip ahead to [the
analysis](analysis-strategy.html), which explains what the modelling pages do
and why.

--------------------------------------------------------------------------------

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
