---
title: "Versions: which one the analyses use, and why"
output: rmarkdown::html_vignette
bibliography: ../references.bib
csl: ../apa.csl
vignette: >
  %\VignetteIndexEntry{Version scope}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5,
  dpi = 150, out.width = "100%", warning = FALSE, message = FALSE
)
library(aphantasiaWMStrats)
library(ggplot2)
theme_set(theme_pdf(base_size = 16))
fmt <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
```

The WM-FTT ran in three versions, and the [task design](task-design.html)
page explains why each replaced the last; the
[sample](sample-description.html) page describes who took part in each. This page answers a narrower
question with a consequence for everything after it: **which versions enter
the inferential analyses.**

The answer is **v1 only**. Every modelling page on this site inherits that
restriction, so the reasoning belongs somewhere a reader can check it
rather than in a sentence inside a methods section.

Framed deliberately as *v1 is primary* rather than *v2 and v3 are
excluded*. Both are kept in `all_data`, described here, and available to
anyone. The pipeline built on v1 is a template to be re-applied when the
later versions have the sample to support it.

```{r data}
block_trials <- all_data |>
  dplyr::filter(grepl("^expe_block", expe_phase))

by_participant <- block_trials |>
  dplyr::summarise(
    imagery_group = dplyr::first(vviq_group_2),
    vviq = dplyr::first(vviq_total_score),
    non_response_word = 1 - mean(responded_word),
    non_response_orientation = 1 - mean(responded_angle),
    non_response_colour = 1 - mean(responded_color),
    .by = c(id, version)
  )
```

## 1. The versions are not interchangeable samples

Pooling would be reasonable if the three versions were the same task run
three times. They are not: each changed the task in a way that could plausibly
change behaviour, and they recruited different populations.

```{r group-balance}
by_participant |>
  dplyr::mutate(imagery_group = dplyr::coalesce(as.character(imagery_group),
                                                "no VVIQ")) |>
  dplyr::count(version, imagery_group) |>
  tidyr::pivot_wider(names_from = imagery_group, values_from = n,
                     values_fill = 0L) |>
  knitr::kable(caption = "Participants by version and imagery group")
```

**Group balance is the decisive number.** v1 is 31 aphantasic to 55 typical
imagers. v2 is 8 to 1, and v3 is 17 to 4. In v2 and v3 a group comparison
is not merely underpowered: with one and four typical imagers respectively,
it is structurally impossible. No amount of modelling recovers a
contrast that the recruitment never sampled.

The later versions are also aphantasia-heavy *because* recruitment shifted,
which matters for the next section.

## 2. Pooling did not just add noise, it manufactured a result

Non-response differs sharply by version.

```{r nonresponse-by-version}
by_participant |>
  dplyr::summarise(
    dplyr::across(tidyselect::starts_with("non_response_"), mean),
    .by = version
  ) |>
  dplyr::arrange(version) |>
  knitr::kable(digits = 3, caption = "Mean non-response rate by version")
```

v3 has roughly three times v1's orientation non-response. It is also the
most aphantasia-heavy version. Those two facts together are enough to
create an association between imagery and non-response in a pooled sample
where none exists within any version.

```{r simpsons}
propensity <- by_participant |>
  dplyr::filter(!is.na(vviq))

dplyr::bind_rows(
  dplyr::mutate(
    correlation_test(propensity$non_response_orientation, propensity$vviq),
    sample = "Pooled"),
  propensity |>
    dplyr::group_split(version) |>
    purrr::map(\(d) dplyr::mutate(
      correlation_test(d$non_response_orientation, d$vviq),
      sample = dplyr::first(d$version))) |>
    purrr::list_rbind()
) |>
  dplyr::select(sample, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "Orientation non-response against VVIQ, pooled and within version")
```

Pooled, the association looks clear and significant. Within v1, which is 80%
of the participants, it is absent. The pooled estimate is a mixture of
strata rather than an effect, and the version that drives it is the one
with 21 participants.

This is the concrete reason the scope decision is not bureaucratic. Pooling
here produced a finding that stratification removed.

## 3. v1 is also the cleanest stratum, not just the largest

```{r v1-associations}
v1 <- dplyr::filter(by_participant, version == "v1")

v1_scores <- block_trials |>
  dplyr::filter(version == "v1") |>
  compose_features(id) |>
  dplyr::left_join(dplyr::select(v1, id, vviq), by = "id")

dplyr::bind_rows(
  dplyr::mutate(correlation_test(v1_scores$mean_word, v1_scores$vviq),
                feature = "Word"),
  dplyr::mutate(correlation_test(v1_scores$mean_angle, v1_scores$vviq),
                feature = "Orientation"),
  dplyr::mutate(correlation_test(v1_scores$mean_color, v1_scores$vviq),
                feature = "Colour")
) |>
  dplyr::select(feature, n, rho, p) |>
  knitr::kable(digits = 3,
               caption = "v1 only: responders-only recall accuracy against VVIQ")
```

Analysed on its own, v1 gives an interpretable pattern rather than a
mixture. The [task validity](scoring.html) material covers what those
associations do and do not support.

There is a second, quieter argument. Per-version standardisation needs each
version to be able to carry its own scale, and v2 cannot: removing any
single participant moves its z-scores by a median of 0.13 SD and by up to
0.945 SD. A scale that unstable is not a scale.

## 4. What this means for every other page

- **All modelling pages are v1 only**, N = `r nrow(v1)`, and say so.
- **v2 and v3 are described, not modelled.** They appear in `all_data`, in
  the [codebook](codebook.html), and in descriptive comparisons.
- **Version is treated as a structural feature, not a nuisance covariate.**
  Adding a version term to a pooled model would assume the versions differ
  by an additive constant, which section 2 shows is exactly what they do
  not do.
- **The restriction is a sample size, not a methodological ceiling.** The
  same pipeline applies to v3 once it recruits, and it is written to be
  re-run rather than rewritten.

## 5. What the excluded versions are still for

v3 retains one job it uniquely enables. **v1 has a fixed recall order**,
word then orientation then colour, so feature differences there are
confounded with output position. v3 randomised the order, and it is the
only version that can estimate that effect. Any claim about differences
between features carries that caveat, and v3 is where it can be checked.

Two costs of the choice are worth stating plainly rather than leaving to
be discovered. **v1 has no parity penalty**, so the distractor there was
free, which shapes what the [task engagement](engagement.html) page can
conclude. And **v1 has the lowest non-response of the three versions**, so
it is the version where declining to report is least common, which is
worth remembering when reading how much weight that behaviour carries.

---

**Continuing through the Extended Online Report:** this page follows [task engagement](engagement.html). To keep reading in order, continue to [task validity](task-validity.html) next. Or skip ahead to [the analysis](analysis-strategy.html), which explains what the modelling pages do and why.

---

```{r}
#| label: session_info
#| echo: false

sessioninfo::session_info()
```
